---
title: "M3: OrdRec with Movie Content Features (Phase A, Nm >= 10)"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(knitr)
```
---

## 0. Settings

```{r config}
# --- edit these paths (same as 07/09/10, plus the 08_M3 cache) --------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/selection_v7_2018-04_6m_Nu10_Nm10.rds"
CACHE_M12 <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
CACHE_M3  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/08_M3"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/08_M3"
# ----------------------------------------------------------------------

dir.create(CACHE_M3, showWarnings = FALSE, recursive = TRUE)
dir.create(OUT_DIR,  showWarnings = FALSE, recursive = TRUE)

SEED     <- 2024L
SVD_DIM  <- 16L
K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)

# Inherited verbatim from the 09 scan. No re-tuning: any difference
# between M2e and M3 must be attributable to the content term alone.
BATCH_SIZE <- 512L
MAX_EPOCHS <- 200L
PATIENCE   <- 8L
ETA_DECAY  <- 0.95
CLIP       <- 25
ETA        <- 0.03
LAMBDA     <- 0.02

# The ONE new hyperparameter. The content block has a different
# dimension and a different scale from the factor block, so libFM's
# grouped-regularisation argument applies: it gets its own lambda,
# scanned on the calibration set. M2e's lambda is untouched.
LAMBDA_C_GRID <- c(0.005, 0.02, 0.08, 0.32, 1.28)

N_BOOT <- 2000L

set.seed(SEED)
```

---

## 1. Load the Selection and the Features

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
cat(sprintf("fit %d | calib %d | test %d | users %d | movies %d\n",
            nrow(fit), nrow(calib), nrow(test), nU, nI))
stopifnot(nrow(fit) == 393364L)

# ---- content features -------------------------------------------------
feat <- readRDS(file.path(CACHE_M3, "m3_features_v1.rds"))
Cmat <- feat$C
# The single most important guard in this file: a silent row misalignment
# between C and movie_idx would attach every film's content to a
# different film, and the model would still train happily.
stopifnot(identical(as.integer(feat$movie_map$movieId),
                    as.integer(movie_map$movieId)),
          nrow(Cmat) == nI,
          all(is.finite(Cmat)))
nF <- ncol(Cmat)
cat(sprintf("Content matrix: %d films x %d features\n", nI, nF))
print(table(feat$block))

support <- feat$support
setorder(support, movie_idx)
stopifnot(identical(support$movie_idx, seq_len(nI)))
```

---

## 2. Core Functions

```{r core-functions}
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))

# Effective item factor: q_i + W' c_i. W = NULL means M3a (or M2e).
q_eff <- function(par, i) {
  if (is.null(par$W)) par$Q[i, , drop = FALSE]
  else par$Q[i, , drop = FALSE] + Cmat[i, , drop = FALSE] %*% par$W
}

calc_scores_m3 <- function(par, u, i) {
  s <- par$bi[i] + rowSums(par$P[u, , drop = FALSE] * q_eff(par, i))
  if (!is.null(par$w)) s <- s + as.numeric(Cmat[i, , drop = FALSE] %*% par$w)
  s
}

user_threshold_rows <- function(par, u) {
  E  <- exp(par$beta[u, , drop = FALSE])
  t1 <- par$t1[u]
  cbind(t1, t1 + E[, 1], t1 + E[, 1] + E[, 2],
        t1 + E[, 1] + E[, 2] + E[, 3])
}

ordrec_forward_e <- function(y, r, Tmat) {
  n <- length(y)
  g_hi <- ifelse(r == K_LEVELS, 1,
                 sigmoid(Tmat[cbind(seq_len(n), pmin(r, K_LEVELS - 1L))] - y))
  g_lo <- ifelse(r == 1L, 0,
                 sigmoid(Tmat[cbind(seq_len(n), pmax(r - 1L, 1L))] - y))
  list(prob = pmax(g_hi - g_lo, 1e-12),
       A = g_hi * (1 - g_hi), B = g_lo * (1 - g_lo))
}

ordrec_mean_ll_m3 <- function(par, dat) {
  y  <- calc_scores_m3(par, dat$u, dat$i)
  Tm <- user_threshold_rows(par, dat$u)
  mean(log(ordrec_forward_e(y, dat$r, Tm)$prob))
}

ordrec_row_ll_m3 <- function(par, dat) {
  y  <- calc_scores_m3(par, dat$u, dat$i)
  Tm <- user_threshold_rows(par, dat$u)
  log(ordrec_forward_e(y, dat$r, Tm)$prob)
}

ordrec_probs_m3 <- function(par, u, i) {
  y  <- calc_scores_m3(par, u, i)
  Tm <- user_threshold_rows(par, u)
  CP <- plogis(Tm - y)
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- metrics, verbatim from 07/09/10 ---------------------------------
tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

metric_set <- function(P, y) {
  y   <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll, row_rps = rps,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  mean(d[, {
    k <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]$ndcg, na.rm = TRUE)
}

fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  pu <- d[, {
    sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
    ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  sum(pu$nc) / (sum(pu$nc) + sum(pu$nd))
}
```

---

## 3. The Trainer


```{r trainer}
train_ordrec_m3 <- function(dat_fit, dat_calib, init, eta, lambda, lambda_c,
                            use_W = FALSE,
                            batch_size = BATCH_SIZE, max_epochs = MAX_EPOCHS,
                            patience = PATIENCE, seed = SEED, clip = CLIP,
                            eta_decay = ETA_DECAY, verbose = TRUE) {
  par <- init
  if (!use_W) par$W <- NULL
  n <- length(dat_fit$u)
  best <- list(ll = -Inf, par = par, epoch = 0L)
  hist <- data.table(epoch = integer(), fit_meanLL = numeric(),
                     calib_meanLL = numeric(), clip_rate = numeric(),
                     norm_w = numeric(), norm_W = numeric())
  set.seed(seed)
  eta_e <- eta

  for (ep in seq_len(max_epochs)) {
    perm <- sample.int(n); n_clip <- 0L
    for (b0 in seq(1L, n, by = batch_size)) {
      idx <- perm[b0:min(b0 + batch_size - 1L, n)]
      u <- dat_fit$u[idx]; i <- dat_fit$i[idx]; r <- dat_fit$r[idx]
      nb <- length(idx)
      Cb <- Cmat[i, , drop = FALSE]

      y  <- calc_scores_m3(par, u, i)
      E  <- exp(par$beta[u, , drop = FALSE])
      t1v <- par$t1[u]
      Tm <- cbind(t1v, t1v + E[, 1], t1v + E[, 1] + E[, 2],
                  t1v + E[, 1] + E[, 2] + E[, 3])
      fw <- ordrec_forward_e(y, r, Tm)
      c_hi <- fw$A / fw$prob; c_lo <- fw$B / fw$prob
      n_clip <- n_clip + sum(c_hi > clip) + sum(c_lo > clip)
      c_hi <- pmin(c_hi, clip); c_lo <- pmin(c_lo, clip)
      dy <- c_lo - c_hi                       # dl/dy per row

      # ---- per-user thresholds: batch-SUM (verbatim from 10) ---------
      g_t1 <- rowsum(c_hi - c_lo, u)
      g_b1 <- rowsum(E[, 1] * (c_hi * (1L <= r - 1L) - c_lo * (1L <= r - 2L)), u)
      g_b2 <- rowsum(E[, 2] * (c_hi * (2L <= r - 1L) - c_lo * (2L <= r - 2L)), u)
      g_b3 <- rowsum(E[, 3] * (c_hi * (3L <= r - 1L) - c_lo * (3L <= r - 2L)), u)
      ug <- as.integer(rownames(g_t1))
      par$t1[ug]      <- par$t1[ug]      + eta_e * as.numeric(g_t1)
      par$beta[ug, 1] <- par$beta[ug, 1] + eta_e * as.numeric(g_b1)
      par$beta[ug, 2] <- par$beta[ug, 2] + eta_e * as.numeric(g_b2)
      par$beta[ug, 3] <- par$beta[ug, 3] + eta_e * as.numeric(g_b3)

      # ---- user factors: gradient uses the EFFECTIVE item factor -----
      Qe <- q_eff(par, i)
      SP <- rowsum(dy * Qe, u)
      nu <- as.numeric(rowsum(rep(1, nb), u))
      par$P[ug, ] <- par$P[ug, ] + eta_e * (SP - lambda * nu * par$P[ug, , drop = FALSE])

      # ---- items: residual bias and residual factor ------------------
      si <- rowsum(dy, i); ig <- as.integer(rownames(si))
      ni <- as.numeric(rowsum(rep(1, nb), i))
      par$bi[ig] <- par$bi[ig] + eta_e * (as.numeric(si) - lambda * ni * par$bi[ig])
      SQ <- rowsum(dy * par$P[u, , drop = FALSE], i)
      par$Q[ig, ] <- par$Q[ig, ] + eta_e * (SQ - lambda * ni * par$Q[ig, , drop = FALSE])

      # ---- content bias w: global, batch-MEAN, lambda_c --------------
      g_w <- as.numeric(crossprod(Cb, dy)) / nb
      par$w <- par$w + eta_e * (g_w - lambda_c * par$w)

      # ---- content factor W: global, batch-MEAN, lambda_c ------------
      if (!is.null(par$W)) {
        g_W <- crossprod(Cb, dy * par$P[u, , drop = FALSE]) / nb
        par$W <- par$W + eta_e * (g_W - lambda_c * par$W)
      }

      if (any(!is.finite(par$w)) || any(!is.finite(par$t1[ug])))
        return(list(diverged = TRUE, best_ll = -Inf, best_epoch = 0L, history = hist))
    }

    ll_c <- ordrec_mean_ll_m3(par, dat_calib)
    ll_f <- ordrec_mean_ll_m3(par, dat_fit)
    hist <- rbind(hist, data.table(
      epoch = ep, fit_meanLL = ll_f, calib_meanLL = ll_c,
      clip_rate = n_clip / n,
      norm_w = sqrt(sum(par$w^2)),
      norm_W = if (is.null(par$W)) NA_real_ else sqrt(sum(par$W^2))))
    if (verbose) cat(sprintf("    epoch %2d  fit %.5f  calib %.5f  |w| %.3f\n",
                             ep, ll_f, ll_c, sqrt(sum(par$w^2))))
    if (ll_c > best$ll + 1e-5) best <- list(ll = ll_c, par = par, epoch = ep)
    else if (ep - best$epoch >= patience) break
    eta_e <- eta_e * eta_decay
  }
  list(diverged = FALSE, best_ll = best$ll, par = best$par,
       best_epoch = best$epoch, history = hist)
}
```

---

## 4. Correctness Checks

### 4a. Numerical gradient check on the two new blocks

```{r grad-check}
set.seed(11)
.nU <- 40L; .nI <- 30L; .K <- 4L; .nF <- 6L; .n <- 500L
.Csave <- Cmat
Cmat <<- matrix(rnorm(.nI * .nF), .nI, .nF)     # temporary global for q_eff

.p <- list(t1 = rnorm(.nU, -1, 0.3),
           beta = matrix(rnorm(.nU * 3, 0, 0.2), .nU, 3),
           bi = rnorm(.nI, 0, 0.3),
           P = matrix(rnorm(.nU * .K, 0, 0.3), .nU, .K),
           Q = matrix(rnorm(.nI * .K, 0, 0.3), .nI, .K),
           w = rnorm(.nF, 0, 0.2),
           W = matrix(rnorm(.nF * .K, 0, 0.2), .nF, .K))
.d <- list(u = sample(.nU, .n, TRUE), i = sample(.nI, .n, TRUE),
           r = sample(5L, .n, TRUE))

.obj <- function(p) sum(log(ordrec_forward_e(
  calc_scores_m3(p, .d$u, .d$i), .d$r, user_threshold_rows(p, .d$u))$prob))

.analytic <- function(p) {
  y  <- calc_scores_m3(p, .d$u, .d$i)
  Tm <- user_threshold_rows(p, .d$u)
  fw <- ordrec_forward_e(y, .d$r, Tm)
  dy <- fw$B / fw$prob - fw$A / fw$prob
  Cb <- Cmat[.d$i, , drop = FALSE]
  list(w = as.numeric(crossprod(Cb, dy)),
       W = crossprod(Cb, dy * p$P[.d$u, , drop = FALSE]))
}
.g <- .analytic(.p)
.eps <- 1e-6
.num_w <- sapply(seq_len(.nF), function(k) {
  pp <- .p; pp$w[k] <- pp$w[k] + .eps; pm <- .p; pm$w[k] <- pm$w[k] - .eps
  (.obj(pp) - .obj(pm)) / (2 * .eps)
})
.num_W <- outer(seq_len(.nF), seq_len(.K), Vectorize(function(a, b) {
  pp <- .p; pp$W[a, b] <- pp$W[a, b] + .eps
  pm <- .p; pm$W[a, b] <- pm$W[a, b] - .eps
  (.obj(pp) - .obj(pm)) / (2 * .eps)
}))
cat(sprintf("max |analytic - numeric|  w: %.3e   W: %.3e\n",
            max(abs(.g$w - .num_w)), max(abs(.g$W - .num_W))))
stopifnot(max(abs(.g$w - .num_w)) < 1e-4,
          max(abs(.g$W - .num_W)) < 1e-4)
cat("GRADIENT CHECK: PASS\n")
Cmat <<- .Csave
rm(.p, .d, .g, .num_w, .num_W, .Csave)
```

### 4b. Reduction check: content off must reproduce M2e


```{r reduction-check}
m2e <- readRDS(file.path(CACHE_M12, "m2e_final.rds"))
stopifnot(!m2e$diverged)
cat(sprintf("Cached M2e: best epoch %d | calib meanLL %.5f\n",
            m2e$best_epoch, m2e$best_ll))

dat_fit   <- list(u = fit$user_idx,   i = fit$movie_idx,   r = fit$rating5)
dat_calib <- list(u = calib$user_idx, i = calib$movie_idx, r = calib$rating5)
dat_test  <- list(u = test$user_idx,  i = test$movie_idx,  r = test$rating5)

marg5  <- prop.table(tabulate(dat_fit$r, K_LEVELS))
t_marg <- qlogis(cumsum(marg5)[1:4])
set.seed(SEED)
init_cold <- list(t1   = rep(t_marg[1], nU),
                  beta = matrix(log(pmax(diff(t_marg), 1e-3)), nU, 3, byrow = TRUE),
                  bi = rep(0, nI),
                  P = matrix(rnorm(nU * SVD_DIM, 0, 0.05), nU, SVD_DIM),
                  Q = matrix(rnorm(nI * SVD_DIM, 0, 0.05), nI, SVD_DIM),
                  w = rep(0, nF),
                  W = matrix(0, nF, SVD_DIM))

red_cache <- file.path(CACHE_M3, "m3_reduction_check.rds")
if (file.exists(red_cache)) {
  m3_null <- readRDS(red_cache)
} else {
  init0 <- init_cold; init0$w <- rep(0, nF)
  m3_null <- train_ordrec_m3(dat_fit, dat_calib, init0, eta = ETA,
                             lambda = LAMBDA, lambda_c = 1e6,  # pins w at 0
                             use_W = FALSE, verbose = FALSE)
  saveRDS(m3_null, red_cache)
}
cat(sprintf("Reduction run: calib %.5f (M2e %.5f, diff %.2e)\n",
            m3_null$best_ll, m2e$best_ll, m3_null$best_ll - m2e$best_ll))
stopifnot(abs(m3_null$best_ll - m2e$best_ll) < 5e-4)
cat("REDUCTION CHECK: PASS\n")
```

---

## 5. Selecting `lambda_c` on Calibration


```{r lambda-scan}
scan_cache <- file.path(CACHE_M3, "m3a_lambda_scan.rds")
if (file.exists(scan_cache)) {
  scan_res <- readRDS(scan_cache)
} else {
  scan_res <- rbindlist(lapply(LAMBDA_C_GRID, function(lc) {
    cat(sprintf("  lambda_c = %.4f\n", lc))
    r <- train_ordrec_m3(dat_fit, dat_calib, init_cold, eta = ETA,
                         lambda = LAMBDA, lambda_c = lc,
                         use_W = FALSE, verbose = FALSE)
    data.table(lambda_c = lc, calib_meanLL = r$best_ll,
               best_epoch = r$best_epoch, norm_w = sqrt(sum(r$par$w^2)),
               diverged = r$diverged)
  }))
  saveRDS(scan_res, scan_cache)
}
kable(scan_res, digits = 5, caption = "M3a: calibration scan over the content penalty")

LAMBDA_C <- scan_res[which.max(calib_meanLL), lambda_c]
cat(sprintf("Selected lambda_c = %g\n", LAMBDA_C))
if (LAMBDA_C == min(LAMBDA_C_GRID) || LAMBDA_C == max(LAMBDA_C_GRID))
  warning("lambda_c selected at a grid boundary - widen LAMBDA_C_GRID.")

p_scan <- ggplot(scan_res, aes(lambda_c, calib_meanLL)) +
  geom_line(colour = "steelblue") + geom_point(size = 2) +
  scale_x_log10() + theme_minimal() +
  geom_hline(yintercept = m2e$best_ll, linetype = "dashed", colour = "grey40") +
  labs(title = "M3a: content penalty scan",
       subtitle = "Dashed line is M2e's calibration likelihood",
       x = expression(lambda[c]), y = "calibration mean log-likelihood")
print(p_scan)
ggsave(file.path(OUT_DIR, "m3b_lambda_scan.png"), p_scan, width = 7, height = 4, dpi = 150)
```

---

## 6. Final Training

```{r final-train}
fit_or_load <- function(tag, use_W) {
  f <- file.path(CACHE_M3, sprintf("%s_final.rds", tag))
  if (file.exists(f)) { cat(sprintf("Loaded cached %s.\n", tag)); return(readRDS(f)) }
  r <- train_ordrec_m3(dat_fit, dat_calib, init_cold, eta = ETA,
                       lambda = LAMBDA, lambda_c = LAMBDA_C,
                       use_W = use_W, verbose = TRUE)
  saveRDS(r, f); r
}
m3a <- fit_or_load("m3a", use_W = FALSE)
m3b <- fit_or_load("m3b", use_W = TRUE)
stopifnot(!m3a$diverged, !m3b$diverged)

train_tbl <- data.table(
  model = c("M2e", "M3a", "M3b"),
  best_epoch   = c(m2e$best_epoch, m3a$best_epoch, m3b$best_epoch),
  calib_meanLL = c(m2e$best_ll,    m3a$best_ll,    m3b$best_ll),
  fit_calib_gap = c(
    m2e$history[epoch == m2e$best_epoch, fit_meanLL - calib_meanLL],
    m3a$history[epoch == m3a$best_epoch, fit_meanLL - calib_meanLL],
    m3b$history[epoch == m3b$best_epoch, fit_meanLL - calib_meanLL]))
kable(train_tbl, digits = 5, caption = "Training summary")
```

---

## 7. Test-Month Evaluation

```{r test-eval}
par_m2e <- m2e$par; par_m2e$w <- NULL; par_m2e$W <- NULL   # scored by the same code path

P_list <- list(M2e = ordrec_probs_m3(par_m2e, dat_test$u, dat_test$i),
               M3a = ordrec_probs_m3(m3a$par, dat_test$u, dat_test$i),
               M3b = ordrec_probs_m3(m3b$par, dat_test$u, dat_test$i))

res_list <- lapply(P_list, metric_set, y = test$rating5)
metrics <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))

rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  data.table(model = m,
             NDCG10 = ndcg_at_10(E, test$rating5, test$userId),
             FCP    = fcp_stat(E, test$rating5, test$userId))
}))
metrics <- merge(metrics, rank_tbl, by = "model", sort = FALSE)
kable(metrics, digits = 5, caption = "Test month (2018-10)")
```

---

## 8. The Result: Paired Contrasts, Stratified by Item Support

```{r paired-boot}
LLm <- sapply(P_list, function(P) log(tidy_probs(P)[cbind(seq_len(nrow(P)), test$rating5)]))
D <- data.table(d_a = LLm[, "M3a"] - LLm[, "M2e"],
                d_b = LLm[, "M3b"] - LLm[, "M2e"],
                d_ba = LLm[, "M3b"] - LLm[, "M3a"])

test[, n_train := support$n_train[movie_idx]]
test[, nm_band := cut(n_train, c(9, 20, 50, 200, Inf),
                      labels = c("10-20", "21-50", "51-200", "200+"))]
cat("Test rows per stratum:\n"); print(test[, .N, by = nm_band][order(nm_band)])

u_key <- as.integer(factor(test$userId))
i_key <- as.integer(factor(test$movieId))
nU_t <- max(u_key); nI_t <- max(i_key); n_test <- nrow(test)

boot_paired <- function(d, rows, B = N_BOOT) {
  dv <- d[rows]; uk <- u_key[rows]; ik <- i_key[rows]
  out <- numeric(B)
  for (b in seq_len(B)) {
    cu <- tabulate(sample.int(nU_t, nU_t, replace = TRUE), nbins = nU_t)
    ci <- tabulate(sample.int(nI_t, nI_t, replace = TRUE), nbins = nI_t)
    w  <- cu[uk] * ci[ik]
    sw <- sum(w)
    out[b] <- if (sw > 0) sum(dv * w) / sw else NA_real_
  }
  out[is.finite(out)]
}

set.seed(SEED)
strata <- c(list(All = seq_len(n_test)),
            split(seq_len(n_test), test$nm_band))
contrasts <- c(d_a = "M3a - M2e", d_b = "M3b - M2e", d_ba = "M3b - M3a")

boot_tbl <- rbindlist(lapply(names(strata), function(s) {
  rows <- strata[[s]]
  rbindlist(lapply(names(contrasts), function(cn) {
    bs <- boot_paired(D[[cn]], rows)
    data.table(stratum = s, contrast = contrasts[[cn]],
               n_rows = length(rows),
               delta = mean(D[[cn]][rows]),
               lo = unname(quantile(bs, 0.025)),
               hi = unname(quantile(bs, 0.975)))
  }))
}))
boot_tbl[, sig := ifelse(lo > 0 | hi < 0, "*", "")]
kable(boot_tbl, digits = 5,
      caption = "Paired mean log-likelihood gain, two-way bootstrap 95% CI")
fwrite(boot_tbl, file.path(OUT_DIR, "m3c_stratified_contrasts.csv"))
```

```{r decay-plot, fig.height=4.2}
pd <- boot_tbl[stratum != "All"]
pd[, stratum := factor(stratum, levels = c("10-20", "21-50", "51-200", "200+"))]
p_decay <- ggplot(pd, aes(stratum, delta, colour = contrast, group = contrast)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.12,
                position = position_dodge(0.35)) +
  geom_point(size = 2.4, position = position_dodge(0.35)) +
  theme_minimal() +
  labs(title = "Does the content gain decay with item support?",
       subtitle = "Pre-registered prediction: monotone decline, near zero in the densest stratum",
       x = "training ratings for the film", y = "paired mean log-likelihood gain (nats)",
       colour = NULL)
print(p_decay)
ggsave(file.path(OUT_DIR, "m3d_support_decay.png"), p_decay,
       width = 8, height = 4.2, dpi = 150)
```

---

## 9. Where the Content Term Puts Its Weight

```{r block-weights}
wt <- data.table(feature = colnames(Cmat), block = feat$block,
                 w_a = m3a$par$w, w_b = m3b$par$w,
                 W_norm = sqrt(rowSums(m3b$par$W^2)))
blk <- wt[, .(n_features = .N,
              L2_w = sqrt(sum(w_a^2)),
              L2_W = sqrt(sum(W_norm^2))), by = block][order(-L2_w)]
kable(blk, digits = 4, caption = "Content weight by feature block (M3a bias, M3b factor)")
fwrite(wt, file.path(OUT_DIR, "m3e_content_weights.csv"))

top <- wt[order(-abs(w_a))][1:20]
p_w <- ggplot(top, aes(reorder(feature, w_a), w_a, fill = block)) +
  geom_col() + coord_flip() + theme_minimal() +
  labs(title = "Largest content-bias coefficients (M3a)",
       subtitle = "Penalised and correlated: read as allocation of capacity, not as effects",
       x = NULL, y = expression(w))
print(p_w)
ggsave(file.path(OUT_DIR, "m3f_content_weights.png"), p_w, width = 8, height = 6, dpi = 150)
```

---

## 10. Save

```{r save}
m3_results <- list(
  metrics    = metrics,
  train_tbl  = train_tbl,
  scan       = scan_res,
  lambda_c   = LAMBDA_C,
  contrasts  = boot_tbl,
  weights    = wt,
  row_delta  = D,
  strata_n   = test[, .N, by = nm_band][order(nm_band)],
  meta = list(built = Sys.time(), n_boot = N_BOOT, seed = SEED,
              feature_version = feat$meta$version,
              n_features = nF, excluded = feat$meta$excluded))
saveRDS(m3_results, file.path(CACHE_M3, "m3_results.rds"))
fwrite(metrics, file.path(OUT_DIR, "m3_vs_m2e_comparison.csv"))
cat("06b COMPLETE.\n")
```
