---
title: "M3: OrdRec with Movie Content Features (Phase A, Nm >= 10)"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(knitr)
```

## Aim

M2e is the paper's configuration and the end of the ordinal ladder. M3
asks the question Koren and Sill leave open in their final paragraph:
whether *item attributes for rarely rated items* add anything on top of
it. Two models are trained, both cold-start, both with M2e's
hyperparameters:

$$\text{M3a:}\quad y_{ui} = b_i + w^\top c_i + q_i^\top p_u$$
$$\text{M3b:}\quad y_{ui} = b_i + w^\top c_i + \big(q_i + W^\top c_i\big)^\top p_u$$

with $c_i$ the standardised content vector from `05b`. Everything else —
the logit link, the per-user thresholds $t^{(u)}$, the likelihood, the
mini-batch trainer — is M2e's, unchanged.

**What M3a actually does, and why it is not vacuous.** Every film already
has a free bias $b_i$, so $w^\top c_i$ adds *no capacity*: with $b_i$
unpenalised the two models would be exactly equivalent. The content term
acts only through the penalty. Because $b_i$ is shrunk toward zero and
the film's total bias is $b_i + w^\top c_i$, the effect of the content
term is to move the **shrinkage target** from the global mean to a
content-predicted value. M3a is therefore a feature-based prior on the
item bias, in the sense of Agarwal and Chen's RLFM, whose MovieLens-100K
result — a test set with no new users or items, where the entire gain
came from "better regularization achieved through feature-based prior" —
is the closest published analogue to this setup. M3b extends the same
logic from the bias to the factor, as in Wang and Blei's
$v_j = \theta_j + \epsilon_j$.

**The prediction being tested.** Shrinkage matters most where the data is
thinnest, so the gain should decay monotonically with item support. Strub
et al. report exactly this on MovieLens-10M: an RMSE improvement of
1.811% in the sparsest quintile of items falling to 0.052% in the
densest, with only 0.206% overall. Pilászy and Tikk locate the crossover
at roughly ten ratings — and this selection prunes at $N_m \ge 10$, so
**the whole catalogue sits on the side where ratings already dominate
metadata.** The pre-registered expectation is therefore a small or
zero overall gain with a monotone decay across strata. The stratified
contrast, not the headline number, is the result.

**Trust chain.** The likelihood/gradient core is M2e's, which passed the
numerical check in 08 and the `clm` cross-check in 07. The two new
gradient blocks ($w$, $W$) get their own numerical check in Section 4,
and the trainer is re-run through M2e's synthetic self-test with the
content term switched off, which must reproduce M2e's behaviour.

---

## 0. Settings

```{r config}
# --- edit these paths (same as 07/09/10, plus the 08_M3 cache) --------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/selection_v7_2018-04_6m_Nu10_Nm10.rds"
CACHE_M12 <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
CACHE_M3  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/08_M3"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/08_M3"
# ----------------------------------------------------------------------

dir.create(CACHE_M3, showWarnings = FALSE, recursive = TRUE)
dir.create(OUT_DIR,  showWarnings = FALSE, recursive = TRUE)

SEED     <- 2024L
SVD_DIM  <- 16L
K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)

# Inherited verbatim from the 09 scan. No re-tuning: any difference
# between M2e and M3 must be attributable to the content term alone.
BATCH_SIZE <- 512L
MAX_EPOCHS <- 200L
PATIENCE   <- 8L
ETA_DECAY  <- 0.95
CLIP       <- 25
ETA        <- 0.03
LAMBDA     <- 0.02

# The ONE new hyperparameter. The content block has a different
# dimension and a different scale from the factor block, so libFM's
# grouped-regularisation argument applies: it gets its own lambda,
# scanned on the calibration set. M2e's lambda is untouched.
LAMBDA_C_GRID <- c(0.005, 0.02, 0.08, 0.32, 1.28)

N_BOOT <- 2000L

set.seed(SEED)
```

---

## 1. Load the Selection and the Features

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
cat(sprintf("fit %d | calib %d | test %d | users %d | movies %d\n",
            nrow(fit), nrow(calib), nrow(test), nU, nI))
stopifnot(nrow(fit) == 393364L)

# ---- content features -------------------------------------------------
feat <- readRDS(file.path(CACHE_M3, "m3_features_v1.rds"))
Cmat <- feat$C
# The single most important guard in this file: a silent row misalignment
# between C and movie_idx would attach every film's content to a
# different film, and the model would still train happily.
stopifnot(identical(as.integer(feat$movie_map$movieId),
                    as.integer(movie_map$movieId)),
          nrow(Cmat) == nI,
          all(is.finite(Cmat)))
nF <- ncol(Cmat)
cat(sprintf("Content matrix: %d films x %d features\n", nI, nF))
print(table(feat$block))

support <- feat$support
setorder(support, movie_idx)
stopifnot(identical(support$movie_idx, seq_len(nI)))
```

---

## 2. Core Functions

Metric functions are verbatim from 07/09/10 so the evaluation protocol
stays bit-identical across the whole ladder. The scoring functions are
M2e's with the content term added; setting `W = NULL` and `w = 0`
recovers M2e exactly, which is what the self-test in Section 4 exploits.

```{r core-functions}
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))

# Effective item factor: q_i + W' c_i. W = NULL means M3a (or M2e).
q_eff <- function(par, i) {
  if (is.null(par$W)) par$Q[i, , drop = FALSE]
  else par$Q[i, , drop = FALSE] + Cmat[i, , drop = FALSE] %*% par$W
}

calc_scores_m3 <- function(par, u, i) {
  s <- par$bi[i] + rowSums(par$P[u, , drop = FALSE] * q_eff(par, i))
  if (!is.null(par$w)) s <- s + as.numeric(Cmat[i, , drop = FALSE] %*% par$w)
  s
}

user_threshold_rows <- function(par, u) {
  E  <- exp(par$beta[u, , drop = FALSE])
  t1 <- par$t1[u]
  cbind(t1, t1 + E[, 1], t1 + E[, 1] + E[, 2],
        t1 + E[, 1] + E[, 2] + E[, 3])
}

ordrec_forward_e <- function(y, r, Tmat) {
  n <- length(y)
  g_hi <- ifelse(r == K_LEVELS, 1,
                 sigmoid(Tmat[cbind(seq_len(n), pmin(r, K_LEVELS - 1L))] - y))
  g_lo <- ifelse(r == 1L, 0,
                 sigmoid(Tmat[cbind(seq_len(n), pmax(r - 1L, 1L))] - y))
  list(prob = pmax(g_hi - g_lo, 1e-12),
       A = g_hi * (1 - g_hi), B = g_lo * (1 - g_lo))
}

ordrec_mean_ll_m3 <- function(par, dat) {
  y  <- calc_scores_m3(par, dat$u, dat$i)
  Tm <- user_threshold_rows(par, dat$u)
  mean(log(ordrec_forward_e(y, dat$r, Tm)$prob))
}

ordrec_row_ll_m3 <- function(par, dat) {
  y  <- calc_scores_m3(par, dat$u, dat$i)
  Tm <- user_threshold_rows(par, dat$u)
  log(ordrec_forward_e(y, dat$r, Tm)$prob)
}

ordrec_probs_m3 <- function(par, u, i) {
  y  <- calc_scores_m3(par, u, i)
  Tm <- user_threshold_rows(par, u)
  CP <- plogis(Tm - y)
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- metrics, verbatim from 07/09/10 ---------------------------------
tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

metric_set <- function(P, y) {
  y   <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll, row_rps = rps,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  mean(d[, {
    k <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]$ndcg, na.rm = TRUE)
}

fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  pu <- d[, {
    sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
    ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  sum(pu$nc) / (sum(pu$nc) + sum(pu$nd))
}
```

---

## 3. The Trainer

M2e's trainer with two new parameter blocks. Three decisions carry over
from 09/10 and one is new:

- **Per-user thresholds, $b_i$, $P$, $Q$**: batch-SUM gradient with
  per-occurrence L2, exactly as in 10. These are entity parameters,
  touched by only some rows of a batch.
- **$w$ and $W$ are global parameters**, touched by *every* row of every
  batch, like the four global thresholds in 09. They therefore take the
  batch-MEAN gradient and a plain (not per-occurrence) L2 at
  `lambda_c`. Using the sum here would make their effective learning
  rate scale with batch size.
- **$q_i$ is penalised toward zero, not toward $W^\top c_i$.** That is
  the whole mechanism: the total item factor is $q_i + W^\top c_i$, so
  shrinking the residual $q_i$ pulls the total toward the content
  prediction. Same for $b_i$ against $w^\top c_i$.
- Clipping, learning-rate decay and early stopping on calibration
  meanLL are unchanged.

```{r trainer}
train_ordrec_m3 <- function(dat_fit, dat_calib, init, eta, lambda, lambda_c,
                            use_W = FALSE,
                            batch_size = BATCH_SIZE, max_epochs = MAX_EPOCHS,
                            patience = PATIENCE, seed = SEED, clip = CLIP,
                            eta_decay = ETA_DECAY, verbose = TRUE) {
  par <- init
  if (!use_W) par$W <- NULL
  n <- length(dat_fit$u)
  best <- list(ll = -Inf, par = par, epoch = 0L)
  hist <- data.table(epoch = integer(), fit_meanLL = numeric(),
                     calib_meanLL = numeric(), clip_rate = numeric(),
                     norm_w = numeric(), norm_W = numeric())
  set.seed(seed)
  eta_e <- eta

  for (ep in seq_len(max_epochs)) {
    perm <- sample.int(n); n_clip <- 0L
    for (b0 in seq(1L, n, by = batch_size)) {
      idx <- perm[b0:min(b0 + batch_size - 1L, n)]
      u <- dat_fit$u[idx]; i <- dat_fit$i[idx]; r <- dat_fit$r[idx]
      nb <- length(idx)
      Cb <- Cmat[i, , drop = FALSE]

      y  <- calc_scores_m3(par, u, i)
      E  <- exp(par$beta[u, , drop = FALSE])
      t1v <- par$t1[u]
      Tm <- cbind(t1v, t1v + E[, 1], t1v + E[, 1] + E[, 2],
                  t1v + E[, 1] + E[, 2] + E[, 3])
      fw <- ordrec_forward_e(y, r, Tm)
      c_hi <- fw$A / fw$prob; c_lo <- fw$B / fw$prob
      n_clip <- n_clip + sum(c_hi > clip) + sum(c_lo > clip)
      c_hi <- pmin(c_hi, clip); c_lo <- pmin(c_lo, clip)
      dy <- c_lo - c_hi                       # dl/dy per row

      # ---- per-user thresholds: batch-SUM (verbatim from 10) ---------
      g_t1 <- rowsum(c_hi - c_lo, u)
      g_b1 <- rowsum(E[, 1] * (c_hi * (1L <= r - 1L) - c_lo * (1L <= r - 2L)), u)
      g_b2 <- rowsum(E[, 2] * (c_hi * (2L <= r - 1L) - c_lo * (2L <= r - 2L)), u)
      g_b3 <- rowsum(E[, 3] * (c_hi * (3L <= r - 1L) - c_lo * (3L <= r - 2L)), u)
      ug <- as.integer(rownames(g_t1))
      par$t1[ug]      <- par$t1[ug]      + eta_e * as.numeric(g_t1)
      par$beta[ug, 1] <- par$beta[ug, 1] + eta_e * as.numeric(g_b1)
      par$beta[ug, 2] <- par$beta[ug, 2] + eta_e * as.numeric(g_b2)
      par$beta[ug, 3] <- par$beta[ug, 3] + eta_e * as.numeric(g_b3)

      # ---- user factors: gradient uses the EFFECTIVE item factor -----
      Qe <- q_eff(par, i)
      SP <- rowsum(dy * Qe, u)
      nu <- as.numeric(rowsum(rep(1, nb), u))
      par$P[ug, ] <- par$P[ug, ] + eta_e * (SP - lambda * nu * par$P[ug, , drop = FALSE])

      # ---- items: residual bias and residual factor ------------------
      si <- rowsum(dy, i); ig <- as.integer(rownames(si))
      ni <- as.numeric(rowsum(rep(1, nb), i))
      par$bi[ig] <- par$bi[ig] + eta_e * (as.numeric(si) - lambda * ni * par$bi[ig])
      SQ <- rowsum(dy * par$P[u, , drop = FALSE], i)
      par$Q[ig, ] <- par$Q[ig, ] + eta_e * (SQ - lambda * ni * par$Q[ig, , drop = FALSE])

      # ---- content bias w: global, batch-MEAN, lambda_c --------------
      g_w <- as.numeric(crossprod(Cb, dy)) / nb
      par$w <- par$w + eta_e * (g_w - lambda_c * par$w)

      # ---- content factor W: global, batch-MEAN, lambda_c ------------
      if (!is.null(par$W)) {
        g_W <- crossprod(Cb, dy * par$P[u, , drop = FALSE]) / nb
        par$W <- par$W + eta_e * (g_W - lambda_c * par$W)
      }

      if (any(!is.finite(par$w)) || any(!is.finite(par$t1[ug])))
        return(list(diverged = TRUE, best_ll = -Inf, best_epoch = 0L, history = hist))
    }

    ll_c <- ordrec_mean_ll_m3(par, dat_calib)
    ll_f <- ordrec_mean_ll_m3(par, dat_fit)
    hist <- rbind(hist, data.table(
      epoch = ep, fit_meanLL = ll_f, calib_meanLL = ll_c,
      clip_rate = n_clip / n,
      norm_w = sqrt(sum(par$w^2)),
      norm_W = if (is.null(par$W)) NA_real_ else sqrt(sum(par$W^2))))
    if (verbose) cat(sprintf("    epoch %2d  fit %.5f  calib %.5f  |w| %.3f\n",
                             ep, ll_f, ll_c, sqrt(sum(par$w^2))))
    if (ll_c > best$ll + 1e-5) best <- list(ll = ll_c, par = par, epoch = ep)
    else if (ep - best$epoch >= patience) break
    eta_e <- eta_e * eta_decay
  }
  list(diverged = FALSE, best_ll = best$ll, par = best$par,
       best_epoch = best$epoch, history = hist)
}
```

---

## 4. Correctness Checks

### 4a. Numerical gradient check on the two new blocks

The analytic gradients for $w$ and $W$ are the only genuinely new algebra
in this file, so they are checked against finite differences on a small
synthetic problem before anything is trained.

```{r grad-check}
set.seed(11)
.nU <- 40L; .nI <- 30L; .K <- 4L; .nF <- 6L; .n <- 500L
.Csave <- Cmat
Cmat <<- matrix(rnorm(.nI * .nF), .nI, .nF)     # temporary global for q_eff

.p <- list(t1 = rnorm(.nU, -1, 0.3),
           beta = matrix(rnorm(.nU * 3, 0, 0.2), .nU, 3),
           bi = rnorm(.nI, 0, 0.3),
           P = matrix(rnorm(.nU * .K, 0, 0.3), .nU, .K),
           Q = matrix(rnorm(.nI * .K, 0, 0.3), .nI, .K),
           w = rnorm(.nF, 0, 0.2),
           W = matrix(rnorm(.nF * .K, 0, 0.2), .nF, .K))
.d <- list(u = sample(.nU, .n, TRUE), i = sample(.nI, .n, TRUE),
           r = sample(5L, .n, TRUE))

.obj <- function(p) sum(log(ordrec_forward_e(
  calc_scores_m3(p, .d$u, .d$i), .d$r, user_threshold_rows(p, .d$u))$prob))

.analytic <- function(p) {
  y  <- calc_scores_m3(p, .d$u, .d$i)
  Tm <- user_threshold_rows(p, .d$u)
  fw <- ordrec_forward_e(y, .d$r, Tm)
  dy <- fw$B / fw$prob - fw$A / fw$prob
  Cb <- Cmat[.d$i, , drop = FALSE]
  list(w = as.numeric(crossprod(Cb, dy)),
       W = crossprod(Cb, dy * p$P[.d$u, , drop = FALSE]))
}
.g <- .analytic(.p)
.eps <- 1e-6
.num_w <- sapply(seq_len(.nF), function(k) {
  pp <- .p; pp$w[k] <- pp$w[k] + .eps; pm <- .p; pm$w[k] <- pm$w[k] - .eps
  (.obj(pp) - .obj(pm)) / (2 * .eps)
})
.num_W <- outer(seq_len(.nF), seq_len(.K), Vectorize(function(a, b) {
  pp <- .p; pp$W[a, b] <- pp$W[a, b] + .eps
  pm <- .p; pm$W[a, b] <- pm$W[a, b] - .eps
  (.obj(pp) - .obj(pm)) / (2 * .eps)
}))
cat(sprintf("max |analytic - numeric|  w: %.3e   W: %.3e\n",
            max(abs(.g$w - .num_w)), max(abs(.g$W - .num_W))))
stopifnot(max(abs(.g$w - .num_w)) < 1e-4,
          max(abs(.g$W - .num_W)) < 1e-4)
cat("GRADIENT CHECK: PASS\n")
Cmat <<- .Csave
rm(.p, .d, .g, .num_w, .num_W, .Csave)
```

### 4b. Reduction check: content off must reproduce M2e

With `w = 0`, `W = NULL` and `lambda_c` irrelevant, the trainer's update
equations reduce to 10's line for line. Running it from M2e's own
cold-start initialisation and the same seed must land on the same
calibration likelihood.

```{r reduction-check}
m2e <- readRDS(file.path(CACHE_M12, "m2e_final.rds"))
stopifnot(!m2e$diverged)
cat(sprintf("Cached M2e: best epoch %d | calib meanLL %.5f\n",
            m2e$best_epoch, m2e$best_ll))

dat_fit   <- list(u = fit$user_idx,   i = fit$movie_idx,   r = fit$rating5)
dat_calib <- list(u = calib$user_idx, i = calib$movie_idx, r = calib$rating5)
dat_test  <- list(u = test$user_idx,  i = test$movie_idx,  r = test$rating5)

marg5  <- prop.table(tabulate(dat_fit$r, K_LEVELS))
t_marg <- qlogis(cumsum(marg5)[1:4])
set.seed(SEED)
init_cold <- list(t1   = rep(t_marg[1], nU),
                  beta = matrix(log(pmax(diff(t_marg), 1e-3)), nU, 3, byrow = TRUE),
                  bi = rep(0, nI),
                  P = matrix(rnorm(nU * SVD_DIM, 0, 0.05), nU, SVD_DIM),
                  Q = matrix(rnorm(nI * SVD_DIM, 0, 0.05), nI, SVD_DIM),
                  w = rep(0, nF),
                  W = matrix(0, nF, SVD_DIM))

red_cache <- file.path(CACHE_M3, "m3_reduction_check.rds")
if (file.exists(red_cache)) {
  m3_null <- readRDS(red_cache)
} else {
  init0 <- init_cold; init0$w <- rep(0, nF)
  m3_null <- train_ordrec_m3(dat_fit, dat_calib, init0, eta = ETA,
                             lambda = LAMBDA, lambda_c = 1e6,  # pins w at 0
                             use_W = FALSE, verbose = FALSE)
  saveRDS(m3_null, red_cache)
}
cat(sprintf("Reduction run: calib %.5f (M2e %.5f, diff %.2e)\n",
            m3_null$best_ll, m2e$best_ll, m3_null$best_ll - m2e$best_ll))
stopifnot(abs(m3_null$best_ll - m2e$best_ll) < 5e-4)
cat("REDUCTION CHECK: PASS\n")
```

---

## 5. Selecting `lambda_c` on Calibration

M2e's `lambda` stays at 0.02 throughout. Only the content block's penalty
is scanned, on the calibration set, as libFM's grouped regularisation
argument prescribes. The scan is run for M3a; M3b reuses the selected
value so the M3a/M3b contrast is not confounded by a second search.

```{r lambda-scan}
scan_cache <- file.path(CACHE_M3, "m3a_lambda_scan.rds")
if (file.exists(scan_cache)) {
  scan_res <- readRDS(scan_cache)
} else {
  scan_res <- rbindlist(lapply(LAMBDA_C_GRID, function(lc) {
    cat(sprintf("  lambda_c = %.4f\n", lc))
    r <- train_ordrec_m3(dat_fit, dat_calib, init_cold, eta = ETA,
                         lambda = LAMBDA, lambda_c = lc,
                         use_W = FALSE, verbose = FALSE)
    data.table(lambda_c = lc, calib_meanLL = r$best_ll,
               best_epoch = r$best_epoch, norm_w = sqrt(sum(r$par$w^2)),
               diverged = r$diverged)
  }))
  saveRDS(scan_res, scan_cache)
}
kable(scan_res, digits = 5, caption = "M3a: calibration scan over the content penalty")

LAMBDA_C <- scan_res[which.max(calib_meanLL), lambda_c]
cat(sprintf("Selected lambda_c = %g\n", LAMBDA_C))
if (LAMBDA_C == min(LAMBDA_C_GRID) || LAMBDA_C == max(LAMBDA_C_GRID))
  warning("lambda_c selected at a grid boundary - widen LAMBDA_C_GRID.")

p_scan <- ggplot(scan_res, aes(lambda_c, calib_meanLL)) +
  geom_line(colour = "steelblue") + geom_point(size = 2) +
  scale_x_log10() + theme_minimal() +
  geom_hline(yintercept = m2e$best_ll, linetype = "dashed", colour = "grey40") +
  labs(title = "M3a: content penalty scan",
       subtitle = "Dashed line is M2e's calibration likelihood",
       x = expression(lambda[c]), y = "calibration mean log-likelihood")
print(p_scan)
ggsave(file.path(OUT_DIR, "m3b_lambda_scan.png"), p_scan, width = 7, height = 4, dpi = 150)
```

---

## 6. Final Training

```{r final-train}
fit_or_load <- function(tag, use_W) {
  f <- file.path(CACHE_M3, sprintf("%s_final.rds", tag))
  if (file.exists(f)) { cat(sprintf("Loaded cached %s.\n", tag)); return(readRDS(f)) }
  r <- train_ordrec_m3(dat_fit, dat_calib, init_cold, eta = ETA,
                       lambda = LAMBDA, lambda_c = LAMBDA_C,
                       use_W = use_W, verbose = TRUE)
  saveRDS(r, f); r
}
m3a <- fit_or_load("m3a", use_W = FALSE)
m3b <- fit_or_load("m3b", use_W = TRUE)
stopifnot(!m3a$diverged, !m3b$diverged)

train_tbl <- data.table(
  model = c("M2e", "M3a", "M3b"),
  best_epoch   = c(m2e$best_epoch, m3a$best_epoch, m3b$best_epoch),
  calib_meanLL = c(m2e$best_ll,    m3a$best_ll,    m3b$best_ll),
  fit_calib_gap = c(
    m2e$history[epoch == m2e$best_epoch, fit_meanLL - calib_meanLL],
    m3a$history[epoch == m3a$best_epoch, fit_meanLL - calib_meanLL],
    m3b$history[epoch == m3b$best_epoch, fit_meanLL - calib_meanLL]))
kable(train_tbl, digits = 5, caption = "Training summary")
```

---

## 7. Test-Month Evaluation

```{r test-eval}
par_m2e <- m2e$par; par_m2e$w <- NULL; par_m2e$W <- NULL   # scored by the same code path

P_list <- list(M2e = ordrec_probs_m3(par_m2e, dat_test$u, dat_test$i),
               M3a = ordrec_probs_m3(m3a$par, dat_test$u, dat_test$i),
               M3b = ordrec_probs_m3(m3b$par, dat_test$u, dat_test$i))

res_list <- lapply(P_list, metric_set, y = test$rating5)
metrics <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))

rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  data.table(model = m,
             NDCG10 = ndcg_at_10(E, test$rating5, test$userId),
             FCP    = fcp_stat(E, test$rating5, test$userId))
}))
metrics <- merge(metrics, rank_tbl, by = "model", sort = FALSE)
kable(metrics, digits = 5, caption = "Test month (2018-10)")
```

---

## 8. The Result: Paired Contrasts, Stratified by Item Support

Two decisions here decide whether the analysis can see anything at all.

**Paired, not marginal.** The quantity of interest is the per-row
difference $\Delta_{ui} = \ell^{M3}_{ui} - \ell^{M2e}_{ui}$, resampled as
a difference. Comparing two separately-resampled marginal means would
carry the between-row variance of the likelihood itself, which is orders
of magnitude larger than the effect being measured and would swamp it.

**Two-way bootstrap.** Test rows are crossed by user and by item, not
independent, so the resampling weights both margins by the product of
independent draw counts, as in file 20 (Owen and Eckles).

```{r paired-boot}
LLm <- sapply(P_list, function(P) log(tidy_probs(P)[cbind(seq_len(nrow(P)), test$rating5)]))
D <- data.table(d_a = LLm[, "M3a"] - LLm[, "M2e"],
                d_b = LLm[, "M3b"] - LLm[, "M2e"],
                d_ba = LLm[, "M3b"] - LLm[, "M3a"])

test[, n_train := support$n_train[movie_idx]]
test[, nm_band := cut(n_train, c(9, 20, 50, 200, Inf),
                      labels = c("10-20", "21-50", "51-200", "200+"))]
cat("Test rows per stratum:\n"); print(test[, .N, by = nm_band][order(nm_band)])

u_key <- as.integer(factor(test$userId))
i_key <- as.integer(factor(test$movieId))
nU_t <- max(u_key); nI_t <- max(i_key); n_test <- nrow(test)

boot_paired <- function(d, rows, B = N_BOOT) {
  dv <- d[rows]; uk <- u_key[rows]; ik <- i_key[rows]
  out <- numeric(B)
  for (b in seq_len(B)) {
    cu <- tabulate(sample.int(nU_t, nU_t, replace = TRUE), nbins = nU_t)
    ci <- tabulate(sample.int(nI_t, nI_t, replace = TRUE), nbins = nI_t)
    w  <- cu[uk] * ci[ik]
    sw <- sum(w)
    out[b] <- if (sw > 0) sum(dv * w) / sw else NA_real_
  }
  out[is.finite(out)]
}

set.seed(SEED)
strata <- c(list(All = seq_len(n_test)),
            split(seq_len(n_test), test$nm_band))
contrasts <- c(d_a = "M3a - M2e", d_b = "M3b - M2e", d_ba = "M3b - M3a")

boot_tbl <- rbindlist(lapply(names(strata), function(s) {
  rows <- strata[[s]]
  rbindlist(lapply(names(contrasts), function(cn) {
    bs <- boot_paired(D[[cn]], rows)
    data.table(stratum = s, contrast = contrasts[[cn]],
               n_rows = length(rows),
               delta = mean(D[[cn]][rows]),
               lo = unname(quantile(bs, 0.025)),
               hi = unname(quantile(bs, 0.975)))
  }))
}))
boot_tbl[, sig := ifelse(lo > 0 | hi < 0, "*", "")]
kable(boot_tbl, digits = 5,
      caption = "Paired mean log-likelihood gain, two-way bootstrap 95% CI")
fwrite(boot_tbl, file.path(OUT_DIR, "m3c_stratified_contrasts.csv"))
```

```{r decay-plot, fig.height=4.2}
pd <- boot_tbl[stratum != "All"]
pd[, stratum := factor(stratum, levels = c("10-20", "21-50", "51-200", "200+"))]
p_decay <- ggplot(pd, aes(stratum, delta, colour = contrast, group = contrast)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.12,
                position = position_dodge(0.35)) +
  geom_point(size = 2.4, position = position_dodge(0.35)) +
  theme_minimal() +
  labs(title = "Does the content gain decay with item support?",
       subtitle = "Pre-registered prediction: monotone decline, near zero in the densest stratum",
       x = "training ratings for the film", y = "paired mean log-likelihood gain (nats)",
       colour = NULL)
print(p_decay)
ggsave(file.path(OUT_DIR, "m3d_support_decay.png"), p_decay,
       width = 8, height = 4.2, dpi = 150)
```

> **Plain read.** The `All` row is the headline and is expected to be
> small. The pattern across the four strata is the result. A monotone
> decline confirms the shrinkage account and reproduces Strub et al.'s
> quintile pattern under an ordinal likelihood rather than RMSE; a flat
> profile would falsify it and would need explaining — the first thing to
> check would be the feature/support correlations written out by `05b`,
> since a content-composition difference between strata would produce a
> gain that does not come from sparsity at all.

---

## 9. Where the Content Term Puts Its Weight

Which blocks earn their parameters. This is descriptive, not inferential:
the coefficients are penalised and correlated, so they are read as
"where the model chose to spend capacity", not as effect estimates.

```{r block-weights}
wt <- data.table(feature = colnames(Cmat), block = feat$block,
                 w_a = m3a$par$w, w_b = m3b$par$w,
                 W_norm = sqrt(rowSums(m3b$par$W^2)))
blk <- wt[, .(n_features = .N,
              L2_w = sqrt(sum(w_a^2)),
              L2_W = sqrt(sum(W_norm^2))), by = block][order(-L2_w)]
kable(blk, digits = 4, caption = "Content weight by feature block (M3a bias, M3b factor)")
fwrite(wt, file.path(OUT_DIR, "m3e_content_weights.csv"))

top <- wt[order(-abs(w_a))][1:20]
p_w <- ggplot(top, aes(reorder(feature, w_a), w_a, fill = block)) +
  geom_col() + coord_flip() + theme_minimal() +
  labs(title = "Largest content-bias coefficients (M3a)",
       subtitle = "Penalised and correlated: read as allocation of capacity, not as effects",
       x = NULL, y = expression(w))
print(p_w)
ggsave(file.path(OUT_DIR, "m3f_content_weights.png"), p_w, width = 8, height = 6, dpi = 150)
```

---

## 10. Save

```{r save}
m3_results <- list(
  metrics    = metrics,
  train_tbl  = train_tbl,
  scan       = scan_res,
  lambda_c   = LAMBDA_C,
  contrasts  = boot_tbl,
  weights    = wt,
  row_delta  = D,
  strata_n   = test[, .N, by = nm_band][order(nm_band)],
  meta = list(built = Sys.time(), n_boot = N_BOOT, seed = SEED,
              feature_version = feat$meta$version,
              n_features = nF, excluded = feat$meta$excluded))
saveRDS(m3_results, file.path(CACHE_M3, "m3_results.rds"))
fwrite(metrics, file.path(OUT_DIR, "m3_vs_m2e_comparison.csv"))
cat("06b COMPLETE.\n")
```

---

## 11. Limitations Specific to M3

- **The content term is collinear with $b_i$ and $q_i$ by construction.**
  Any gain is a regularisation effect, not new capacity. This is the
  intended mechanism, but it means M3 cannot be described as "adding
  information the model did not have"; it re-aims the shrinkage.
- **`lambda_c` is tuned but M2e's `lambda` is not re-tuned alongside it.**
  A joint scan might find a different optimum for the pair. The
  single-axis scan is the price of keeping the M2e comparison clean.
- **The sparsest stratum here is 10–20 ratings**, because the selection
  prunes at $N_m \ge 10$. Pilászy and Tikk place the ratings-vs-metadata
  crossover at about ten ratings, so this design cannot observe the
  regime where content is predicted to dominate. That is what the
  Phase B reselection at lower $N_m$ is for, and this limitation is the
  reason it is not optional.
- **Text features are a TF-IDF/SVD bag of words.** Kim et al. report that
  bag-of-words item models add almost nothing over plain MF on dense
  MovieLens (CTR 0.8969 vs PMF 0.8971 on ML-1m) and attribute this to the
  loss of word order. A contextual encoder might do better; that is out
  of scope.
