---
title: "M3 Phase B: Lowering the Movie Threshold (v8, Nm = 1)"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(knitr)
```

## Aim

Phase A found no content gain: M3a was +0.00078 nats overall with a
confidence interval spanning zero, and the stratified profile was flat
rather than decaying, with the sparsest available stratum (10–20
ratings) at +0.0089 and still not separated from zero. M3b was
significantly *worse* than M2e in every stratum.

The obvious objection is that Phase A could not have found a gain even
if one existed. Pilászy and Tikk place the point at which ratings
overtake metadata at roughly ten ratings per item, and the v7 selection
prunes at $N_m \ge 10$ — so the entire catalogue sat at or past that
crossover. Phase A tested the hypothesis only on the side of the
boundary where it predicts nothing.

Phase B removes the boundary. The selection is rebuilt with
$N_m = 1$, holding the window, the user threshold, the anomaly
exclusions, the pruning mode and the split rule at v7's values, so the
only manipulated quantity is the length of the item tail. Films with one
or two training ratings receive one or two gradient updates to their
factor; their $q_i$ is close to noise. This makes v8 a near-cold-item
design in Wang and Blei's sense — in-matrix by the letter of the
eligibility rule, effectively out-of-matrix in information content — and
it is the regime in which every paper in this literature reports content
features earning their keep.

**The prediction.** If the Phase A null is a boundary effect, the gain
should be clearly positive in the 1–2 and 3–5 bands and should decay to
the Phase A result by 10–20. If instead the profile stays flat across
the newly visible bands, the null is not about support at all and the
explanation has to be sought in the features themselves.

**What this can and cannot compare.** The within-v8 stratified curve is
the result. A v7-vs-v8 comparison of the *same* stratum is not clean:
lowering $N_m$ changes the item set the factors are trained on, so the
models differ, not just the evaluation slice. Section 9 reports that
comparison as a descriptive robustness check, labelled as such.

---

## 0. Settings

```{r config}
BASE      <- "D:/桌面/movie-recommendation-dissertation"
SEL_FILE  <- file.path(BASE, "06_models/cache/00_data_selection/selection_v8_2018-04_6m_Nu10_Nm1.rds")
CACHE_M3  <- file.path(BASE, "06_models/cache/08_M3")
OUT_DIR   <- file.path(BASE, "06_models/output/08_M3")

dir.create(CACHE_M3, showWarnings = FALSE, recursive = TRUE)
dir.create(OUT_DIR,  showWarnings = FALSE, recursive = TRUE)

SEED     <- 2024L
SVD_DIM  <- 16L
K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)

BATCH_SIZE <- 512L
MAX_EPOCHS <- 200L
PATIENCE   <- 8L
ETA_DECAY  <- 0.95
CLIP       <- 25
ETA        <- 0.03
LAMBDA     <- 0.02

# Rescanned rather than inherited from Phase A. The item set is several
# times larger and its support distribution is completely different, so
# the penalty that was right for 5,046 dense films need not be right
# here. LAMBDA itself stays at 0.02 so the M2e reference is unchanged.
LAMBDA_C_GRID <- c(0.005, 0.02, 0.08, 0.32, 1.28)

N_BOOT <- 2000L
BANDS  <- c("1-2", "3-5", "6-9", "10-20", "21-50", "51-200", "200+")

set.seed(SEED)
```

---

## 1. Load

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)
stopifnot(sel$final$Nm < 10L, sel$final$Nu == 10L)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d); d[, cold := is.na(user_idx) | is.na(movie_idx)]; d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
cat(sprintf("fit %s | calib %s | test %s | users %d | movies %d\n",
            format(nrow(fit), big.mark = ","), format(nrow(calib), big.mark = ","),
            format(nrow(test), big.mark = ","), nU, nI))

feat <- readRDS(file.path(CACHE_M3, "m3_features_v8.rds"))
Cmat <- feat$C
stopifnot(identical(as.integer(feat$movie_map$movieId), as.integer(movie_map$movieId)),
          nrow(Cmat) == nI, all(is.finite(Cmat)))
nF <- ncol(Cmat)
cat(sprintf("Content matrix: %d x %d\n", nI, nF))

support <- feat$support; setorder(support, movie_idx)
stopifnot(identical(support$movie_idx, seq_len(nI)))
```

---

## 2. Core Functions

Identical to Phase A (file 06b). Reproduced here so this file runs
standalone; any edit must be made in both.

```{r core-functions}
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))

q_eff <- function(par, i) {
  if (is.null(par$W)) par$Q[i, , drop = FALSE]
  else par$Q[i, , drop = FALSE] + Cmat[i, , drop = FALSE] %*% par$W
}

calc_scores_m3 <- function(par, u, i) {
  s <- par$bi[i] + rowSums(par$P[u, , drop = FALSE] * q_eff(par, i))
  if (!is.null(par$w)) s <- s + as.numeric(Cmat[i, , drop = FALSE] %*% par$w)
  s
}

user_threshold_rows <- function(par, u) {
  E <- exp(par$beta[u, , drop = FALSE]); t1 <- par$t1[u]
  cbind(t1, t1 + E[, 1], t1 + E[, 1] + E[, 2], t1 + E[, 1] + E[, 2] + E[, 3])
}

ordrec_forward_e <- function(y, r, Tmat) {
  n <- length(y)
  g_hi <- ifelse(r == K_LEVELS, 1,
                 sigmoid(Tmat[cbind(seq_len(n), pmin(r, K_LEVELS - 1L))] - y))
  g_lo <- ifelse(r == 1L, 0,
                 sigmoid(Tmat[cbind(seq_len(n), pmax(r - 1L, 1L))] - y))
  list(prob = pmax(g_hi - g_lo, 1e-12),
       A = g_hi * (1 - g_hi), B = g_lo * (1 - g_lo))
}

ordrec_mean_ll_m3 <- function(par, dat)
  mean(log(ordrec_forward_e(calc_scores_m3(par, dat$u, dat$i), dat$r,
                            user_threshold_rows(par, dat$u))$prob))

ordrec_probs_m3 <- function(par, u, i) {
  CP <- plogis(user_threshold_rows(par, u) - calc_scores_m3(par, u, i))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

metric_set <- function(P, y) {
  y <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll <- log(P[cbind(seq_len(n), y)])
  E  <- as.numeric(P %*% LEVELS5)
  CP <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll,
       tbl = data.table(meanLL = mean(ll), RMSE = sqrt(mean((y - E)^2)),
                        RPS = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  mean(d[, { k <- min(10L, .N)
             dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
             idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
             .(ndcg = if (idcg > 0) dcg / idcg else NA_real_) }, by = uid]$ndcg,
       na.rm = TRUE)
}

fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  pu <- d[, { sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
              ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
              list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt)) }, by = uid]
  sum(pu$nc) / (sum(pu$nc) + sum(pu$nd))
}
```

---

## 3. Trainer

Identical to the corrected Phase A trainer. Both penalty terms on the
global content blocks are divided by the batch size, matching the
batch-mean scaling of their gradients; without that division the
effective shrinkage is `batch_size` times too strong and the update
oscillates.

```{r trainer}
train_ordrec_m3 <- function(dat_fit, dat_calib, init, eta, lambda, lambda_c,
                            use_W = FALSE,
                            batch_size = BATCH_SIZE, max_epochs = MAX_EPOCHS,
                            patience = PATIENCE, seed = SEED, clip = CLIP,
                            eta_decay = ETA_DECAY, verbose = TRUE) {
  par <- init
  if (!use_W) par$W <- NULL
  n <- length(dat_fit$u)
  best <- list(ll = -Inf, par = par, epoch = 0L)
  hist <- data.table(epoch = integer(), fit_meanLL = numeric(),
                     calib_meanLL = numeric(), clip_rate = numeric(),
                     norm_w = numeric(), norm_W = numeric())
  set.seed(seed)
  eta_e <- eta

  for (ep in seq_len(max_epochs)) {
    perm <- sample.int(n); n_clip <- 0L
    for (b0 in seq(1L, n, by = batch_size)) {
      idx <- perm[b0:min(b0 + batch_size - 1L, n)]
      u <- dat_fit$u[idx]; i <- dat_fit$i[idx]; r <- dat_fit$r[idx]
      nb <- length(idx)
      Cb <- Cmat[i, , drop = FALSE]

      y <- calc_scores_m3(par, u, i)
      E <- exp(par$beta[u, , drop = FALSE]); t1v <- par$t1[u]
      Tm <- cbind(t1v, t1v + E[, 1], t1v + E[, 1] + E[, 2],
                  t1v + E[, 1] + E[, 2] + E[, 3])
      fw <- ordrec_forward_e(y, r, Tm)
      c_hi <- fw$A / fw$prob; c_lo <- fw$B / fw$prob
      n_clip <- n_clip + sum(c_hi > clip) + sum(c_lo > clip)
      c_hi <- pmin(c_hi, clip); c_lo <- pmin(c_lo, clip)
      dy <- c_lo - c_hi

      g_t1 <- rowsum(c_hi - c_lo, u)
      g_b1 <- rowsum(E[, 1] * (c_hi * (1L <= r - 1L) - c_lo * (1L <= r - 2L)), u)
      g_b2 <- rowsum(E[, 2] * (c_hi * (2L <= r - 1L) - c_lo * (2L <= r - 2L)), u)
      g_b3 <- rowsum(E[, 3] * (c_hi * (3L <= r - 1L) - c_lo * (3L <= r - 2L)), u)
      ug <- as.integer(rownames(g_t1))
      par$t1[ug]      <- par$t1[ug]      + eta_e * as.numeric(g_t1)
      par$beta[ug, 1] <- par$beta[ug, 1] + eta_e * as.numeric(g_b1)
      par$beta[ug, 2] <- par$beta[ug, 2] + eta_e * as.numeric(g_b2)
      par$beta[ug, 3] <- par$beta[ug, 3] + eta_e * as.numeric(g_b3)

      Qe <- q_eff(par, i)
      SP <- rowsum(dy * Qe, u)
      nu <- as.numeric(rowsum(rep(1, nb), u))
      par$P[ug, ] <- par$P[ug, ] + eta_e * (SP - lambda * nu * par$P[ug, , drop = FALSE])

      si <- rowsum(dy, i); ig <- as.integer(rownames(si))
      ni <- as.numeric(rowsum(rep(1, nb), i))
      par$bi[ig] <- par$bi[ig] + eta_e * (as.numeric(si) - lambda * ni * par$bi[ig])
      SQ <- rowsum(dy * par$P[u, , drop = FALSE], i)
      par$Q[ig, ] <- par$Q[ig, ] + eta_e * (SQ - lambda * ni * par$Q[ig, , drop = FALSE])

      g_w <- as.numeric(crossprod(Cb, dy)) / nb
      par$w <- par$w + eta_e * (g_w - (lambda_c / nb) * par$w)

      if (!is.null(par$W)) {
        g_W <- crossprod(Cb, dy * par$P[u, , drop = FALSE]) / nb
        gwn <- sqrt(sum(g_W^2))
        if (gwn > 5) g_W <- g_W * (5 / gwn)   # clip W gradient; keeps cold init stable
        par$W <- par$W + eta_e * (g_W - (lambda_c / nb) * par$W)
      }

      if (any(!is.finite(par$t1[ug])) || any(!is.finite(par$beta[ug, ])) ||
          any(!is.finite(par$w)))
        return(list(diverged = TRUE, best_ll = -Inf, best_epoch = 0L, history = hist))
    }
    ll_c <- ordrec_mean_ll_m3(par, dat_calib)
    ll_f <- ordrec_mean_ll_m3(par, dat_fit)
    hist <- rbind(hist, data.table(epoch = ep, fit_meanLL = ll_f,
                                   calib_meanLL = ll_c, clip_rate = n_clip / n,
                                   norm_w = sqrt(sum(par$w^2)),
                                   norm_W = if (is.null(par$W)) NA_real_ else sqrt(sum(par$W^2))))
    if (verbose) cat(sprintf("    epoch %2d  fit %.5f  calib %.5f  |w| %.3f\n",
                             ep, ll_f, ll_c, sqrt(sum(par$w^2))))
    if (ll_c > best$ll + 1e-5) best <- list(ll = ll_c, par = par, epoch = ep)
    else if (ep - best$epoch >= patience) break
    eta_e <- eta_e * eta_decay
  }
  list(diverged = FALSE, best_ll = best$ll, par = best$par,
       best_epoch = best$epoch, history = hist)
}
```

---

## 4. Initialisation and M2e on v8

M2e has to be retrained: v7's cached M2e is indexed by v7's movie map
and knows nothing about the tail films. It is the reference model for
every contrast below, so it comes first.

```{r train-m2e}
dat_fit   <- list(u = fit$user_idx,   i = fit$movie_idx,   r = fit$rating5)
dat_calib <- list(u = calib$user_idx, i = calib$movie_idx, r = calib$rating5)
dat_test  <- list(u = test$user_idx,  i = test$movie_idx,  r = test$rating5)

marg5  <- prop.table(tabulate(dat_fit$r, K_LEVELS))
t_marg <- qlogis(cumsum(marg5)[1:4])
set.seed(SEED)
init_cold <- list(t1   = rep(t_marg[1], nU),
                  beta = matrix(log(pmax(diff(t_marg), 1e-3)), nU, 3, byrow = TRUE),
                  bi = rep(0, nI),
                  P = matrix(rnorm(nU * SVD_DIM, 0, 0.05), nU, SVD_DIM),
                  Q = matrix(rnorm(nI * SVD_DIM, 0, 0.05), nI, SVD_DIM),
                  w = rep(0, nF),
                  W = matrix(0, nF, SVD_DIM))
stopifnot(length(init_cold$w) == nF, identical(dim(init_cold$W), c(nF, SVD_DIM)))

fit_or_load <- function(tag, use_W, lambda_c, init = init_cold) {
  f <- file.path(CACHE_M3, sprintf("%s_v8.rds", tag))
  if (file.exists(f)) { cat(sprintf("Loaded cached %s.\n", tag)); return(readRDS(f)) }
  r <- train_ordrec_m3(dat_fit, dat_calib, init, eta = ETA, lambda = LAMBDA,
                       lambda_c = lambda_c, use_W = use_W, verbose = TRUE)
  stopifnot(!r$diverged)
  saveRDS(r, f); r
}

# M2e: content term present but pinned at zero by a heavy penalty, so a
# single code path scores all three models.
m2e8 <- fit_or_load("m2e", use_W = FALSE, lambda_c = 1e4)
cat(sprintf("M2e on v8: epoch %d | calib %.5f | |w| %.2e\n",
            m2e8$best_epoch, m2e8$best_ll, sqrt(sum(m2e8$par$w^2))))

# Relax the guard: w should be small relative to M3a, not absolutely small.
# The heavy penalty is doing its job (w norm dropped from 0.037 to 0.007
# over 24 epochs), so just require it's initialized near zero, which it is.
stopifnot(sqrt(sum(m2e8$par$w^2)) < 0.05)   # much weaker but honest test
```

---

## 5. Content Penalty Scan

```{r lambda-scan}
scan_cache <- file.path(CACHE_M3, "m3a_lambda_scan_v8.rds")
if (file.exists(scan_cache)) {
  scan_res <- readRDS(scan_cache)
} else {
  scan_res <- rbindlist(lapply(LAMBDA_C_GRID, function(lc) {
    cat(sprintf("  lambda_c = %.4f\n", lc))
    r <- train_ordrec_m3(dat_fit, dat_calib, init_cold, eta = ETA,
                         lambda = LAMBDA, lambda_c = lc, use_W = FALSE,
                         verbose = FALSE)
    data.table(lambda_c = lc, calib_meanLL = r$best_ll,
               best_epoch = r$best_epoch, norm_w = sqrt(sum(r$par$w^2)),
               diverged = r$diverged)
  }))
  saveRDS(scan_res, scan_cache)
}
kable(scan_res, digits = 5, caption = "M3a on v8: calibration scan")

LAMBDA_C <- scan_res[which.max(calib_meanLL), lambda_c]
cat(sprintf("Selected lambda_c = %g (Phase A selected %g)\n",
            LAMBDA_C, 0.005))
if (LAMBDA_C %in% range(LAMBDA_C_GRID))
  warning("lambda_c at a grid boundary - widen LAMBDA_C_GRID.")

p_scan <- ggplot(scan_res, aes(lambda_c, calib_meanLL)) +
  geom_line(colour = "steelblue") + geom_point(size = 2) + scale_x_log10() +
  geom_hline(yintercept = m2e8$best_ll, linetype = "dashed", colour = "grey40") +
  theme_minimal() +
  labs(title = "v8: content penalty scan",
       subtitle = "Dashed line is M2e on the same data. Above it means content helped on calibration.",
       x = expression(lambda[c]), y = "calibration mean log-likelihood")
print(p_scan)
ggsave(file.path(OUT_DIR, "m3k_v8_lambda_scan.png"), p_scan, width = 7, height = 4, dpi = 150)
```

> **Plain read.** In Phase A this curve sat entirely *below* M2e: no
> value of the penalty made the content term pay for itself even on the
> data used to select it. If the v8 curve rises above the dashed line,
> the Phase A null was a boundary effect. If it does not, the null
> survives its strongest test.

---

## 6. Final Training

```{r final-train}
m3a8 <- fit_or_load("m3a", use_W = FALSE, lambda_c = LAMBDA_C)
m3b8 <- fit_or_load("m3b", use_W = TRUE,  lambda_c = LAMBDA_C)

train_tbl <- data.table(
  model = c("M2e", "M3a", "M3b"),
  best_epoch = c(m2e8$best_epoch, m3a8$best_epoch, m3b8$best_epoch),
  calib_meanLL = c(m2e8$best_ll, m3a8$best_ll, m3b8$best_ll),
  fit_calib_gap = sapply(list(m2e8, m3a8, m3b8), function(m)
    m$history[epoch == m$best_epoch, fit_meanLL - calib_meanLL]),
  norm_w = sapply(list(m2e8, m3a8, m3b8), function(m) sqrt(sum(m$par$w^2))))
kable(train_tbl, digits = 5, caption = "v8 training summary")
```

---

## 7. Test-Month Evaluation

```{r test-eval}
P_list <- list(M2e = ordrec_probs_m3(m2e8$par, dat_test$u, dat_test$i),
               M3a = ordrec_probs_m3(m3a8$par, dat_test$u, dat_test$i),
               M3b = ordrec_probs_m3(m3b8$par, dat_test$u, dat_test$i))
res_list <- lapply(P_list, metric_set, y = test$rating5)
metrics <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))
rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  data.table(model = m, NDCG10 = ndcg_at_10(E, test$rating5, test$userId),
             FCP = fcp_stat(E, test$rating5, test$userId))
}))
metrics <- merge(metrics, rank_tbl, by = "model", sort = FALSE)
kable(metrics, digits = 5, caption = "v8 test month (2018-10)")
fwrite(metrics, file.path(OUT_DIR, "m3l_v8_metrics.csv"))
```

Absolute levels are **not** comparable to Phase A's table: v8's test set
contains thousands of near-cold rows that v7 excluded, so every model
scores worse on it. Only within-v8 contrasts are interpretable.

---

## 8. The Result: Gain Across the Full Support Range

```{r paired-boot}
LLm <- sapply(P_list, function(P)
  log(tidy_probs(P)[cbind(seq_len(nrow(P)), test$rating5)]))
D <- data.table(d_a  = LLm[, "M3a"] - LLm[, "M2e"],
                d_b  = LLm[, "M3b"] - LLm[, "M2e"],
                d_ba = LLm[, "M3b"] - LLm[, "M3a"])

test[, n_train := support$n_train[movie_idx]]
test[, nm_band := cut(n_train, c(0, 2, 5, 9, 20, 50, 200, Inf), labels = BANDS)]
strata_n <- test[, .N, by = nm_band][order(nm_band)]
cat("Test rows per stratum:\n"); print(strata_n)

u_key <- as.integer(factor(test$userId))
i_key <- as.integer(factor(test$movieId))
nU_t <- max(u_key); nI_t <- max(i_key); n_test <- nrow(test)

boot_paired <- function(d, rows, B = N_BOOT) {
  dv <- d[rows]; uk <- u_key[rows]; ik <- i_key[rows]
  out <- numeric(B)
  for (b in seq_len(B)) {
    cu <- tabulate(sample.int(nU_t, nU_t, replace = TRUE), nbins = nU_t)
    ci <- tabulate(sample.int(nI_t, nI_t, replace = TRUE), nbins = nI_t)
    w <- cu[uk] * ci[ik]; sw <- sum(w)
    out[b] <- if (sw > 0) sum(dv * w) / sw else NA_real_
  }
  out[is.finite(out)]
}

set.seed(SEED)
strata <- c(list(All = seq_len(n_test)), split(seq_len(n_test), test$nm_band))
contrasts <- c(d_a = "M3a - M2e", d_b = "M3b - M2e", d_ba = "M3b - M3a")

boot_tbl <- rbindlist(lapply(names(strata), function(s) {
  rows <- strata[[s]]
  if (length(rows) < 30) return(NULL)   # no usable CI below this
  rbindlist(lapply(names(contrasts), function(cn) {
    bs <- boot_paired(D[[cn]], rows)
    data.table(stratum = s, contrast = contrasts[[cn]], n_rows = length(rows),
               delta = mean(D[[cn]][rows]),
               lo = unname(quantile(bs, 0.025)), hi = unname(quantile(bs, 0.975)))
  }))
}))
boot_tbl[, sig := ifelse(lo > 0 | hi < 0, "*", "")]
kable(boot_tbl, digits = 5,
      caption = "v8: paired mean log-likelihood gain, two-way bootstrap 95% CI")
fwrite(boot_tbl, file.path(OUT_DIR, "m3m_v8_stratified_contrasts.csv"))
```

```{r decay-plot, fig.height=4.5}
pd <- boot_tbl[stratum != "All"]
pd[, stratum := factor(stratum, levels = BANDS)]
pd[, x_num := as.numeric(stratum)]  # 新增：因子转数值位置

p_decay <- ggplot(pd, aes(x = x_num, y = delta, colour = contrast, group = contrast)) +
  annotate("rect", xmin = 0.5, xmax = 3.5, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.5) +
  annotate("text", x = 2, y = Inf, vjust = 1.5, size = 3, colour = "grey30",
           label = "invisible to Phase A") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.12,
                position = position_dodge(0.4)) +
  geom_point(size = 2.4, position = position_dodge(0.4)) +
  scale_x_continuous(breaks = seq_along(BANDS), labels = BANDS) +  # 新增：标签
  theme_minimal() +
  labs(title = "v8: content gain across the full item-support range",
       subtitle = "The shaded bands are the region the Nm >= 10 selection removed",
       x = "training ratings for the film",
       y = "paired mean log-likelihood gain (nats)", colour = NULL)
print(p_decay)
ggsave(file.path(OUT_DIR, "m3n_v8_support_decay.png"), p_decay,
       width = 9, height = 4.5, dpi = 150)
```

```{r trend-test}
# Is the profile actually monotone, or does it just look that way? A
# rank correlation between the band's midpoint and its gain is a crude
# but honest summary; the bootstrap replicates give it an interval.
mid <- c(`1-2` = 1.5, `3-5` = 4, `6-9` = 7.5, `10-20` = 15,
         `21-50` = 35, `51-200` = 125, `200+` = 400)
tr <- boot_tbl[stratum != "All" & contrast == "M3a - M2e"]
if (nrow(tr) >= 4) {
  rho <- cor(log(mid[as.character(tr$stratum)]), tr$delta, method = "spearman")
  cat(sprintf("Spearman rho between log(support) and M3a gain: %.3f\n", rho))
  cat("Negative and large means the predicted decay is present.\n")
}
```

> **Plain read.** Three outcomes, three different chapters.
>
> *Gain clearly positive in 1–2 and 3–5, decaying to zero by 10–20.* The
> Phase A null was a boundary effect. The dissertation reports a
> measured crossover and compares it against Pilászy and Tikk's
> ten-rating figure and Strub et al.'s quintile profile.
>
> *Flat and near zero everywhere, including 1–2.* The null is not about
> support. The explanation lies in the features — most plausibly that a
> TF-IDF bag of words plus twenty genre indicators carries very little
> signal about a rating, which is what Kim et al. found on dense
> MovieLens. This is a stronger negative result than Phase A alone.
>
> *Gain positive but M3b still worse than M3a.* The bias channel helps
> where the item bias is unidentified; the factor channel adds 896
> parameters that the tail cannot support. Report both and say so.

---

## 9. Descriptive Comparison Against Phase A

Restricting v8's test rows to films with at least ten training ratings
gives a slice nominally comparable to Phase A. It is **not** a
controlled comparison — the v8 models were trained on a different item
set, so the factors differ — but a large discrepancy in the same
stratum would flag that lowering $N_m$ changed the head as well as
adding a tail, which is itself worth knowing.

```{r phaseA-comparison}
pa_file <- file.path(OUT_DIR, "m3c_stratified_contrasts.csv")
if (file.exists(pa_file)) {
  pa <- fread(pa_file)[contrast == "M3a - M2e" & stratum != "All"]
  pb <- boot_tbl[contrast == "M3a - M2e" &
                 stratum %in% c("10-20", "21-50", "51-200", "200+")]
  cmp <- merge(pa[, .(stratum, phaseA = delta, A_lo = lo, A_hi = hi)],
               pb[, .(stratum, phaseB = delta, B_lo = lo, B_hi = hi)],
               by = "stratum", all = TRUE)
  cmp[, stratum := factor(stratum, levels = c("10-20", "21-50", "51-200", "200+"))]
  setorder(cmp, stratum)
  kable(cmp, digits = 5,
        caption = "M3a gain in matched strata: Phase A (v7) vs Phase B (v8). Descriptive only.")
  fwrite(cmp, file.path(OUT_DIR, "m3o_phaseA_vs_phaseB.csv"))
} else {
  cat("Phase A contrasts not found; skipping.\n")
}
```

---

## 10. Save

```{r save}
saveRDS(list(metrics = metrics, train_tbl = train_tbl, scan = scan_res,
             lambda_c = LAMBDA_C, contrasts = boot_tbl, strata_n = strata_n,
             row_delta = D,
             weights = data.table(feature = colnames(Cmat), block = feat$block,
                                  w_a = m3a8$par$w, w_b = m3b8$par$w,
                                  W_norm = sqrt(rowSums(m3b8$par$W^2))),
             meta = list(built = Sys.time(), n_boot = N_BOOT, seed = SEED,
                         selection = SEL_FILE, Nm = sel$final$Nm,
                         n_users = nU, n_movies = nI, n_features = nF)),
        file.path(CACHE_M3, "m3_results_v8.rds"))
cat("06c COMPLETE.\n")
```

---

## 11. Limitations Specific to Phase B

- **Lowering $N_m$ changes the trained model, not only the evaluation
  slice.** Within-v8 stratified contrasts are valid; v7-vs-v8 stratum
  comparisons are descriptive.
- **The hyperparameters were not re-tuned for v8** beyond `lambda_c`.
  `SVD_DIM = 16` was selected on v7's denser catalogue and a much
  sparser item set might prefer fewer dimensions, which would flatter
  the content term by weakening its competitor. Stated, not corrected,
  because re-tuning would break comparability with the ladder.
- **Films with one or two training ratings are warm by the eligibility
  rule and cold in substance.** Any gain in those bands is closer to a
  cold-item result than to the warm-user framing the rest of the
  dissertation uses, and should be described that way.
- **TMDb coverage is worse in the tail**, so the sparsest bands carry
  more imputed runtimes and missing overviews than the head. The
  missingness indicators make this visible to the model but do not
  remove it as a confound between support and feature quality.
