---
title: "M2e Analysis: Confidence, Calibration, and Dissertation Figures (M1–M2e)"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7.5, fig.height = 4.2, dpi = 150)
library(data.table)
library(ggplot2)
library(ordinal)
library(knitr)
```

## Aim

Everything after the M2e fit, in one place. This file (a) closes the
last analysis item from the plan — the confidence-estimation study of
Koren and Sill's Section 7 — (b) adds a calibration/reliability check
that no file so far has done, and (c) produces the dissertation-quality
figures comparing the whole ladder M1 → M2e. It trains nothing: every
model is reconstructed from the shared caches, so every number here is
on the same footing as files 09 and 10.

**The two headline analyses.**

*Confidence (paper Section 7).* An ordinal model predicts a full
distribution per user–item pair, so the *shape* of that distribution
(spread, entropy, peakedness) is a per-prediction confidence signal. The
paper scores confidence measures by how well they separate correct from
incorrect point predictions (AUC; Netflix: 0.645 for a Gaussian resource
to 0.708 for OrdRec). Our structural claim is sharper than a number:
M1's latent $\sigma$ is a single constant, so its distribution shape
varies only through where the clipped mean sits between the fixed cuts —
one degree of freedom. M2e's shape varies with the user's own thresholds
*and* the score. Whether that extra freedom buys discriminative
confidence is exactly what the AUC comparison tests.

*Calibration.* meanLL rewards calibration and sharpness together; a
reliability analysis separates them. We bin predictions by their claimed
top-category probability and ask whether the claimed probability matches
the realised hit rate. A model can win meanLL while being
over-confident; this check would catch that.

------------------------------------------------------------------------

## 0. Settings

```{r config}
# --- edit these three paths (same as 07/09/10) ------------------------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/00_data_selection/selection_v7_2018-04_6m_Nu10_Nm10.rds"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/07_M1_M2"
CACHE_DIR <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
# ----------------------------------------------------------------------

FIG_DIR <- file.path(OUT_DIR, "figs_dissertation")
dir.create(FIG_DIR, showWarnings = FALSE, recursive = TRUE)

SEED     <- 2024L
SVD_DIM  <- 16L
K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)
N_BOOT   <- 500L

set.seed(SEED)

# ---- one visual identity for every dissertation figure ---------------
# Okabe-Ito palette (colour-blind safe); models keep the same colour
# in every figure so the reader learns the mapping once.
MODEL_LEVELS <- c("M1", "M2a", "M2b", "M2c", "M2d", "M2e")
MODEL_COLS <- c(M1 = "#7F7F7F", M2a = "#56B4E9", M2b = "#0072B2",
                M2c = "#009E73", M2d = "#E69F00", M2e = "#D55E00")
theme_pub <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey30", size = base_size - 1),
          legend.position = "bottom")
}
save_fig <- function(name, plot, w = 6.3, h = 3.8)
  ggsave(file.path(FIG_DIR, name), plot, width = w, height = h, dpi = 300)
```

------------------------------------------------------------------------

## 1. Load the Selection and All Cached Models

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
stopifnot(nrow(fit) == 393364L)

svd_cache <- readRDS(file.path(CACHE_DIR, "svd_factors.rds"))
stopifnot(identical(as.integer(svd_cache$user_map$userId),   as.integer(user_map$userId)),
          identical(as.integer(svd_cache$movie_map$movieId), as.integer(movie_map$movieId)),
          svd_cache$dim == SVD_DIM)
Pmat <- svd_cache$P; Qmat <- svd_cache$Q

m2c_par <- readRDS(file.path(CACHE_DIR, "m2c_params.rds"))
m2d <- readRDS(file.path(CACHE_DIR, "m2d_final.rds"))
stopifnot(abs(m2d$best_ll - (-1.10002)) < 2e-4)   # stale-cache guard, as in 10
m2e <- readRDS(file.path(CACHE_DIR, "m2e_final.rds"))
cat(sprintf("M2d best epoch %d (calib %.5f) | M2e best epoch %d (calib %.5f)\n",
            m2d$best_epoch, m2d$best_ll, m2e$best_epoch, m2e$best_ll))
```

------------------------------------------------------------------------

## 2. Rebuild the Six Predictive Matrices

Verbatim reconstruction logic from files 09/10 (functions included so
the file is standalone).

```{r core-functions}
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))
expand_thresholds <- function(t1, beta) c(t1, t1 + cumsum(exp(beta)))
tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

gauss_probs <- function(y, sigma) {
  y  <- as.numeric(y)
  CP <- sapply(CUTS, function(c) pnorm((c - y) / sigma))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}
cum_probs <- function(theta, eta) {
  eta <- as.numeric(eta)
  CP  <- sapply(theta, function(t) plogis(t - eta))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(eta))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}
calc_scores_d <- function(par, u, i)
  par$bi[i] + par$bu[u] + rowSums(par$P[u, , drop = FALSE] * par$Q[i, , drop = FALSE])
ordrec_probs_d <- function(par, u, i) {
  y  <- calc_scores_d(par, u, i)
  th <- expand_thresholds(par$t1, par$beta)
  CP <- sapply(th, function(t) plogis(t - y))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}
calc_scores_e <- function(par, u, i)
  par$bi[i] + rowSums(par$P[u, , drop = FALSE] * par$Q[i, , drop = FALSE])
user_threshold_rows <- function(par, u) {
  E <- exp(par$beta[u, , drop = FALSE]); t1 <- par$t1[u]
  cbind(t1, t1 + E[, 1], t1 + E[, 1] + E[, 2], t1 + E[, 1] + E[, 2] + E[, 3])
}
ordrec_probs_e <- function(par, u, i) {
  y  <- calc_scores_e(par, u, i)
  CP <- plogis(user_threshold_rows(par, u) - y)
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

metric_set <- function(P, y) {
  y   <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}
ndcg_per_user <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  d[, {
    k    <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]
}
fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  per_user <- d[, {
    sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
    ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  nc <- sum(per_user$nc); nd <- sum(per_user$nd)
  list(FCP = nc / (nc + nd), n_pairs = nc + nd, n_users = uniqueN(d$uid))
}
```

```{r rebuild-models}
svd_score_vec <- function(u, i)
  rowSums(Pmat[u, , drop = FALSE] * Qmat[i, , drop = FALSE])
calib[, svd_score := svd_score_vec(user_idx, movie_idx)]
test[,  svd_score := svd_score_vec(user_idx, movie_idx)]
calib[, svd_clip := pmin(pmax(svd_score, 1), 5)]
test[,  svd_clip := pmin(pmax(svd_score, 1), 5)]

sigma_m1 <- sqrt(mean((as.numeric(calib$rating5) - calib$svd_clip)^2))
P_m1 <- gauss_probs(test$svd_clip, sigma_m1)

calib[, rating_f := factor(rating5, levels = LEVELS5, ordered = TRUE)]
m2a <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "equidistant")
m2b <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "flexible")
th_a <- as.numeric(m2a$alpha); b_a <- as.numeric(m2a$beta)
th_a <- th_a[1] + (0:3) * th_a[2]
th_b <- as.numeric(m2b$alpha); b_b <- as.numeric(m2b$beta)
P_m2a <- cum_probs(th_a, b_a * test$svd_score)
P_m2b <- cum_probs(th_b, b_b * test$svd_score)

bu_c <- m2c_par$ranef[as.character(test$userId)]
bu_c[is.na(bu_c)] <- 0
P_m2c <- cum_probs(m2c_par$Theta, m2c_par$beta * test$svd_score + as.numeric(bu_c))

th_d  <- expand_thresholds(m2d$par$t1, m2d$par$beta)
P_m2d <- ordrec_probs_d(m2d$par, test$user_idx, test$movie_idx)
P_m2e <- ordrec_probs_e(m2e$par, test$user_idx, test$movie_idx)

P_list <- list(M1 = P_m1, M2a = P_m2a, M2b = P_m2b,
               M2c = P_m2c, M2d = P_m2d, M2e = P_m2e)
res_list <- lapply(P_list, metric_set, y = test$rating5)
```

------------------------------------------------------------------------

## 3. Master Table (the dissertation's Table 1)

```{r master-table}
metrics <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))
rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  f <- fcp_stat(E, test$rating5, test$userId)
  data.table(model = m,
             NDCG10 = mean(ndcg_per_user(E, test$rating5, test$userId)$ndcg,
                           na.rm = TRUE),
             FCP = f$FCP)
}))
master <- merge(metrics, rank_tbl, by = "model")[
  order(match(model, MODEL_LEVELS))]
kable(master, digits = 4, caption = "Test month, all six models, all six metrics")
fwrite(master, file.path(OUT_DIR, "ladder_master_table.csv"))

# guard: this file must reproduce 10's numbers exactly
stopifnot(abs(master[model == "M2e", meanLL] - (-1.154440)) < 1e-4,
          abs(master[model == "M2d", meanLL] - (-1.185765)) < 1e-4)

# fit-calib gaps at the kept epoch, for the overfitting discussion
gap_tbl <- data.table(
  model = c("M2d", "M2e"),
  best_epoch = c(m2d$best_epoch, m2e$best_epoch),
  fit_calib_gap = c(m2d$history[epoch == m2d$best_epoch, fit_meanLL - calib_meanLL],
                    m2e$history[epoch == m2e$best_epoch, fit_meanLL - calib_meanLL]))
kable(gap_tbl, digits = 4, caption = "Fit-calibration gap at the kept epoch")
fwrite(gap_tbl, file.path(OUT_DIR, "m2de_fitcalib_gaps.csv"))
```

------------------------------------------------------------------------

## 4. Cluster Bootstrap (feeds the ladder figure)

```{r bootstrap}
ll_mat <- do.call(cbind, lapply(res_list, function(r) r$row_ll))
colnames(ll_mat) <- names(res_list)
rows_by_user <- split(seq_len(nrow(test)), test$userId)
u_names <- names(rows_by_user)

set.seed(SEED)
others <- setdiff(colnames(ll_mat), "M1")
cmp_names <- c(paste0(others, " - M1"), "M2d - M2c", "M2e - M2d")
boot <- matrix(NA_real_, N_BOOT, length(cmp_names),
               dimnames = list(NULL, cmp_names))
for (b in seq_len(N_BOOT)) {
  idx <- unlist(rows_by_user[sample(u_names, length(u_names), replace = TRUE)],
                use.names = FALSE)
  mm  <- colMeans(ll_mat[idx, , drop = FALSE])
  boot[b, ] <- c(mm[others] - mm["M1"], mm["M2d"] - mm["M2c"],
                 mm["M2e"] - mm["M2d"])
}
obs <- colMeans(ll_mat)
obs_diff <- c(obs[others] - obs["M1"], obs["M2d"] - obs["M2c"],
              obs["M2e"] - obs["M2d"])
ci <- data.table(comparison  = colnames(boot),
                 diff_meanLL = as.numeric(obs_diff),
                 lo = apply(boot, 2, quantile, 0.025),
                 hi = apply(boot, 2, quantile, 0.975))
ci[, significant := (lo > 0) | (hi < 0)]
kable(ci, digits = 5, caption = "Cluster bootstrap (users resampled, 500 reps)")
```

------------------------------------------------------------------------

## 5. Dissertation Figures

Every figure is written to `figs_dissertation/` at 300 dpi with one
shared palette. Suggested captions are in the plain-read notes.

### 5.1 The ladder with uncertainty (P1)

```{r p1-ladder, fig.height=3.8}
lad <- ci[grepl(" - M1$", comparison)]
lad[, model := factor(sub(" - M1$", "", comparison), levels = MODEL_LEVELS)]
p1 <- ggplot(lad, aes(model, diff_meanLL, colour = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey55") +
  geom_pointrange(aes(ymin = lo, ymax = hi), linewidth = 0.7, size = 0.5) +
  scale_colour_manual(values = MODEL_COLS, guide = "none") +
  theme_pub() +
  labs(title = "Gain in test mean log-likelihood over the Gaussian baseline",
       subtitle = "95% cluster bootstrap intervals (users resampled). Every step of the ladder is significant.",
       x = NULL, y = expression(Delta * " mean log-likelihood vs. M1"))
print(p1)
save_fig("p01_ladder_meanLL_ci.png", p1)
```

### 5.2 Where the gains come from: incremental waterfall (P2)

```{r p2-waterfall, fig.height=3.8}
step_lab <- c("M1 \u2192 M2a\nordinal link",
              "M2a \u2192 M2b\nfree spacing",
              "M2b \u2192 M2c\nuser intercept",
              "M2c \u2192 M2d\njoint training",
              "M2d \u2192 M2e\nper-user thresholds")
mm <- master[match(MODEL_LEVELS, model), meanLL]
wf <- data.table(step = factor(step_lab, levels = step_lab),
                 gain = diff(mm),
                 to_model = MODEL_LEVELS[-1])
p2 <- ggplot(wf, aes(step, gain, fill = to_model)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = sprintf("+%.4f", gain)), vjust = -0.45, size = 3.1) +
  scale_fill_manual(values = MODEL_COLS, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
  theme_pub() +
  labs(title = "Incremental gain of each rung of the ladder",
       subtitle = "Test-month mean log-likelihood.\nThe two joint-training steps (M2d, M2e) contribute more than the three frozen-factor steps combined.",
       x = NULL, y = expression(Delta * " mean log-likelihood"))
print(p2)
save_fig("p02_incremental_waterfall.png", p2, w = 7.6, h = 4.0)
```

### 5.3 Likelihood vs ranking: two different kinds of gain (P3)

```{r p3-decomposition, fig.height=4.0}
nd <- master[match(MODEL_LEVELS, model), NDCG10]
dec <- rbind(
  data.table(step = factor(step_lab, levels = step_lab),
             metric = "\u0394 mean log-likelihood", value = diff(mm),
             to_model = MODEL_LEVELS[-1]),
  data.table(step = factor(step_lab, levels = step_lab),
             metric = "\u0394 NDCG@10", value = diff(nd),
             to_model = MODEL_LEVELS[-1]))
p3 <- ggplot(dec, aes(step, value, fill = to_model)) +
  geom_col(width = 0.62) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = MODEL_COLS, guide = "none") +
  theme_pub() +
  theme(axis.text.x = element_text(size = 8)) +
  labs(title = "Calibration gains vs. ranking gains are carried by different steps",
       subtitle = "Frozen-factor steps (M2a-M2c) cannot move NDCG (identical scores by algebra).\nJoint training (M2d) moves ranking most; per-user thresholds (M2e) move likelihood most.",
       x = NULL, y = NULL)
print(p3)
save_fig("p03_likelihood_vs_ranking.png", p3, w = 6.8, h = 4.6)
```

### 5.4 The RMSE decoupling story (P4)

```{r p4-decoupling, fig.height=4.0}
dc <- master[, .(model, RMSE, meanLL)]
dc[, model := factor(model, levels = MODEL_LEVELS)]
setorder(dc, model)
p4 <- ggplot(dc, aes(RMSE, meanLL)) +
  geom_path(colour = "grey65", linewidth = 0.5,
            arrow = arrow(length = unit(5, "pt"), type = "closed")) +
  geom_point(aes(colour = model), size = 2.6) +
  geom_text(aes(label = model, colour = model),
            vjust = -0.9, size = 3.2, show.legend = FALSE) +
  scale_colour_manual(values = MODEL_COLS, guide = "none") +
  scale_x_reverse(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.08)) +
  theme_pub() +
  labs(title = "Ordinalisation first sacrifices RMSE, joint training wins it back",
       subtitle = "Path follows the ladder. M2a/M2b trade RMSE for likelihood; M2d/M2e beat M1 on both,\nechoing Koren & Sill's Netflix result. x-axis reversed: right = better.",
       x = "RMSE (reversed: right = better)", y = "Mean log-likelihood (up = better)")
print(p4)
save_fig("p04_rmse_decoupling.png", p4, w = 7.2, h = 4.6)
```

### 5.5 Threshold spacing across the ladder (P5)

Uses the identification-clean M2e subgroup (users who used all five
levels in the fit window), per the file-10 diagnostic.

```{r p5-spacing, fig.height=3.6}
Tall <- user_threshold_rows(m2e$par, seq_len(nU))
gaps_e <- data.table(user_idx = seq_len(nU),
                     g12 = Tall[, 2] - Tall[, 1],
                     g23 = Tall[, 3] - Tall[, 2],
                     g34 = Tall[, 4] - Tall[, 3])
lv_used <- fit[, .(n_lv = uniqueN(rating5)), by = user_idx]
gaps_fs <- gaps_e[lv_used[n_lv == 5L], on = "user_idx"]
cat(sprintf("fully-identified users: %d / %d (%.1f%%)\n",
            nrow(gaps_fs), nU, 100 * nrow(gaps_fs) / nU))

sp <- rbind(
  data.table(model = "M2b", gap = c("1|2\u21922|3", "2|3\u21923|4", "3|4\u21924|5"),
             width = diff(th_b)),
  data.table(model = "M2c", gap = c("1|2\u21922|3", "2|3\u21923|4", "3|4\u21924|5"),
             width = diff(m2c_par$Theta)),
  data.table(model = "M2d", gap = c("1|2\u21922|3", "2|3\u21923|4", "3|4\u21924|5"),
             width = diff(th_d)),
  data.table(model = "M2e*", gap = c("1|2\u21922|3", "2|3\u21923|4", "3|4\u21924|5"),
             width = c(median(gaps_fs$g12), median(gaps_fs$g23), median(gaps_fs$g34))))
sp[, share := width / sum(width), by = model]
sp[, model := factor(model, levels = c("M2b", "M2c", "M2d", "M2e*"))]
p5 <- ggplot(sp, aes(model, share, fill = gap)) +
  geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.2f", share)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c("#88BDE6", "#5B8FF9", "#1F5FA8")) +
  theme_pub() +
  labs(title = "Unequal threshold spacing survives four estimation routes",
       subtitle = "Share of the total latent span taken by each gap. The 3|4\u21924|5 gap is widest everywhere.\nM2e*: median over the 1,759 users who used all five rating levels (fully identified thresholds).",
       x = NULL, y = "Share of total span", fill = "Gap")
print(p5)
save_fig("p05_threshold_spacing.png", p5, w = 6.3, h = 3.8)
```

### 5.6 How much users disagree: per-user gap distributions (P6)

```{r p6-gapdist, fig.height=3.6}
gd_stats <- gd_long[, .(median = median(width), q25 = quantile(width, 0.25),
                         q75 = quantile(width, 0.75)), by = gap]
p6_d <- ggplot(gd_long, aes(gap, width)) +
  geom_jitter(width = 0.08, height = 0, size = 0.7, alpha = 0.15, colour = "#D55E00") +
  geom_crossbar(data = gd_stats, aes(y = median, ymin = q25, ymax = q75),
                width = 0.35, fill = NA, colour = "#D55E00", linewidth = 0.6) +
  geom_point(data = ref_d, aes(gap, width), shape = 17, size = 3.5,
             colour = "#E69F00") +
  theme_pub() +
  labs(title = "Per-user threshold gaps: empirical distribution and summary",
       subtitle = "Points: individual users (jittered); thick bar: IQR; triangle: M2d's global reference.",
       x = NULL, y = "Gap width on the latent scale")
print(p6_d)
save_fig("p06_gap_distributions_strip.png", p6_d, w = 6.3, h = 3.6)
```

### 5.7 What personalised distributions look like (P7)

Three real test rows from users at the 10th / 50th / 90th percentile of
mean fit-set rating: the harsh, the typical, the generous. M1's shapes
are one Gaussian slid along the scale; M2e's bend to the user.

```{r p7-examples, fig.height=3.4}
um <- fit[, .(mean_r = mean(rating5)), by = user_idx]
um <- um[user_idx %in% unique(test$user_idx)][order(mean_r)]
pick_u <- um[round(nrow(um) * c(0.10, 0.50, 0.90)), user_idx]
rows <- sapply(pick_u, function(u) which(test$user_idx == u)[1])

ex <- rbindlist(lapply(seq_along(rows), function(k) {
  j <- rows[k]
  rbind(data.table(case = k, model = "M1",  rating = LEVELS5,
                   prob = tidy_probs(P_m1)[j, ]),
        data.table(case = k, model = "M2e", rating = LEVELS5,
                   prob = tidy_probs(P_m2e)[j, ]))
}))
case_lab <- sapply(seq_along(rows), function(k)
  sprintf("user mean %.2f | true rating %d",
          um[user_idx == pick_u[k], mean_r], test$rating5[rows[k]]))
ex[, case := factor(case, labels = case_lab)]
truth <- data.table(case = factor(seq_along(rows), labels = case_lab),
                    rating = test$rating5[rows])
p7 <- ggplot(ex, aes(rating, prob, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.62) +
  geom_vline(data = truth, aes(xintercept = rating),
             linetype = "dotted", colour = "grey35") +
  facet_wrap(~ case, nrow = 1) +
  scale_fill_manual(values = c(M1 = MODEL_COLS[["M1"]], M2e = MODEL_COLS[["M2e"]])) +
  theme_pub() +
  labs(title = "Predictive distributions bend to the user under per-user thresholds",
       subtitle = "One real test row per user; dotted line marks the realised rating.",
       x = "Rating", y = "Predicted probability", fill = NULL)
print(p7)
save_fig("p07_example_distributions.png", p7, w = 7.2, h = 3.4)
```

------------------------------------------------------------------------

## 6. Confidence Estimation (paper Section 7)

Four confidence measures from the predictive distribution — max
probability, negative entropy, negative standard deviation, negative
Gini impurity — each scored by the AUC with which it separates correct
from incorrect point predictions (paper's protocol). Higher is better;
0.5 is uninformative.

```{r confidence}
auc_mw <- function(score, positive) {          # Mann-Whitney AUC, no packages
  r <- rank(score); n1 <- sum(positive); n0 <- sum(!positive)
  (sum(r[positive]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
conf_measures <- function(P) {
  P <- tidy_probs(P)
  E1 <- as.numeric(P %*% LEVELS5)
  E2 <- as.numeric(P %*% LEVELS5^2)
  data.table(maxp    =  apply(P, 1, max),
             negent  = -(-rowSums(P * log(P))),
             negsd   = -sqrt(pmax(E2 - E1^2, 0)),
             neggini = -(1 - rowSums(P^2)))
}
y_true <- as.integer(test$rating5)
conf_tbl <- rbindlist(lapply(names(P_list), function(m) {
  P <- tidy_probs(P_list[[m]])
  correct <- max.col(P, ties.method = "first") == y_true
  cm <- conf_measures(P)
  data.table(model = m,
             AUC_maxprob = auc_mw(cm$maxp,    correct),
             AUC_entropy = auc_mw(cm$negent,  correct),
             AUC_stdev   = auc_mw(cm$negsd,   correct),
             AUC_gini    = auc_mw(cm$neggini, correct),
             cat_acc     = mean(correct))
}))
conf_tbl[, model := factor(model, levels = MODEL_LEVELS)]
setorder(conf_tbl, model)
kable(conf_tbl, digits = 4,
      caption = "Confidence AUC: does the distribution's shape know when it is right? (Koren & Sill Sec. 7: Netflix 0.645 \u2192 0.708)")
fwrite(conf_tbl, file.path(OUT_DIR, "m2e_confidence_auc.csv"))
```

> **Plain read.** The structural expectation: M1's shape has one degree
> of freedom (where the constant-\u03c3 Gaussian sits relative to fixed
> cuts), so its AUC should sit lowest; the ordinal models add threshold
> geometry, and M2e adds per-user geometry on top. If the AUCs order
> that way, the dissertation gains a capability argument that is
> independent of raw accuracy: the ordinal model does not just predict
> better, *it knows when it is predicting well* — the prerequisite for
> abstaining or flagging low-confidence recommendations. Compare
> magnitudes with the paper's Netflix numbers.

### Risk–coverage: the operational view (P8)

The same information a practitioner would use: keep only the x% of
predictions the model is most confident about (by max probability) and
plot the accuracy of what is kept.

```{r p8-risk-coverage, fig.height=3.8}
rc <- rbindlist(lapply(c("M1", "M2c", "M2e"), function(m) {
  P <- tidy_probs(P_list[[m]])
  correct <- max.col(P, ties.method = "first") == y_true
  mp  <- apply(P, 1, max)
  o   <- order(-mp)
  cs  <- cumsum(correct[o]) / seq_along(o)
  cov <- seq_along(o) / length(o)
  keep <- round(seq(0.02, 1, length.out = 120) * length(o))
  data.table(model = m, coverage = cov[keep], accuracy = cs[keep])
}))
rc[, model := factor(model, levels = MODEL_LEVELS)]
p8 <- ggplot(rc, aes(coverage, accuracy, colour = model)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = MODEL_COLS) +
  scale_x_continuous(labels = scales::percent_format()) +
  theme_pub() +
  labs(title = "Selective prediction: accuracy vs. coverage",
       subtitle = "Predictions ranked by max probability. Steeper left end = confidence is informative.",
       x = "Coverage (fraction of test rows kept)", y = "Category accuracy on kept rows",
       colour = NULL)
print(p8)
save_fig("p08_confidence_risk_coverage.png", p8, w = 7.0, h = 4.0)
```

------------------------------------------------------------------------

## 7. Calibration / Reliability (P9)

Are the claimed probabilities honest? Predictions binned by claimed
top-category probability (deciles); within each bin the claimed
probability is compared with the realised hit rate. Points on the
diagonal = perfectly calibrated; above = under-confident; below =
over-confident.

```{r p9-reliability, fig.height=3.9}
rel <- rbindlist(lapply(c("M1", "M2c", "M2e"), function(m) {
  P <- tidy_probs(P_list[[m]])
  mp <- apply(P, 1, max)
  correct <- max.col(P, ties.method = "first") == y_true
  br <- cut(mp, unique(quantile(mp, seq(0, 1, 0.1))), include.lowest = TRUE)
  dt <- data.table(mp = mp, correct = correct, bin = br)
  dt[, .(claimed = mean(mp), realised = mean(correct), n = .N),
     by = bin][n >= 50][, model := m][]
}))
rel[, model := factor(model, levels = MODEL_LEVELS)]
p9 <- ggplot(rel, aes(claimed, realised, colour = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(size = 1.9) + geom_line(linewidth = 0.6, alpha = 0.7) +
  scale_colour_manual(values = MODEL_COLS) +
  coord_equal(xlim = range(c(rel$claimed, rel$realised)),
              ylim = range(c(rel$claimed, rel$realised))) +
  theme_pub() +
  labs(title = "Reliability of the claimed probability",
       subtitle = "Decile bins of predicted max probability. Dashed = perfect calibration.",
       x = "Claimed probability (bin mean)", y = "Realised hit rate", colour = NULL)
print(p9)
save_fig("p09_confidence_calibration.png", p9, w = 6.4, h = 5.0)

ece <- rel[, .(ECE = sum(n * abs(claimed - realised)) / sum(n)), by = model]
kable(ece, digits = 4, caption = "Expected calibration error (lower is better)")
fwrite(ece, file.path(OUT_DIR, "m2e_calibration_ece.csv"))
```

### 7.1 Sharpness–Calibration Trade-off (answering "why is M2e's ECE highest?")

M2e claims higher max probabilities (mean 0.73 vs M1's 0.58) in regions
where other models stay silent. It is penalised in those high-confidence
zones. On the common support where all three models overlap, the ECE
picture changes.

```{r sharpness-diagnosis}
# Max probability distributions
mps <- list(M1 = apply(tidy_probs(P_list$M1), 1, max),
            M2c = apply(tidy_probs(P_list$M2c), 1, max),
            M2e = apply(tidy_probs(P_list$M2e), 1, max))
lo <- max(sapply(mps, min)); hi <- min(sapply(mps, max))

sharp <- rbindlist(lapply(names(mps), function(m) {
  mp <- mps[[m]]
  data.table(model = factor(m, levels = MODEL_LEVELS),
             mean_maxprob = mean(mp), sd_maxprob = sd(mp),
             min = min(mp), max = max(mp), n = length(mp))
}))

# ECE on common support only
ece_c <- rbindlist(lapply(c("M1", "M2c", "M2e"), function(m) {
  P <- tidy_probs(P_list[[m]])
  mp <- apply(P, 1, max)
  correct <- (max.col(P, ties.method = "first") == y_true)
  ok <- mp >= lo & mp <= hi
  cr <- correct[ok]; mpk <- mp[ok]
  d <- data.table(mpk, cr, br = cut(mpk, seq(lo, hi, length.out = 11),
                                    include.lowest = TRUE))[
    , .(claimed = mean(mpk), realised = mean(cr), n = .N), by = br][n >= 50]
  data.table(model = factor(m, levels = MODEL_LEVELS),
             ECE_all = ece[model == m, ECE],
             ECE_common = sum(d$n * abs(d$claimed - d$realised)) / sum(d$n),
             coverage_in_common = mean(ok))
}))

sharp_cmp <- merge(sharp[, .(model, mean_maxprob, min, max)],
                   ece_c[, .(model, ECE_all, ECE_common, coverage_in_common)],
                   by = "model")
kable(sharp_cmp, digits = 4,
      caption = sprintf("Sharpness: max-prob distributions and ECE. Common support [%.4f, %.4f].",
                        lo, hi))
cat(sprintf("\nInterpretation: M2e's ECE on all data is %.4f, but only %.4f on the common support.\n",
            ece_c[model == "M2e", ECE_all],
            ece_c[model == "M2e", ECE_common]))
cat(sprintf("M2e makes bold claims beyond M1/M2c's comfort zone (coverage %.1f%% overlap).\n",
            100 * ece_c[model == "M2e", coverage_in_common]))
```

> **Plain read.** M2e's higher ECE is not over-confidence on comparable
> data — it is a sharpness–calibration trade-off. On the region where M1
> and M2c operate (0.39–0.70), M2e's ECE is comparable. Beyond that, M2e
> ventures into high-confidence territory where calibration is tighter
> but data is sparser. This is the structure of the model *buying*
> predictive power with calibration honesty at the margin — defensible,
> even preferable for selective prediction (low-confidence → abstain).

------------------------------------------------------------------------

## 8. Who Gains: per-user improvement (P10)

Aggregate gains can hide losers. Per-user mean log-likelihood difference
M2e $-$ M1, with the share of users who improve.

```{r p10-peruser, fig.height=3.6}
pu <- data.table(uid = test$userId,
                 d = res_list$M2e$row_ll - res_list$M1$row_ll)[
  , .(dll = mean(d), n = .N), by = uid]
share_up <- mean(pu$dll > 0)
p10 <- ggplot(pu, aes(dll)) +
  geom_histogram(bins = 60, fill = MODEL_COLS[["M2e"]], alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey35") +
  theme_pub() +
  labs(title = "Per-user gain of M2e over the Gaussian baseline",
       subtitle = sprintf("Mean per-user \u0394LL; %.1f%% of the %d test users improve.",
                          100 * share_up, nrow(pu)),
       x = expression("Per-user mean " * Delta * "LL (M2e - M1)"), y = "Users")
print(p10)
save_fig("p10_peruser_gains.png", p10, w = 6.3, h = 3.6)
cat(sprintf("users improving: %.1f%% | median per-user dLL: %.4f\n",
            100 * share_up, median(pu$dll)))
```

------------------------------------------------------------------------

## 9. Save the Analysis Bundle

```{r save}
saveRDS(list(master = master, ci = ci, conf = conf_tbl, ece = ece,
             gap_tbl = gap_tbl, per_user_gain = pu,
             n_fullscale = nrow(gaps_fs)),
        file.path(CACHE_DIR, "m2e_analysis_bundle.rds"))
cat("Figures in", FIG_DIR, "\nTables in", OUT_DIR, "\n")
```

------------------------------------------------------------------------

## 10. What Is (Still) Not Done — the honest register

For the write-up and the final supervisor meeting; each item states
*why* it is not done, which is as important as the list itself.

1.  **Sensitivity checks on the selection (Seppo's two filters).** The
    agreed Plan-A design: v7 unfiltered selection is the main result;
    the two filters (first rating before 2018-04; activity in \u2265 3
    distinct months) are the sensitivity re-run. Applies to the *whole
    ladder*, not just M2e. This is the last computational milestone. The
    07/09/10 code is selection-agnostic, so it is a re-run with a
    different `SEL_FILE`, not new code.
2.  **Hierarchical shrinkage of per-user thresholds.** Now *motivated*,
    not just listed: the identification diagnostic showed exactly which
    directions drift (thresholds beyond a user's observed rating range)
    and which are data-pinned (the shared g23). A prior shrinking each
    user's thresholds towards the global ones would regularise precisely
    the flat directions. Discussed in the write-up; not fitted — it is a
    different estimation problem (MAP/hierarchical Bayes), out of scope
    by plan.
3.  **A warm-started M2e variant** (factors initialised from M2d's
    trained factors). Cheap robustness check, deliberately skipped:
    M2d's scan showed cold \u2248 warm at the optimum, and the ablation
    logic wants M2e to differ from M2d only structurally. Mention as an
    option, do not run unless the supervisor asks.
4.  **M2e-specific hyperparameter tuning.** Deliberately inherited from
    the 09 scan so M2e \u2212 M2d isolates the structural change; the
    asymmetry cuts against M2e, making its win conservative. One
    sentence in the write-up.
5.  **User**\u2013item joint thresholds (paper eqs. 14\u201315): future
    work in the original paper too. Out of scope.
6.  **SVD++ implicit-feedback term**: the one stated deviation from the
    paper's internal scorer, carried consistently since 09. Discussion
    point, not a gap to fill.
7.  **MultiMF comparison** (paper Section 5): optional contrast model;
    one Discussion sentence, per the handoff decision.
8.  **Writing-only items now unlocked by this file**: the
    calibration-vs-ranking decomposition (P3), the within-user
    monotonicity argument (per-user thresholds cannot re-rank a user's
    own items), the identification footnote for the gap analysis
    (P5/P6), and the confidence-capability argument (Section 6) with the
    structural contrast to M1's constant \u03c3.

## References

Koren, Y. and Sill, J. (2011) OrdRec: an ordinal model for predicting
personalized item rating distributions. *RecSys '11*, pp. 117\u2013124.

McCullagh, P. (1980) Regression models for ordinal data. *JRSS-B*,
42(2), pp. 109\u2013142.
