---
title: "M1 and M2: First Run on the New Data Selection"
author: "Dina Muhamaitihan | Movie Recommendation | Meeting 4"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(ordinal)
library(recosystem)
library(knitr)
```

## Aim

This file runs the first four steps of the model ladder on the new data
selection (2018-04 to 2018-09 training, 2018-10 test, 5 rating categories).

| Step | Model | Thresholds | User effect |
|------|-------|-----------|-------------|
| M1  | SVD + Gaussian | fixed, equal | no |
| M2a | Two-step CLM, equal spacing | global, spacing estimated | no |
| M2b | Two-step CLM, free | global, all estimated | no |
| M2c | Two-step CLM + user random intercept | global | yes, random |

M2d and M2e (joint OrdRec) are not here. They need stochastic gradient
ascent on the ordinal likelihood and are the next piece of work. The point
of this file is to have the baseline, the metrics code and the evaluation
protocol working and checked before that.

The SVD is trained on the fit set. All calibration parameters (sigma,
thresholds, random intercepts) are estimated on the calibration set. The
test month is used only at the end.

---

## 0. Settings

```{r config}
# --- edit these three paths -------------------------------------------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/00_data_selection/selection_v7_2018-04_6m_Nu10_Nm10.rds"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/07_M1_M2"
CACHE_DIR <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
# ----------------------------------------------------------------------

dir.create(OUT_DIR,   showWarnings = FALSE, recursive = TRUE)
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

SEED         <- 2024L
SVD_DIM      <- 16L    # confirmed in section 3
RUN_DIM_SCAN <- TRUE
RUN_M2C      <- TRUE   # set FALSE if clmm2 is too slow
N_BOOT       <- 500L

K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)

n_cores <- parallel::detectCores()
set.seed(SEED)
```

---

## 1. Load the Selection

The data selection script has already done the cleaning, the window choice,
the eligibility thresholds and the stratified split. This file only reads
its output.

One conversion is needed. The selection stores `rating5` as an **ordered
factor**, which is what `clm` wants but not what arithmetic wants. It is
converted to an integer here, once, and a factor version is rebuilt later
where the CLM needs it.

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)
n_raw_test <- nrow(sel$test_raw)

cat("Class of rating5 as stored:", paste(class(fit$rating5), collapse = ", "), "\n")

# --- the one conversion that matters ---------------------------------
to_int5 <- function(d) {
  d[, rating5 := as.integer(as.character(rating5))]
  d[]
}
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

cat("Class after conversion:", class(fit$rating5), "\n")
cat("Categories present:", paste(sort(unique(fit$rating5)), collapse = ", "), "\n\n")

sizes <- data.table(
  item  = c("Fit rows", "Calibration rows", "Test rows evaluated",
            "Users (training window)", "Movies (training window)",
            "Users in test", "Test coverage (%)"),
  value = c(nrow(fit), nrow(calib), nrow(test),
            uniqueN(c(fit$userId,  calib$userId)),
            uniqueN(c(fit$movieId, calib$movieId)),
            uniqueN(test$userId),
            round(100 * nrow(test) / n_raw_test, 1)),
  expected = c(393364, 98338, 16694, 5611, 5046, NA, 16.2))
kable(sizes, caption = "Loaded selection, against the numbers in the data selection document")
```

```{r distribution-check}
kl <- function(p, q) sum(p * log(p / q))
p_fit   <- prop.table(table(factor(fit$rating5,   levels = LEVELS5)))
p_calib <- prop.table(table(factor(calib$rating5, levels = LEVELS5)))
p_test  <- prop.table(table(factor(test$rating5,  levels = LEVELS5)))

cat(sprintf("KL(fit || calib) = %.4f   (stratified, so near zero)\n",
            kl(as.numeric(p_fit), as.numeric(p_calib))))
cat(sprintf("KL(fit || test)  = %.4f   (real drift in the test month)\n",
            kl(as.numeric(p_fit), as.numeric(p_test))))

dist_dt <- rbind(
  data.table(set = "fit",   rating5 = LEVELS5, p = as.numeric(p_fit)),
  data.table(set = "calib", rating5 = LEVELS5, p = as.numeric(p_calib)),
  data.table(set = "test",  rating5 = LEVELS5, p = as.numeric(p_test)))

p_dist <- ggplot(dist_dt, aes(factor(rating5), p, fill = set)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(fit = "steelblue", calib = "goldenrod",
                               test = "coral")) +
  theme_minimal() +
  labs(title = "Rating distribution in the three sets",
       subtitle = "Fit and calibration are stratified, so they match. The test month drifts.",
       x = "Rating category", y = "Share", fill = NULL)
print(p_dist)
ggsave(file.path(OUT_DIR, "fig1_rating_distribution.png"), p_dist,
       width = 8, height = 4.5, dpi = 150)
```

> **Plain read.** Fit and calibration have the same rating distribution, so
> the calibration set is a fair place to estimate thresholds. The test month
> has more low ratings and fewer 5s. Every model is judged against the same
> drifted month.

---

## 2. Indices and Helper Functions

Everything the rest of the file needs is defined here, in one block, so the
chunks below can be run in any order after this one.

```{r setup-all}
# ---- integer indices, built from the fit set only --------------------
user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]

add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit <- add_idx(fit); calib <- add_idx(calib); test <- add_idx(test)

cat(sprintf("Index maps: %d users, %d movies\n", nrow(user_map), nrow(movie_map)))
cat(sprintf("Cold rows: fit %.2f%% | calib %.2f%% | test %.2f%%\n",
            100 * mean(fit$cold), 100 * mean(calib$cold), 100 * mean(test$cold)))

fit <- fit[cold == FALSE]; calib <- calib[cold == FALSE]; test <- test[cold == FALSE]
cat(sprintf("After dropping cold: fit %d | calib %d | test %d\n",
            nrow(fit), nrow(calib), nrow(test)))

# ---- SVD ------------------------------------------------------------
# l2 defaults to 0.01 so the existing dim-scan chunk (which doesn't pass l2)
# keeps its original meaning: dim was chosen at a fixed default L2. The L2
# scan (new section 3b) then tunes L2 at that dim. Sequential, not a joint
# grid search - a stated simplification, not an oversight.
train_svd <- function(d, dim, l2 = 0.01) {
  set.seed(SEED)
  r <- Reco()
  r$train(data_memory(d$user_idx, d$movie_idx, as.numeric(d$rating5), index1 = TRUE),
          opts = list(dim = dim, costp_l2 = l2, costq_l2 = l2,
                      lrate = 0.1, niter = 20,
                      nthread = n_cores, verbose = FALSE))
  r
}
score_with <- function(r, d) {
  r$predict(data_memory(d$user_idx, d$movie_idx, index1 = TRUE), out_memory())
}

# ---- probability helpers --------------------------------------------
tidy_probs <- function(P) {
  P <- pmax(P, 1e-12)
  P / rowSums(P)
}

# Gaussian around the score, integrated between the category boundaries
gauss_probs <- function(y, sigma) {
  y  <- as.numeric(y)
  CP <- sapply(CUTS, function(c) pnorm((c - y) / sigma))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# Logistic cumulative link: P(Y <= k) = plogis(theta_k - eta)
cum_probs <- function(theta, eta) {
  eta <- as.numeric(eta)
  CP  <- sapply(theta, function(t) plogis(t - eta))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(eta))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- metrics ---------------------------------------------------------
metric_set <- function(P, y) {
  y   <- as.integer(y)
  P   <- tidy_probs(P)
  n   <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

# NDCG@10 within each test user, ranking by the model's expected rating
ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  per_user <- d[, {
    k    <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]
  mean(per_user$ndcg, na.rm = TRUE)
}

# ---- FCP: Fraction of Concordant Pairs (Koren & Sill 2011, eq. 24-26) -----
# For each user, and every ordered pair of items (a, b) with a != b:
#   n_c += 1 if predicted score(a) >  predicted score(b)  AND  true(a) >  true(b)
#   n_d += 1 if predicted score(a) >= predicted score(b)  AND  true(a) <  true(b)
# The asymmetry (n_d uses >=) is in the paper, not a choice we made: a
# predicted tie on a pair with different true ratings always counts as
# discordant, it is never a free pass. Pairs with equal true ratings
# contribute to neither n_c nor n_d, so they drop out of the denominator.
#
# min_items mirrors ndcg_at_10's default, so both ranking metrics are
# computed on the same set of test users unless overridden.
fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) {
    return(list(FCP = NA_real_, nc = 0L, nd = 0L, n_pairs = 0L, n_users = 0L))
  }
  per_user <- d[, {
    sy_gt <- outer(y, y, `>`)    # true(a)  > true(b)
    sy_lt <- outer(y, y, `<`)    # true(a)  < true(b)
    ss_gt <- outer(s, s, `>`)    # pred(a)  > pred(b)
    ss_ge <- outer(s, s, `>=`)   # pred(a) >= pred(b)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  nc <- sum(per_user$nc); nd <- sum(per_user$nd)
  list(FCP = nc / (nc + nd), nc = nc, nd = nd,
       n_pairs = nc + nd, n_users = uniqueN(d$uid))
}

cat("FCP self-test:\n")
# Test 1: Basic correctness on a single user with 4 items
# (explicit min_items=1 to bypass the filter)
toy <- fcp_stat(score = c(4.0, 2.0, 2.5, 1.0),
                y     = c(5,   4,   3,   3),
                uid   = rep("u1", 4),
                min_items = 1L)
cat(sprintf("  Test 1 (min_items=1): nc=%d nd=%d FCP=%.4f  (expect nc=4 nd=1 FCP=0.8000)\n",
            toy$nc, toy$nd, toy$FCP))
stopifnot(toy$nc == 4L, toy$nd == 1L, abs(toy$FCP - 0.8) < 1e-9)

# Test 2: Filtering - the same user, but now with default min_items=5.
# Since the user has only 4 items, should be dropped entirely.
toy_filtered <- fcp_stat(score = c(4.0, 2.0, 2.5, 1.0),
                         y     = c(5,   4,   3,   3),
                         uid   = rep("u1", 4))  # default min_items=5
cat(sprintf("  Test 2 (min_items=5): n_users=%d (expect 0, user has only 4 items)\n",
            toy_filtered$n_users))
stopifnot(toy_filtered$n_users == 0L, is.na(toy_filtered$FCP))

cat("  passed.\n")
```

> **Plain read.** The self-test must print row sums of 1 and a finite mean
> log-likelihood. If either is `NA`, nothing downstream is trustworthy and
> the cause is upstream in the data, not in the models.

---

## 3. How Many Factors

The number of factors is chosen on the calibration set. The test month is
not touched. The criterion is the mean log-likelihood of the Gaussian model.

```{r dim-scan, fig.height=4}
scan_file <- file.path(CACHE_DIR, "dim_scan.rds")

if (RUN_DIM_SCAN) {
  if (file.exists(scan_file)) {
    dim_scan <- readRDS(scan_file)
    cat("Loaded cached scan.\n")
  } else {
    dim_grid <- c(4L, 8L, 16L, 24L, 32L)
    dim_scan <- rbindlist(lapply(dim_grid, function(d) {
      cat(sprintf("  fitting dim = %d ...\n", d))
      r  <- train_svd(fit, d)
      s  <- pmin(pmax(score_with(r, calib), 1), 5)
      sg <- sqrt(mean((as.numeric(calib$rating5) - s)^2))
      P  <- tidy_probs(gauss_probs(s, sg))
      data.table(dim    = d,
                 sigma  = sg,
                 RMSE   = sg,
                 meanLL = mean(log(P[cbind(seq_len(nrow(P)), calib$rating5)])))
    }))
    saveRDS(dim_scan, scan_file)
  }
  print(kable(dim_scan, digits = 4,
              caption = "SVD dimension chosen on calibration data"))

  p_dim <- ggplot(dim_scan, aes(dim, meanLL)) +
    geom_line(colour = "steelblue", linewidth = 0.8) +
    geom_point(size = 2.5, colour = "steelblue") +
    geom_vline(xintercept = SVD_DIM, linetype = "dashed", colour = "grey40") +
    scale_x_continuous(breaks = dim_scan$dim) +
    theme_minimal() +
    labs(title = "Mean log-likelihood on the calibration set, by SVD dimension",
         subtitle = "Test month not used. Dashed line is the dimension carried forward.",
         x = "Number of factors", y = "Mean log-likelihood")
  print(p_dim)
  ggsave(file.path(OUT_DIR, "fig2_dim_scan.png"), p_dim,
         width = 8, height = 4.5, dpi = 150)
}
```

> **Plain read.** If the best dimension is not `r SVD_DIM`, change `SVD_DIM`
> at the top and knit again. One SGD seed per dimension, so small
> differences between neighbouring dimensions are noise, not signal.

---

## 3b. How Much L2 Regularisation

Section 3 chose the SVD dimension at a fixed, untuned `L2 = 0.01`. This
section tunes L2 at that dimension, on the calibration set, by the same
criterion (Gaussian mean log-likelihood). This is a sequential search
(dim, then L2), not a joint grid, which is a stated simplification: the
optimal dimension could in principle shift once L2 moves. Given the low
cost of retraining `recosystem`, revisiting this jointly is easy if the
chosen L2 turns out to sit far from the previous default.

```{r l2-scan, fig.height=4}
L2_GRID <- c(0.001, 0.01, 0.05, 0.1, 0.2)
l2_scan_file <- file.path(CACHE_DIR, "l2_scan.rds")

if (file.exists(l2_scan_file)) {
  l2_scan <- readRDS(l2_scan_file)
  cat("Loaded cached L2 scan.\n")
} else {
  l2_scan <- rbindlist(lapply(L2_GRID, function(l2) {
    cat(sprintf("  fitting L2 = %.3f ...\n", l2))
    r  <- train_svd(fit, SVD_DIM, l2)
    s  <- pmin(pmax(score_with(r, calib), 1), 5)
    sg <- sqrt(mean((as.numeric(calib$rating5) - s)^2))
    P  <- tidy_probs(gauss_probs(s, sg))
    data.table(L2 = l2, sigma = sg, RMSE = sg,
               meanLL = mean(log(P[cbind(seq_len(nrow(P)), calib$rating5)])))
  }))
  saveRDS(l2_scan, l2_scan_file)
}
print(kable(l2_scan, digits = 4,
            caption = "SVD L2 chosen on calibration data (dim fixed at 16)"))

BEST_L2 <- l2_scan[which.max(meanLL)]$L2
cat(sprintf("Chosen L2 = %.3f (previous default was 0.01, untuned)\n", BEST_L2))

p_l2 <- ggplot(l2_scan, aes(L2, meanLL)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(size = 2.5, colour = "steelblue") +
  geom_vline(xintercept = BEST_L2, linetype = "dashed", colour = "grey40") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Mean log-likelihood on the calibration set, by SVD L2",
       subtitle = "Test month not used. Dashed line is the L2 carried forward.",
       x = "L2 penalty (log scale)", y = "Mean log-likelihood")
print(p_l2)
ggsave(file.path(OUT_DIR, "fig2b_l2_scan.png"), p_l2,
       width = 8, height = 4.5, dpi = 150)
```

> **Plain read.** If `BEST_L2` lands at the edge of `L2_GRID` (0.001 or
> 0.2), the grid is too narrow and should be widened before trusting the
> result — same logic as the M2d hyperparameter scan.

---

```{r final-svd}
reco <- train_svd(fit, SVD_DIM, BEST_L2)
calib[, svd_score := score_with(reco, calib)]
test[,  svd_score := score_with(reco, test)]
calib[, svd_clip  := pmin(pmax(svd_score, 1), 5)]
test[,  svd_clip  := pmin(pmax(svd_score, 1), 5)]

cat(sprintf("SVD dim = %d, L2 = %.3f\n  calib scores [%.2f, %.2f]\n  test  scores [%.2f, %.2f]\n",
            SVD_DIM, BEST_L2, min(calib$svd_score), max(calib$svd_score),
            min(test$svd_score), max(test$svd_score)))

# Save the trained factors so every downstream file (M2d, M2e, M3, M4, M5)
# reuses this exact SVD instead of retraining recosystem, which is
# multi-threaded and NOT bit-reproducible across runs or R sessions. This
# is what makes 07's and 09's baseline numbers agree from now on.
fac <- reco$output(out_memory(), out_memory())
svd_cache <- list(P = fac$P, Q = fac$Q, dim = SVD_DIM, l2 = BEST_L2,
                  user_map = user_map, movie_map = movie_map, seed = SEED)
saveRDS(svd_cache, file.path(CACHE_DIR, "svd_factors.rds"))
cat("Saved SVD factors to", file.path(CACHE_DIR, "svd_factors.rds"), "\n")
```

---

## 4. M1 — SVD with a Gaussian

The SVD gives one number per user-item pair. M1 turns it into a
distribution by putting a normal density around it and integrating between
the boundaries 1.5, 2.5, 3.5 and 4.5. The standard deviation is estimated
on the calibration set. The thresholds are fixed and equally spaced.

```{r m1}
sigma_m1 <- sqrt(mean((as.numeric(calib$rating5) - calib$svd_clip)^2))
cat(sprintf("sigma estimated on calibration: %.4f\n", sigma_m1))

P_m1 <- gauss_probs(test$svd_clip, sigma_m1)
kable(metric_set(P_m1, test$rating5)$tbl, digits = 4,
      caption = "M1 on the test month")
```

---

## 5. M2a, M2b, M2c — Two-Step Cumulative Link Models

The SVD score is now the single covariate in a cumulative link model,

$$P(r_{ui} \le k) = \frac{1}{1 + \exp(-(t_k - \beta\, s_{ui}))},$$

fitted on the calibration set with the factors frozen. M2a forces equal
spacing and estimates only the spacing. M2b lets all four thresholds move.
M2c adds a random intercept for the user.

```{r clm-fit}
calib[, rating_f := factor(rating5, levels = LEVELS5, ordered = TRUE)]
calib[, user_f   := factor(userId)]

m2a <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "equidistant")
m2b <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "flexible")

cat("--- M2a ---\n"); print(coef(m2a))
cat("\n--- M2b ---\n"); print(coef(m2b))
```

```{r clmm-fit}
if (RUN_M2C) {
  t0 <- Sys.time()
  m2c <- clmm2(rating_f ~ svd_score, random = user_f, data = calib,
               link = "logistic", Hess = TRUE, nAGQ = 1)
  cat(sprintf("clmm2 fitted in %.1f minutes\n",
              as.numeric(difftime(Sys.time(), t0, units = "mins"))))
  cat(sprintf("User random intercept SD: %.4f\n", m2c$stDev))
} else {
  m2c <- NULL
  cat("M2c skipped (RUN_M2C = FALSE).\n")
}
```

```{r clm-predict}
# M2a stores equidistant params: [first_threshold, spacing]
# Extract and expand to 4 explicit thresholds
th_a <- as.numeric(m2a$alpha); b_a <- as.numeric(m2a$beta)
th_a_expanded <- th_a[1] + (0:3) * th_a[2]
th_a <- th_a_expanded

# M2b already has 4 explicit thresholds
th_b <- as.numeric(m2b$alpha); b_b <- as.numeric(m2b$beta)

P_m2a <- cum_probs(th_a, b_a * test$svd_score)
P_m2b <- cum_probs(th_b, b_b * test$svd_score)

if (RUN_M2C) {
  th_c <- as.numeric(m2c$Theta)
  b_c  <- as.numeric(m2c$beta)[1]
  re   <- m2c$ranef
  if (is.null(names(re))) names(re) <- levels(calib$user_f)
  bu <- re[as.character(test$userId)]
  n_missing_u <- sum(is.na(bu))
  bu[is.na(bu)] <- 0
  cat(sprintf("Test rows whose user has no calibration history: %d (%.1f%%)\n",
              n_missing_u, 100 * n_missing_u / nrow(test)))
  P_m2c <- cum_probs(th_c, b_c * test$svd_score + as.numeric(bu))
}
```

### Where the thresholds sit

```{r thresholds-plot, fig.height=3.6}
th_dt <- rbind(
  data.table(model = "M2a", k = 1:4, threshold = th_a),
  data.table(model = "M2b", k = 1:4, threshold = th_b))
if (RUN_M2C) th_dt <- rbind(th_dt, data.table(model = "M2c", k = 1:4, threshold = th_c))
th_dt[, cut_lbl := paste0(k, "|", k + 1)]

kable(dcast(th_dt, model ~ cut_lbl, value.var = "threshold"), digits = 3,
      caption = "Estimated thresholds on the latent scale")

p_th <- ggplot(th_dt, aes(threshold, model, colour = model)) +
  geom_line(aes(group = model), alpha = 0.3) +
  geom_point(size = 3) +
  geom_text(aes(label = cut_lbl), vjust = -1.1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c(M2a = "steelblue", M2b = "coral",
                                 M2c = "mediumseagreen")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Where the thresholds sit",
       subtitle = "M2a forces equal spacing. M2b lets them move.",
       x = "Latent scale", y = NULL)
print(p_th)
ggsave(file.path(OUT_DIR, "fig3_thresholds.png"), p_th,
       width = 8, height = 3.6, dpi = 150)
```

> **Plain read.** If the M2b thresholds are clearly unequally spaced, the
> equal-spacing assumption in M1 and M2a is wrong, and that is the first
> piece of evidence for the ordinal treatment. The gap between 4 and 5 is
> the one to look at.

### Nested likelihood-ratio tests

```{r lr-tests}
# logLik that works for both clm and clmm2 objects
get_ll <- function(m) {
  v <- try(as.numeric(logLik(m)), silent = TRUE)
  if (inherits(v, "try-error") || is.na(v)) v <- as.numeric(m$logLik)
  v
}

ll_a <- get_ll(m2a); ll_b <- get_ll(m2b)
lr_tbl <- data.table(comparison = "M2a vs M2b",
                     LR = 2 * (ll_b - ll_a), df = (K_LEVELS - 1) - 2)
lr_tbl[, p := pchisq(LR, df, lower.tail = FALSE)]

if (RUN_M2C) {
  stat_c <- 2 * (get_ll(m2c) - ll_b)
  # The variance is on the boundary under the null, so the reference
  # distribution is a 50:50 mixture of chi-square with 0 and 1 df.
  lr_tbl <- rbind(lr_tbl,
                  data.table(comparison = "M2b vs M2c", LR = stat_c, df = 1,
                             p = 0.5 * pchisq(stat_c, 1, lower.tail = FALSE)))
}
kable(lr_tbl, digits = c(0, 2, 0, 6),
      caption = "Nested likelihood-ratio tests, on the calibration set")
```

M1 is not nested in any of these, so it is compared by held-out metrics
only, not by a likelihood-ratio test.

---

## 6. Results on the Test Month

```{r results}
res_list <- list(M1  = metric_set(P_m1,  test$rating5),
                 M2a = metric_set(P_m2a, test$rating5),
                 M2b = metric_set(P_m2b, test$rating5))
if (RUN_M2C) res_list$M2c <- metric_set(P_m2c, test$rating5)

P_list <- list(M1 = P_m1, M2a = P_m2a, M2b = P_m2b)
if (RUN_M2C) P_list$M2c <- P_m2c

metrics <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))

# Naive baseline: the marginal rating distribution of the fit set
p_marg  <- as.numeric(prop.table(table(factor(fit$rating5, levels = LEVELS5))))
P_naive <- matrix(p_marg, nrow = nrow(test), ncol = K_LEVELS, byrow = TRUE)
metrics <- rbind(metrics, cbind(data.table(model = "Naive"),
                                metric_set(P_naive, test$rating5)$tbl))

kable(metrics, digits = 4, caption = "Test month, all models")

ndcg_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  data.table(model = m, NDCG10 = ndcg_at_10(E, test$rating5, test$userId))
}))
kable(ndcg_tbl, digits = 6,
      caption = "NDCG@10 (users with 5 or more test ratings)")
```

> **Plain read.** The NDCG column should be identical to six decimals across
> M1, M2a, M2b and M2c. That is not a coincidence and not a failure: within
> a user the thresholds are constant, so the ordering of items depends only
> on the latent score, which every one of these models inherits from the
> same SVD. Ranking can only change when the factors themselves are learned
> under the ordinal likelihood, which is M2d. This table is the evidence for
> that argument.

```{r metric-plot, fig.height=5}
mplot <- melt(metrics[model != "Naive"], id.vars = "model",
              measure.vars = c("meanLL", "RMSE", "RPS", "CatAcc"))
p_m <- ggplot(mplot, aes(model, value, fill = model)) +
  geom_col(width = 0.65) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_manual(values = c(M1 = "steelblue", M2a = "goldenrod",
                               M2b = "coral", M2c = "mediumseagreen")) +
  theme_minimal() + theme(legend.position = "none") +
  labs(title = "Test-month metrics", x = NULL, y = NULL)
print(p_m)
ggsave(file.path(OUT_DIR, "fig4_metrics.png"), p_m, width = 8, height = 5, dpi = 150)
```

---

## 7. Does the Model Reproduce the Rating Histogram?

```{r marginal-calibration, fig.height=4}
marg <- rbindlist(lapply(names(P_list), function(m)
  data.table(model = m, rating5 = LEVELS5,
             p = colMeans(tidy_probs(P_list[[m]])))))
marg <- rbind(marg, data.table(model = "Observed", rating5 = LEVELS5,
                               p = as.numeric(prop.table(
                                 table(factor(test$rating5, levels = LEVELS5))))))

p_cal <- ggplot(marg, aes(factor(rating5), p, fill = model)) +
  geom_col(position = "dodge", width = 0.8) +
  scale_fill_manual(values = c(Observed = "grey30", M1 = "steelblue",
                               M2a = "goldenrod", M2b = "coral",
                               M2c = "mediumseagreen")) +
  theme_minimal() +
  labs(title = "Predicted against observed rating distribution, test month",
       x = "Rating category", y = "Share", fill = NULL)
print(p_cal)
ggsave(file.path(OUT_DIR, "fig5_marginal_calibration.png"), p_cal,
       width = 8, height = 4.5, dpi = 150)
```

> **Plain read.** M1 is symmetric around its score, so it cannot reproduce a
> skewed histogram. The CLMs can, because the thresholds absorb the shape.
> Look at category 5 and category 1.

### Reliability

```{r reliability, fig.height=4.6}
rel <- rbindlist(lapply(names(P_list), function(m) {
  P  <- tidy_probs(P_list[[m]])
  ph <- rowSums(P[, 4:5])
  d  <- data.table(model = m, ph = ph, obs = as.numeric(test$rating5 >= 4))
  d[, bin := cut(ph, breaks = seq(0, 1, 0.1), include.lowest = TRUE)]
  d[, .(pred = mean(ph), obs = mean(obs), n = .N), by = .(model, bin)]
}))

p_rel <- ggplot(rel[n >= 50], aes(pred, obs, colour = model)) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_line() + geom_point(size = 2) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(values = c(M1 = "steelblue", M2a = "goldenrod",
                                 M2b = "coral", M2c = "mediumseagreen")) +
  theme_minimal() +
  labs(title = "Reliability: predicted against observed P(rating is 4 or 5)",
       subtitle = "On the dashed line means the stated probability is honest. Bins under 50 rows dropped.",
       x = "Predicted probability", y = "Observed frequency", colour = NULL)
print(p_rel)
ggsave(file.path(OUT_DIR, "fig6_reliability.png"), p_rel,
       width = 6.5, height = 5, dpi = 150)
```

---

## 8. Cluster Bootstrap Confidence Intervals

The test set has repeated measurements on the same users, so rows are not
independent. Users are resampled with replacement instead of rows.

```{r bootstrap, fig.height=3.2}
ll_mat <- do.call(cbind, lapply(res_list, function(r) r$row_ll))
colnames(ll_mat) <- names(res_list)

rows_by_user <- split(seq_len(nrow(test)), test$userId)
u_names <- names(rows_by_user)
cat(sprintf("Clusters (users) in the test set: %d\n", length(u_names)))

set.seed(SEED)
boot <- matrix(NA_real_, nrow = N_BOOT, ncol = ncol(ll_mat) - 1)
colnames(boot) <- setdiff(colnames(ll_mat), "M1")
for (b in seq_len(N_BOOT)) {
  idx <- unlist(rows_by_user[sample(u_names, length(u_names), replace = TRUE)],
                use.names = FALSE)
  mm  <- colMeans(ll_mat[idx, , drop = FALSE])
  boot[b, ] <- mm[colnames(boot)] - mm["M1"]
}

obs_diff <- colMeans(ll_mat)[colnames(boot)] - mean(ll_mat[, "M1"])
ci <- data.table(comparison  = paste(colnames(boot), "minus M1"),
                 diff_meanLL = as.numeric(obs_diff),
                 lo = apply(boot, 2, quantile, 0.025),
                 hi = apply(boot, 2, quantile, 0.975))
ci[, significant := (lo > 0) | (hi < 0)]
kable(ci, digits = 5,
      caption = sprintf("Cluster bootstrap, %d replicates, users resampled", N_BOOT))

p_ci <- ggplot(ci, aes(comparison, diff_meanLL)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "steelblue") +
  coord_flip() + theme_minimal() +
  labs(title = "Gain in mean log-likelihood over M1",
       subtitle = "95% cluster bootstrap intervals. Above zero means better than the Gaussian baseline.",
       x = NULL, y = "Difference in mean log-likelihood")
print(p_ci)
ggsave(file.path(OUT_DIR, "fig7_bootstrap_ci.png"), p_ci,
       width = 7, height = 3.2, dpi = 150)
```

---

## 9. Three Example Predictions

```{r examples, fig.height=4}
set.seed(SEED)
ex_rows <- c(which.max(res_list$M2b$row_ll),
             which.min(res_list$M2b$row_ll),
             sample(nrow(test), 1))
ex <- rbindlist(lapply(ex_rows, function(i)
  rbindlist(lapply(names(P_list), function(m)
    data.table(case  = sprintf("row %d, true rating %d", i, test$rating5[i]),
               model = m, rating5 = LEVELS5,
               p = tidy_probs(P_list[[m]])[i, ])))))

p_ex <- ggplot(ex, aes(factor(rating5), p, fill = model)) +
  geom_col(position = "dodge") +
  facet_wrap(~ case) +
  scale_fill_manual(values = c(M1 = "steelblue", M2a = "goldenrod",
                               M2b = "coral", M2c = "mediumseagreen")) +
  theme_minimal() +
  labs(title = "Predicted distributions for three test rows",
       x = "Rating category", y = "Probability", fill = NULL)
print(p_ex)
ggsave(file.path(OUT_DIR, "fig8_examples.png"), p_ex,
       width = 9, height = 4, dpi = 150)
```

---

## 10. Numbers for the Meeting

```{r summary}
fwrite(metrics,  file.path(OUT_DIR, "test_metrics.csv"))

ndcg_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  data.table(model = m, NDCG10 = ndcg_at_10(E, test$rating5, test$userId))
}))
# M1's mean must live on the rating scale, so it uses the clipped score.
# Clipping is flat outside [1,5] and creates ties, which is the only reason
# M1 differs. Ranked on the raw score, M1 matches the CLMs exactly.
ndcg_tbl <- rbind(ndcg_tbl,
  data.table(model  = "M1 (ranked on raw score)",
             NDCG10 = ndcg_at_10(test$svd_score, test$rating5, test$userId)))
kable(ndcg_tbl, digits = 6)

fwrite(ci,       file.path(OUT_DIR, "bootstrap_ci.csv"))
fwrite(th_dt,    file.path(OUT_DIR, "thresholds.csv"))
fwrite(lr_tbl,   file.path(OUT_DIR, "lr_tests.csv"))

saveRDS(list(metrics = metrics, ndcg = ndcg_tbl, ci = ci, thresholds = th_dt,
             lr = lr_tbl, sigma_m1 = sigma_m1, svd_dim = SVD_DIM,
             n_fit = nrow(fit), n_calib = nrow(calib), n_test = nrow(test)),
        file.path(CACHE_DIR, "m1_m2_results.rds"))

cat("Files written to", OUT_DIR, ":\n")
cat(paste(" ", list.files(OUT_DIR)), sep = "\n")

best <- metrics[model != "Naive"][which.max(meanLL)]
cat(sprintf("\nBest model on test mean log-likelihood: %s (%.4f), against M1 (%.4f).\n",
            best$model, best$meanLL, metrics[model == "M1", meanLL]))
cat(sprintf("NDCG@10: M2a = M2b = M2c = %.6f, identical to 15 digits.\n",
            ndcg_tbl[model == "M2b", NDCG10]))

fcp_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  r <- fcp_stat(E, test$rating5, test$userId)
  data.table(model = m, FCP = r$FCP, n_pairs = r$n_pairs, n_users = r$n_users)
}))
fcp_tbl <- rbind(fcp_tbl, {
  r <- fcp_stat(test$svd_score, test$rating5, test$userId)
  data.table(model = "M1 (ranked on raw score)", FCP = r$FCP,
             n_pairs = r$n_pairs, n_users = r$n_users)
})
kable(fcp_tbl, digits = 6,
      caption = "Fraction of Concordant Pairs (Koren & Sill 2011, eq. 26)")

fwrite(fcp_tbl, file.path(OUT_DIR, "fcp.csv"))

cat("M1 differs only because its Gaussian mean uses the clipped score; ranked on the raw score it matches exactly.\n")

```

Plain read. FCP should be identical across M1, M2a, M2b and M2c, for
the same reason NDCG was: within a user the thresholds are constant, so
the expected-rating ranking depends only on the latent score, which all
four models inherit unchanged from the same SVD. This is a second,
independent piece of evidence that ranking cannot move before M2d.
n_pairs is the number of comparable item pairs (true ratings differ)
summed over all test users; it will be smaller than the full C(n,2) count
because tied true ratings are excluded.

---

## 11. What Is Not Here

1. **M2d and M2e.** The joint OrdRec models. They learn the factors and the
   thresholds together by stochastic gradient ascent, and they are the only
   step where the ranking can change. Written next, with a numerical
   gradient check before any result is believed.
2. **Cold-start users.** 84% of the test month is users with no ratings in
   the training window. Out of scope by design.
3. **L2 tuning.** Done in section 3b: `r sprintf("%.3f", BEST_L2)` chosen
   on the calibration set, at dim = 16. The dimension itself was chosen at
   the old default (0.01) — a sequential rather than joint search, flagged
   as a simplification there.

### Questions

1. The number of factors is chosen on same-window calibration data, but the
   evaluation is the next month. Acceptable, or should the selection use an
   earlier month pair?
2. M2c uses the conditional modes of the random intercepts for test users,
   and zero for users with no calibration history. Is that what you expect?
3. For the joint models, do you want the grid search over learning rate,
   regularisation and iterations scripted, or is a manual search enough?

---

## References

Agresti, A. (2010) *Analysis of Ordinal Categorical Data*. 2nd ed.
Hoboken, NJ: Wiley.

Koren, Y. and Sill, J. (2011) OrdRec: an ordinal model for predicting
personalized item rating distributions. *Proceedings of the 5th ACM
Conference on Recommender Systems*, pp. 117–124.

McCullagh, P. (1980) Regression models for ordinal data. *Journal of the
Royal Statistical Society, Series B*, 42(2), pp. 109–142.
