---
title: "M2d: Joint OrdRec with Global Thresholds"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(ordinal)
library(recosystem)
library(knitr)
```

## Aim

M2d is the first model in the ladder where the factors are **learned under
the ordinal likelihood** instead of being frozen at the squared-error SVD.
The latent score is plain matrix factorisation with biases,

$$y_{ui} = b_i + b_u + q_i^\top p_u,$$

wrapped in the OrdRec cumulative-logit shell of Koren and Sill (2011):
$t_1$ free, $t_{r+1} = t_r + \exp(\beta_r)$ (their eqs. 4–5, monotone by
construction), $P(r_{ui} \le r) = \sigma(t_r - y_{ui})$ (their eq. 9), and
stochastic gradient ascent on the log-likelihood (their eqs. 11–12).

Our stated deviation from the paper: they wrap SVD++ (their eq. 3, with the
implicit-feedback term $|R(u)|^{-1/2}\sum_j x_j$); we wrap plain MF. The
ordinal shell is unchanged, and the paper itself notes any internal scorer
can be used. A second, smaller deviation is recorded in Section 3: gradient
updates are applied in mini-batches rather than one row at a time.

**Why M2d matters for this dissertation.** NDCG\@10 and FCP are identical
across M1–M2c because those models share one frozen latent score. Joint
training is the only step on the ladder where the ranking can move. This
file answers whether it does.

**Trust chain.** Every likelihood and gradient formula used below is
byte-identical to the functions in `08_M2d_gradient_check.R`, which passed
(a) a numerical gradient check to max relative error 5e-09, and (b) a
frozen-factor cross-check against `ordinal::clm` agreeing to 2e-08 in the
thresholds and to machine precision in the log-likelihood. The mini-batch
trainer additionally passes the synthetic-recovery self-test in Section 3.

---

## 0. Settings

```{r config}
# --- edit these three paths (same as 07) ------------------------------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/00_data_selection/selection_v7_2018-04_6m_Nu10_Nm10.rds"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/07_M1_M2"
CACHE_DIR <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
# ----------------------------------------------------------------------

dir.create(OUT_DIR,   showWarnings = FALSE, recursive = TRUE)
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

# ---- scan slicing ----------------------------------------------------
# Set BEFORE sourcing config_grids.R. Split the grid across several R
# sessions to parallelise by hand: session j of m uses c(j, m).
# Single session: c(1L, 1L).
SCAN_SLICE <- c(1L, 1L)

# Shared grid, protocol and constants for M2d and M2e. Defined in one
# place so the two models cannot drift apart: the supervisor asked for
# the M2e tuning to be done "for consistency", and consistency here means
# an identical search space under an identical protocol.
source("config_grids.R")

RUN_M2C <- TRUE      # clmm2 takes ~20 min; the fit is cached after the first run

n_cores <- parallel::detectCores()
set.seed(SEED)
```

---

## 1. Load the Selection (identical to 07)

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
cat(sprintf("fit %d | calib %d | test %d | users %d | movies %d\n",
            nrow(fit), nrow(calib), nrow(test), nU, nI))
stopifnot(nrow(fit) == 393364L)   # guard against loading a stale selection
```

---

## 2. Core Functions

The model core (`expand_thresholds`, `sigmoid`, `ordrec_forward`) is copied
verbatim from `08_M2d_gradient_check.R`. Metric functions are copied from
07 so the evaluation protocol is bit-identical across the ladder.

```{r core-functions}
# ---- OrdRec core (verbatim from 08, where it passed both checks) -----
expand_thresholds <- function(t1, beta) c(t1, t1 + cumsum(exp(beta)))
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))

ordrec_forward <- function(y, r, th) {
  g_hi <- ifelse(r == K_LEVELS, 1, sigmoid(th[pmin(r, K_LEVELS - 1L)] - y))
  g_lo <- ifelse(r == 1L,       0, sigmoid(th[pmax(r - 1L, 1L)]       - y))
  A <- g_hi * (1 - g_hi)
  B <- g_lo * (1 - g_lo)
  list(prob = pmax(g_hi - g_lo, 1e-12), A = A, B = B)
}

calc_scores <- function(par, u, i)
  par$bi[i] + par$bu[u] + rowSums(par$P[u, , drop = FALSE] *
                                  par$Q[i, , drop = FALSE])

ordrec_mean_ll <- function(par, dat) {
  y  <- calc_scores(par, dat$u, dat$i)
  th <- expand_thresholds(par$t1, par$beta)
  mean(log(ordrec_forward(y, dat$r, th)$prob))
}

# Full 5-column probability matrix for evaluation. Same cumulative-logit
# formula as cum_probs in 07; written against the par list for convenience.
ordrec_probs <- function(par, u, i) {
  y  <- calc_scores(par, u, i)
  th <- expand_thresholds(par$t1, par$beta)
  CP <- sapply(th, function(t) plogis(t - y))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- metrics, verbatim from 07 ---------------------------------------
tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

gauss_probs <- function(y, sigma) {
  y  <- as.numeric(y)
  CP <- sapply(CUTS, function(c) pnorm((c - y) / sigma))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

cum_probs <- function(theta, eta) {
  eta <- as.numeric(eta)
  CP  <- sapply(theta, function(t) plogis(t - eta))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(eta))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

metric_set <- function(P, y) {
  y   <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  per_user <- d[, {
    k    <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]
  mean(per_user$ndcg, na.rm = TRUE)
}

fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0)
    return(list(FCP = NA_real_, nc = 0L, nd = 0L, n_pairs = 0L, n_users = 0L))
  per_user <- d[, {
    sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
    ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  nc <- sum(per_user$nc); nd <- sum(per_user$nd)
  list(FCP = nc / (nc + nd), nc = nc, nd = nd,
       n_pairs = nc + nd, n_users = uniqueN(d$uid))
}
```

---

## 3. The Mini-Batch SGD Trainer

Koren and Sill update after every single rating (their eq. 12). Here
gradients are accumulated over shuffled mini-batches of `r BATCH_SIZE` rows
and applied once per batch — identical in expectation, and the practical
choice in R, where per-row loops are two orders of magnitude slower than
vectorised batches.

Two details, both discovered by testing rather than assumed:

1. **Thresholds get the batch-mean gradient, entity parameters the batch
   sum.** A user or item appears in a batch a handful of times, so the sum
   over the batch is the same movement per-row SGD would have produced.
   The thresholds are touched by *every* row; summing their gradients over
   a batch multiplies the step by the batch size and the training diverges
   (verified). Dividing by the batch size restores the scale. This is a
   per-parameter-type learning rate, which is exactly what Koren and Sill
   describe using in their Section 6.1.
2. **Per-row gradient terms are capped at `r CLIP`.** The terms
   $A/p$ and $B/p$ blow up when the model is confidently wrong ($p \to 0$).
   The cap is a safeguard; the fraction of rows where it binds is reported
   for every run and should be ~0 for the chosen configuration.

Regularisation is L2 on biases and factors, applied per occurrence as in
the paper's eq. 12; the thresholds are left unpenalised (four global
parameters against 393k rows do not overfit, and shrinking thresholds
towards zero has no meaningful prior interpretation).

```{r trainer}
train_ordrec_m2d <- function(dat_fit, dat_calib, init, eta, lambda,
                             batch_size = BATCH_SIZE, max_epochs = MAX_EPOCHS,
                             patience = PATIENCE, seed = SEED, clip = CLIP,
                             eta_decay = ETA_DECAY, verbose = TRUE) {
  par <- init
  n <- length(dat_fit$u)
  best <- list(ll = -Inf, par = par, epoch = 0L)
  hist <- data.table(epoch = integer(), fit_meanLL = numeric(),
                     calib_meanLL = numeric(), clip_rate = numeric())
  set.seed(seed)
  eta_e <- eta
  for (ep in seq_len(max_epochs)) {
    perm <- sample.int(n)
    n_clip <- 0L
    for (b0 in seq(1L, n, by = batch_size)) {
      idx <- perm[b0:min(b0 + batch_size - 1L, n)]
      u <- dat_fit$u[idx]; i <- dat_fit$i[idx]; r <- dat_fit$r[idx]

      y  <- calc_scores(par, u, i)
      th <- expand_thresholds(par$t1, par$beta)
      fw <- ordrec_forward(y, r, th)
      c_hi <- fw$A / fw$prob; c_lo <- fw$B / fw$prob
      n_clip <- n_clip + sum(c_hi > clip) + sum(c_lo > clip)
      c_hi <- pmin(c_hi, clip); c_lo <- pmin(c_lo, clip)
      dy <- c_lo - c_hi                                   # dl/dy per row

      # thresholds: batch-MEAN gradient (see note above)
      d_t1 <- sum(c_hi - c_lo)
      d_be <- vapply(1:(K_LEVELS - 2L), function(s)
        exp(par$beta[s]) * sum(c_hi * (s <= r - 1L) - c_lo * (s <= r - 2L)),
        numeric(1))
      par$t1   <- par$t1   + eta_e * d_t1 / length(idx)
      par$beta <- par$beta + eta_e * d_be / length(idx)

      # users: batch-SUM gradient, per-occurrence L2 (paper eq. 12)
      su <- rowsum(dy, u); ug <- as.integer(rownames(su))
      nu <- as.numeric(rowsum(rep(1, length(u)), u))
      par$bu[ug] <- par$bu[ug] + eta_e * (as.numeric(su) - lambda * nu * par$bu[ug])
      SP <- rowsum(dy * par$Q[i, , drop = FALSE], u)
      par$P[ug, ] <- par$P[ug, ] + eta_e * (SP - lambda * nu * par$P[ug, , drop = FALSE])

      # items
      si <- rowsum(dy, i); ig <- as.integer(rownames(si))
      ni <- as.numeric(rowsum(rep(1, length(i)), i))
      par$bi[ig] <- par$bi[ig] + eta_e * (as.numeric(si) - lambda * ni * par$bi[ig])
      SQ <- rowsum(dy * par$P[u, , drop = FALSE], i)
      par$Q[ig, ] <- par$Q[ig, ] + eta_e * (SQ - lambda * ni * par$Q[ig, , drop = FALSE])

      if (!is.finite(par$t1) || any(!is.finite(par$beta)))
        return(list(diverged = TRUE, best_ll = -Inf, best_epoch = 0L,
                    history = hist))
    }
    ll_c <- ordrec_mean_ll(par, dat_calib)
    ll_f <- ordrec_mean_ll(par, dat_fit)
    hist <- rbind(hist, data.table(epoch = ep, fit_meanLL = ll_f,
                                   calib_meanLL = ll_c, clip_rate = n_clip / n))
    if (verbose) cat(sprintf("    epoch %2d  fit %.5f  calib %.5f  clip %.5f%%\n",
                             ep, ll_f, ll_c, 100 * n_clip / n))
    if (ll_c > best$ll + 1e-5) {
      best <- list(ll = ll_c, par = par, epoch = ep)
    } else if (ep - best$epoch >= patience) break
    eta_e <- eta_e * eta_decay
  }
  list(diverged = FALSE, best_ll = best$ll, par = best$par,
       best_epoch = best$epoch, history = hist)
}
```

### Synthetic self-test

Before touching real data: generate ratings from known parameters, train
from a cold start, and require the trainer to (a) clearly beat the marginal
baseline and (b) land near the oracle likelihood. Runs in a few seconds; if
it ever fails after an edit to the trainer, believe the test, not the edit.

```{r trainer-selftest}
set.seed(7)
.nU <- 400L; .nI <- 200L; .K <- 8L
.true <- list(t1 = 0.5, beta = log(c(1.2, 1.6, 2.0)),
              bu = rnorm(.nU, 0, 0.5), bi = rnorm(.nI, 0, 0.5),
              P = matrix(rnorm(.nU * .K, 0, 0.55), .nU, .K),
              Q = matrix(rnorm(.nI * .K, 0, 0.55), .nI, .K))
.gen <- function(n) {
  u <- sample(.nU, n, TRUE); i <- sample(.nI, n, TRUE)
  y  <- calc_scores(.true, u, i)
  th <- expand_thresholds(.true$t1, .true$beta)
  G  <- sapply(th, function(t) sigmoid(t - y)); G <- cbind(G, 1)
  P5 <- cbind(G[, 1], G[, 2:5] - G[, 1:4])
  list(u = u, i = i,
       r = as.integer(apply(P5, 1, function(p) sample.int(5, 1, prob = p))))
}
.df <- .gen(60000L); .dc <- .gen(15000L)
.oracle <- ordrec_mean_ll(.true, .dc)
.marg   <- prop.table(tabulate(.df$r, 5))
.mll    <- mean(log(.marg[.dc$r]))
.cum    <- qlogis(cumsum(.marg)[1:4])
.cold   <- list(t1 = .cum[1], beta = log(pmax(diff(.cum), 1e-3)),
                bu = rep(0, .nU), bi = rep(0, .nI),
                P = matrix(rnorm(.nU * .K, 0, 0.05), .nU, .K),
                Q = matrix(rnorm(.nI * .K, 0, 0.05), .nI, .K))
.res <- train_ordrec_m2d(.df, .dc, .cold, eta = 0.03, lambda = 0.01,
                         verbose = FALSE)
cat(sprintf("synthetic: trained %.5f | oracle %.5f | marginal %.5f\n",
            .res$best_ll, .oracle, .mll))
stopifnot(!.res$diverged, .res$best_ll > .mll + 0.05,
          .res$best_ll > .oracle - 0.06)
cat("TRAINER SELF-TEST: PASS\n")
rm(.df, .dc, .true, .cold, .res)
```

---

## 4. Baselines Recomputed (M1, M2a, M2b, M2c)

Self-contained rebuild of the comparison models so the bootstrap can use
per-row log-likelihoods from a single session. The SVD here also supplies
M2d's warm start. `recosystem` is multi-threaded and not bit-reproducible,
so these numbers may differ from 07 in the third or fourth decimal; every
comparison in this file is internally consistent because all models see
this session's scores. The 20-minute `clmm2` fit for M2c is cached.

```{r baselines}

# 读 07 缓存的调优 SVD 因子(dim=16, L2=0.05)。09 不再自己训 recosystem
nU <- nrow(user_map); nI <- nrow(movie_map)  # 来自第1节,这里确保一下
svd_cache <- readRDS(file.path(CACHE_DIR, "svd_factors.rds"))
stopifnot(identical(as.integer(svd_cache$user_map$userId),   as.integer(user_map$userId)),
          identical(as.integer(svd_cache$movie_map$movieId), as.integer(movie_map$movieId)),
          svd_cache$dim == SVD_DIM,
          nrow(svd_cache$P) == nU, nrow(svd_cache$Q) == nI)

# 读 07 缓存的调优 SVD 因子(dim=16, L2=0.05)。09 不再自己训 recosystem:
# 它多线程、非 bit-可复现,正是旧 09 基线和 07 在第 3–4 位小数漂移的根源。
# 读缓存后,M1/M2a/M2b/M2c 与 M2d 的 warm start 共享同一组固定因子,与 07 对齐。
svd_cache <- readRDS(file.path(CACHE_DIR, "svd_factors.rds"))
stopifnot(identical(as.integer(svd_cache$user_map$userId),   as.integer(user_map$userId)),
          identical(as.integer(svd_cache$movie_map$movieId), as.integer(movie_map$movieId)),
          svd_cache$dim == SVD_DIM,
          nrow(svd_cache$P) == nU, nrow(svd_cache$Q) == nI)
Pmat <- svd_cache$P    # nU x dim, 第 k 行 = user_map$userId[k]
Qmat <- svd_cache$Q    # nI x dim, 第 k 行 = movie_map$movieId[k]
cat(sprintf("Loaded SVD factors from cache (dim=%d, L2=%g)\n",
            svd_cache$dim, svd_cache$l2))

svd_score_vec <- function(u, i)
  rowSums(Pmat[u, , drop = FALSE] * Qmat[i, , drop = FALSE])
calib[, svd_score := svd_score_vec(user_idx, movie_idx)]
test[,  svd_score := svd_score_vec(user_idx, movie_idx)]
calib[, svd_clip := pmin(pmax(svd_score, 1), 5)]
test[,  svd_clip := pmin(pmax(svd_score, 1), 5)]

# M1
sigma_m1 <- sqrt(mean((as.numeric(calib$rating5) - calib$svd_clip)^2))
P_m1 <- gauss_probs(test$svd_clip, sigma_m1)

# M2a / M2b
calib[, rating_f := factor(rating5, levels = LEVELS5, ordered = TRUE)]
m2a <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "equidistant")
m2b <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "flexible")
th_a <- as.numeric(m2a$alpha); b_a <- as.numeric(m2a$beta)
th_a <- th_a[1] + (0:3) * th_a[2]
th_b <- as.numeric(m2b$alpha); b_b <- as.numeric(m2b$beta)
P_m2a <- cum_probs(th_a, b_a * test$svd_score)
P_m2b <- cum_probs(th_b, b_b * test$svd_score)

# M2c (cached: clmm2 is the slow one)
m2c_cache <- file.path(CACHE_DIR, "m2c_params.rds")
if (RUN_M2C) {
  if (file.exists(m2c_cache)) {
    m2c_par <- readRDS(m2c_cache)
    cat("Loaded cached M2c parameters.\n")
  } else {
    calib[, user_f := factor(userId)]
    t0 <- Sys.time()
    m2c <- clmm2(rating_f ~ svd_score, random = user_f, data = calib,
                 link = "logistic", Hess = TRUE, nAGQ = 1)
    cat(sprintf("clmm2 fitted in %.1f minutes\n",
                as.numeric(difftime(Sys.time(), t0, units = "mins"))))
    re <- m2c$ranef
    if (is.null(names(re))) names(re) <- levels(calib$user_f)
    m2c_par <- list(Theta = as.numeric(m2c$Theta),
                    beta  = as.numeric(m2c$beta)[1],
                    ranef = re, stDev = as.numeric(m2c$stDev))
    saveRDS(m2c_par, m2c_cache)
  }
  bu_c <- m2c_par$ranef[as.character(test$userId)]
  bu_c[is.na(bu_c)] <- 0
  P_m2c <- cum_probs(m2c_par$Theta,
                     m2c_par$beta * test$svd_score + as.numeric(bu_c))
}
```

---

## 5. Initialisation for M2d

**Warm start** takes the SVD's own factor matrices (so M2d starts exactly
where M1's internal scorer stands) with zero biases, and sets the four
thresholds by maximising the ordinal likelihood on the fit set with the
factors frozen — the cheap four-parameter BFGS that was cross-checked
against `clm` in 08. **Cold start** uses small random factors and
thresholds matched to the marginal rating distribution of the fit set
($t_k = \mathrm{logit}$ of the empirical $P(r \le k)$, so the model starts
as the marginal predictor). Running both answers the reviewer question of
whether the joint solution merely orbits the SVD it was seeded with.

```{r inits}
# warm-start 因子就是第 4 节读入的缓存 SVD 因子
fac <- list(P = Pmat, Q = Qmat)
stopifnot(nrow(fac$P) == nU, nrow(fac$Q) == nI)

dat_fit   <- list(u = fit$user_idx,   i = fit$movie_idx,   r = fit$rating5)
dat_calib <- list(u = calib$user_idx, i = calib$movie_idx, r = calib$rating5)
dat_test  <- list(u = test$user_idx,  i = test$movie_idx,  r = test$rating5)

# thresholds for the warm start: frozen-factor BFGS on the fit set
ordrec_frozen_llgr <- function(v, y, r) {
  th <- expand_thresholds(v[1], v[2:4])
  fw <- ordrec_forward(y, r, th)
  d_t1 <- sum((fw$A - fw$B) / fw$prob)
  d_be <- vapply(1:3, function(s)
    exp(v[1 + s]) * sum((fw$A * (s <= r - 1L) - fw$B * (s <= r - 2L)) / fw$prob),
    numeric(1))
  list(ll = sum(log(fw$prob)), gr = c(d_t1, d_be))
}
y_warm_fit <- rowSums(fac$P[dat_fit$u, ] * fac$Q[dat_fit$i, ])
opt <- optim(c(0, 0, 0, 0),
             fn = function(v) -ordrec_frozen_llgr(v, y_warm_fit, dat_fit$r)$ll,
             gr = function(v) -ordrec_frozen_llgr(v, y_warm_fit, dat_fit$r)$gr,
             method = "BFGS", control = list(maxit = 500, reltol = 1e-12))
cat("warm-start thresholds:",
    round(expand_thresholds(opt$par[1], opt$par[2:4]), 4), "\n")

init_warm <- list(t1 = opt$par[1], beta = opt$par[2:4],
                  bu = rep(0, nU), bi = rep(0, nI),
                  P = fac$P, Q = fac$Q)

marg5 <- prop.table(tabulate(dat_fit$r, K_LEVELS))
t_cold <- qlogis(cumsum(marg5)[1:4])
set.seed(SEED)
init_cold <- list(t1 = t_cold[1], beta = log(pmax(diff(t_cold), 1e-3)),
                  bu = rep(0, nU), bi = rep(0, nI),
                  P = matrix(rnorm(nU * SVD_DIM, 0, 0.05), nU, SVD_DIM),
                  Q = matrix(rnorm(nI * SVD_DIM, 0, 0.05), nI, SVD_DIM))
```

---

## 6. Hyperparameter Scan on the Calibration Set

One unified learning rate and one unified L2 for biases and factors
(supervisor-agreed simplification of the paper's per-type grid; the
threshold step is the one exception, explained in Section 3). Each
configuration trains with early stopping on the calibration mean
log-likelihood and is cached. The test month is not touched.

```{r scan}
## 6a. Compute pass (slice-aware; trains only this session's share)
grid <- CJ(eta = ETA_GRID, lambda = LAMBDA_GRID, init = INIT_GRID)

# Failure floor: the marginal predictor's calibration mean log-likelihood.
# Any configuration that cannot beat it, or that peaks at epoch <= 1
# (never settled), is not a candidate. marg5 is the fit-set rating
# distribution from Section 5.
marg_ll_calib <- mean(log(marg5[dat_calib$r]))

scan_file <- function(g)
  file.path(CACHE_DIR, sprintf("m2d_scan_eta%g_lam%g_%s.rds",
                               g$eta, g$lambda, g$init))

rows_mine <- seq(SCAN_SLICE[1], nrow(grid), by = SCAN_SLICE[2])
cat(sprintf("Grid: %d configurations | this session runs %d (slice %d/%d)\n",
            nrow(grid), length(rows_mine), SCAN_SLICE[1], SCAN_SLICE[2]))
cat(sprintf("Marginal-predictor calib meanLL = %.5f (failure floor)\n",
            marg_ll_calib))

for (k in rows_mine) {
  g  <- grid[k]
  cf <- scan_file(g)
  if (file.exists(cf)) next
  cat(sprintf("  [%d/%d] eta=%g lambda=%g init=%s ... ",
              k, nrow(grid), g$eta, g$lambda, g$init))
  t0 <- Sys.time()
  r  <- train_ordrec_m2d(dat_fit, dat_calib,
                         if (g$init == "warm") init_warm else init_cold,
                         eta = g$eta, lambda = g$lambda, verbose = FALSE)
  saveRDS(list(best_ll = r$best_ll, best_epoch = r$best_epoch,
               diverged = r$diverged, history = r$history), cf)
  cat(sprintf("%.1f min (best epoch %d)\n",
              as.numeric(difftime(Sys.time(), t0, units = "mins")),
              r$best_epoch))
}

## 6b. Assembly (requires every cache file, including other sessions')
missing <- grid[!vapply(seq_len(nrow(grid)),
                        function(k) file.exists(scan_file(grid[k])), logical(1))]
if (nrow(missing)) {
  print(missing)
  stop(sprintf("%d configurations missing. Run the remaining slices before assembling.",
               nrow(missing)))
}

scan <- rbindlist(lapply(seq_len(nrow(grid)), function(k) {
  g <- grid[k]; res <- readRDS(scan_file(g))
  data.table(g,
    calib_meanLL = res$best_ll,
    best_epoch   = res$best_epoch,
    diverged     = res$diverged,
    # Budget-limited, not converged: early stopping never fired, so the
    # calibration curve was still rising when MAX_EPOCHS cut it off. Under
    # a geometric decay the total step budget is proportional to eta, so
    # this is expected at the small learning rates and such rows are NOT
    # evidence that the learning rate itself is poor.
    truncated    = !res$diverged && res$best_epoch >= MAX_EPOCHS,
    failed       = res$diverged || res$best_epoch <= 1L ||
                   res$best_ll < marg_ll_calib,
    clip_rate    = if (nrow(res$history)) max(res$history$clip_rate) else NA)
}))

kable(scan[order(-calib_meanLL)], digits = 5,
      caption = "M2d hyperparameter scan, ranked by calibration mean log-likelihood")

best_cfg <- scan[failed == FALSE][which.max(calib_meanLL)]
cat(sprintf("\nChosen: eta=%g lambda=%g init=%s (calib meanLL %.5f, epoch %d)\n",
            best_cfg$eta, best_cfg$lambda, best_cfg$init,
            best_cfg$calib_meanLL, best_cfg$best_epoch))

## 6c. Boundary and truncation checks
on_edge <- function(v, grid_v)
  if (v == min(grid_v)) "LOWER BOUNDARY" else
  if (v == max(grid_v)) "UPPER BOUNDARY" else "interior"

cat(sprintf("  eta    = %g  -> %s of %s\n", best_cfg$eta,
            on_edge(best_cfg$eta, ETA_GRID),
            paste(ETA_GRID, collapse = ", ")))
cat(sprintf("  lambda = %g  -> %s of %s\n", best_cfg$lambda,
            on_edge(best_cfg$lambda, LAMBDA_GRID),
            paste(LAMBDA_GRID, collapse = ", ")))
if (on_edge(best_cfg$eta, ETA_GRID) != "interior" ||
    on_edge(best_cfg$lambda, LAMBDA_GRID) != "interior")
  warning("Selected optimum sits on a grid boundary: the grid must be ",
          "extended in that direction before the value can be reported ",
          "as an optimum rather than as the best of what was searched.")

n_trunc <- sum(scan$truncated)
cat(sprintf("\nBudget-limited (best epoch == MAX_EPOCHS = %d): %d of %d configs\n",
            MAX_EPOCHS, n_trunc, nrow(scan)))
if (n_trunc) print(scan[truncated == TRUE][order(eta, lambda)])
```


> **Plain read.** Three things to check, in order. **(1)** Is the selected
> `eta` interior? Until it is, the reported value is the best of what was
> searched, not an optimum, and the boundary warning says so. **(2)** Are
> the smallest learning rates flagged `truncated`? Under a geometric decay
> the total step budget is proportional to `eta`, so a run at 0.003 has one
> tenth the budget of one at 0.03; if it loses while still improving, it
> lost on budget rather than on merit, and the write-up says that rather
> than claiming small rates were tested and rejected. **(3)** Compare the
> best warm row against the best cold row: if they land at similar
> calibration likelihoods, the joint optimum is not an artefact of the SVD
> seeding. Diverged rows should all sit at the largest learning rate —
> that is the grid doing its job, not a failure.

---

## 7. Final M2d Training

```{r final-train, fig.height=4}
final_cache <- file.path(CACHE_DIR, "m2d_final.rds")
if (file.exists(final_cache)) {
  m2d <- readRDS(final_cache)
  cat("Loaded cached final M2d.\n")
} else {
  m2d <- train_ordrec_m2d(dat_fit, dat_calib,
                          if (best_cfg$init == "warm") init_warm else init_cold,
                          eta = best_cfg$eta, lambda = best_cfg$lambda,
                          verbose = TRUE)
  saveRDS(m2d, final_cache)
}
th_d <- expand_thresholds(m2d$par$t1, m2d$par$beta)
cat(sprintf("best epoch %d | calib meanLL %.5f | max clip rate %.5f%%\n",
            m2d$best_epoch, m2d$best_ll, 100 * max(m2d$history$clip_rate)))
cat("final thresholds:", round(th_d, 4), "\n")

hc <- melt(m2d$history[, .(epoch, fit_meanLL, calib_meanLL)], id.vars = "epoch")
p_curve <- ggplot(hc, aes(epoch, value, colour = variable)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.5) +
  geom_vline(xintercept = m2d$best_epoch, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(fit_meanLL = "steelblue",
                                 calib_meanLL = "coral"),
                      labels = c("fit", "calibration")) +
  theme_minimal() +
  labs(title = "M2d training curves",
       subtitle = "Dashed line: epoch with the best calibration likelihood (parameters kept).",
       x = "Epoch", y = "Mean log-likelihood", colour = NULL)
print(p_curve)
ggsave(file.path(OUT_DIR, "fig9_m2d_training_curve.png"), p_curve,
       width = 8, height = 4, dpi = 150)
```

---

## 8. Test-Month Evaluation

Same protocol as 07, one new row.

```{r test-eval}
P_m2d <- ordrec_probs(m2d$par, dat_test$u, dat_test$i)

P_list <- list(M1 = P_m1, M2a = P_m2a, M2b = P_m2b)
if (RUN_M2C) P_list$M2c <- P_m2c
P_list$M2d <- P_m2d

res_list <- lapply(P_list, metric_set, y = test$rating5)
metrics  <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))
kable(metrics, digits = 4, caption = "Test month, all models")

rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  f <- fcp_stat(E, test$rating5, test$userId)
  data.table(model = m,
             NDCG10 = ndcg_at_10(E, test$rating5, test$userId),
             FCP = f$FCP, n_pairs = f$n_pairs, n_users = f$n_users)
}))
kable(rank_tbl, digits = 6, caption = "Ranking metrics (users with >= 5 test ratings)")
```

> **Plain read.** The single most important comparison in this file is the
> M2d row of the ranking table against the frozen M2a–M2c rows. M1–M2c are
> identical by algebra; M2d is the first model whose latent scores were
> allowed to move. Whether NDCG and FCP improve, worsen, or barely move is
> a result either way and goes in the dissertation as such. The likelihood
> comparison (meanLL against M2c's -1.2080 territory) answers whether joint
> training buys anything beyond the two-step shortcut.

### Thresholds across the ladder

```{r thresholds}
th_tbl <- data.table(model = c("M2a", "M2b", "M2c"[RUN_M2C], "M2d"),
                     rbind(th_a, th_b,
                           if (RUN_M2C) m2c_par$Theta, th_d))
setnames(th_tbl, c("model", paste0(1:4, "|", 2:5)))
th_long <- melt(th_tbl, id.vars = "model")
th_wide <- dcast(th_long, model ~ variable)
gaps <- th_wide[, .(model,
                    g12 = `2|3` - `1|2`, g23 = `3|4` - `2|3`,
                    g34 = `4|5` - `3|4`)]
kable(th_wide, digits = 3, caption = "Thresholds on the latent scale")
kable(gaps, digits = 3, caption = "Threshold gaps (unequal spacing carries over?)")
```

> **Plain read.** M2d's latent scale is its own (its scores are trained,
> not inherited), so absolute threshold values are not comparable across
> rows — the *pattern of gaps* is. If the 4|5 gap remains the widest, the
> unequal-spacing finding survives joint training.

---

## 9. Cluster Bootstrap

Users resampled with replacement, as in 07. Comparisons against M1, plus
M2d against M2c — the question of whether joint training beats the
two-step random-intercept model.

```{r bootstrap, fig.height=3.6}
ll_mat <- do.call(cbind, lapply(res_list, function(r) r$row_ll))
colnames(ll_mat) <- names(res_list)

rows_by_user <- split(seq_len(nrow(test)), test$userId)
u_names <- names(rows_by_user)

set.seed(SEED)
others <- setdiff(colnames(ll_mat), "M1")
cmp_names <- c(paste0(others, " - M1"), if (RUN_M2C) "M2d - M2c")
boot <- matrix(NA_real_, N_BOOT, length(cmp_names),
               dimnames = list(NULL, cmp_names))
for (b in seq_len(N_BOOT)) {
  idx <- unlist(rows_by_user[sample(u_names, length(u_names), replace = TRUE)],
                use.names = FALSE)
  mm  <- colMeans(ll_mat[idx, , drop = FALSE])
  boot[b, ] <- c(mm[others] - mm["M1"],
                 if (RUN_M2C) mm["M2d"] - mm["M2c"])
}
obs <- colMeans(ll_mat)
obs_diff <- c(obs[others] - obs["M1"],
              if (RUN_M2C) obs["M2d"] - obs["M2c"])
ci <- data.table(comparison  = colnames(boot),
                 diff_meanLL = as.numeric(obs_diff),
                 lo = apply(boot, 2, quantile, 0.025, na.rm = TRUE),
                 hi = apply(boot, 2, quantile, 0.975, na.rm = TRUE))
ci[, significant := (lo > 0) | (hi < 0)]
kable(ci, digits = 5,
      caption = sprintf("Cluster bootstrap, %d replicates, users resampled", N_BOOT))

p_ci <- ggplot(ci, aes(comparison, diff_meanLL)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "steelblue") +
  coord_flip() + theme_minimal() +
  labs(title = "Gain in mean log-likelihood",
       subtitle = "95% cluster bootstrap intervals",
       x = NULL, y = "Difference in mean log-likelihood")
print(p_ci)
ggsave(file.path(OUT_DIR, "fig10_m2d_bootstrap.png"), p_ci,
       width = 7, height = 3.6, dpi = 150)
```

---

## 10. Save Everything

```{r save}
fwrite(metrics,  file.path(OUT_DIR, "m2d_test_metrics.csv"))
fwrite(rank_tbl, file.path(OUT_DIR, "m2d_ranking_metrics.csv"))
fwrite(ci,       file.path(OUT_DIR, "m2d_bootstrap_ci.csv"))
fwrite(th_wide,  file.path(OUT_DIR, "m2d_thresholds.csv"))
fwrite(scan,     file.path(OUT_DIR, "m2d_hyperparameter_scan.csv"))

saveRDS(list(best_cfg = best_cfg, thresholds = th_d,
             calib_meanLL = m2d$best_ll, best_epoch = m2d$best_epoch,
             metrics = metrics, ranking = rank_tbl, ci = ci,
             sigma_m1 = sigma_m1),
        file.path(CACHE_DIR, "m2d_results.rds"))
cat("Written to", OUT_DIR, "\n")
```

---

## 11. What Is Not Here

1. **M2e** (per-user thresholds, no $b_u$ — the configuration the paper
   itself uses, their eq. 13). Next file; it reuses this trainer with a
   per-user threshold index and drops the user bias.
2. **Per-parameter-type learning-rate grids** (paper Section 6.1). One
   unified rate here by design; revisit only if M2d underperforms M2c.
3. **The identifiability note for the write-up**: with global thresholds a
   constant can be added to every $b_i$ and every threshold without
   changing the likelihood; the L2 penalty pins this down. In M2e the
   same argument removes $b_u$ entirely.

## References

Koren, Y. and Sill, J. (2011) OrdRec: an ordinal model for predicting
personalized item rating distributions. *RecSys '11*, pp. 117–124.

McCullagh, P. (1980) Regression models for ordinal data. *JRSS-B*, 42(2),
pp. 109–142.
