---
title: "M2e: Joint OrdRec with Per-User Thresholds (the Paper's Configuration)"
author: "Dina Muhamaitihan | Movie Recommendation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
library(data.table)
library(ggplot2)
library(ordinal)
library(knitr)
```

## Aim

M2e is the configuration Koren and Sill (2011) actually use. Ahead of their
eq. 13 they state that in their setting the thresholds are user-specific,
and that once they are, the user bias $b_u$ is no longer needed: any
constant that $b_u$ could add to the score can be absorbed by shifting that
user's four thresholds. The latent score therefore drops $b_u$,

$$y_{ui} = b_i + q_i^\top p_u,$$

and every user owns a threshold vector
$t^{(u)} = (t^{(u)}_1, \dots, t^{(u)}_4)$, parameterised exactly as before
($t^{(u)}_1$ free, $t^{(u)}_{r+1} = t^{(u)}_r + \exp(\beta^{(u)}_r)$, so
monotone by construction for every user). Everything else — the logit link,
the likelihood, the gradient structure of their eq. 12 — is unchanged.

**Position on the ladder.** M2d (file 09) used *global* thresholds; that
was our own ablation step, inserted so the effect of joint training could
be measured before user-specific thresholds were introduced. M2e completes
the ladder: the M2e $-$ M2d comparison isolates what per-user thresholds
buy *on top of* joint training, and M2e itself is the model the original
paper describes (with plain MF inside instead of SVD++, the same stated
deviation as in 09).

**Design choice: same hyperparameters as M2d, no new scan.** M2e trains
with the configuration the 09 scan selected (`eta = 0.03`, `lambda = 0.02`,
cold start). Re-tuning would confound the M2e $-$ M2d comparison: with
everything held fixed, any difference is attributable to the structural
change alone. The cost — M2e's hyperparameters are borrowed, not tuned for
it — is acknowledged in Section 11.

**Trust chain.** The likelihood/gradient core is the same algebra that
passed the numerical gradient check and the `ordinal::clm` cross-check in
08, re-indexed by user. The per-user trainer additionally passes its own
synthetic-recovery self-test in Section 3, in which the generating model
itself has heterogeneous per-user thresholds.

---

## 0. Settings

```{r config}
# --- edit these three paths (same as 07/09) ---------------------------
SEL_FILE  <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/00_data_selection/selection_v7_2018-04_6m_Nu10_Nm10.rds"
OUT_DIR   <- "D:/桌面/movie-recommendation-dissertation/06_models/output/07_M1_M2"
CACHE_DIR <- "D:/桌面/movie-recommendation-dissertation/06_models/cache/07_M1_M2"
# ----------------------------------------------------------------------

dir.create(OUT_DIR,   showWarnings = FALSE, recursive = TRUE)
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

SEED     <- 2024L
SVD_DIM  <- 16L
K_LEVELS <- 5L
LEVELS5  <- 1:5
CUTS     <- c(1.5, 2.5, 3.5, 4.5)
N_BOOT   <- 500L

# ---- M2e training configuration --------------------------------------
# Fixed at the optimum selected by the 09 scan (eta=0.03, lambda=0.02,
# cold start). No new hyperparameter search: see "Design choice" above.
BATCH_SIZE <- 512L
MAX_EPOCHS <- 200L
PATIENCE   <- 8L
ETA_DECAY  <- 0.95
CLIP       <- 25
ETA        <- 0.03
LAMBDA     <- 0.02

RUN_M2C <- TRUE   # clmm2 parameters are read from 09's cache (m2c_params.rds)

set.seed(SEED)
```

---

## 1. Load the Selection (identical to 07/09)

```{r load-data}
sel   <- readRDS(SEL_FILE)
fit   <- as.data.table(sel$fit)
calib <- as.data.table(sel$calib)
test  <- as.data.table(sel$test)

to_int5 <- function(d) { d[, rating5 := as.integer(as.character(rating5))]; d[] }
fit <- to_int5(fit); calib <- to_int5(calib); test <- to_int5(test)

user_map  <- data.table(userId  = sort(unique(fit$userId)))[,  user_idx  := .I]
movie_map <- data.table(movieId = sort(unique(fit$movieId)))[, movie_idx := .I]
add_idx <- function(d) {
  d <- merge(d, user_map,  by = "userId",  all.x = TRUE, sort = FALSE)
  d <- merge(d, movie_map, by = "movieId", all.x = TRUE, sort = FALSE)
  setDT(d)
  d[, cold := is.na(user_idx) | is.na(movie_idx)]
  d[]
}
fit   <- add_idx(fit)[cold == FALSE]
calib <- add_idx(calib)[cold == FALSE]
test  <- add_idx(test)[cold == FALSE]

nU <- nrow(user_map); nI <- nrow(movie_map)
cat(sprintf("fit %d | calib %d | test %d | users %d | movies %d\n",
            nrow(fit), nrow(calib), nrow(test), nU, nI))
stopifnot(nrow(fit) == 393364L)   # guard against loading a stale selection
```

---

## 2. Core Functions

Two families live side by side in this file: the **global-threshold**
functions (verbatim from 09) are needed to score the cached M2d model, and
the **per-user** functions (`_e` suffix) drive M2e. The metric functions
are verbatim from 07/09 so the evaluation protocol stays bit-identical.

```{r core-functions}
sigmoid <- function(x) 1 / (1 + exp(-pmin(pmax(x, -30), 30)))

# ---- global-threshold OrdRec (verbatim from 09, used to score M2d) ---
expand_thresholds <- function(t1, beta) c(t1, t1 + cumsum(exp(beta)))

calc_scores_d <- function(par, u, i)
  par$bi[i] + par$bu[u] + rowSums(par$P[u, , drop = FALSE] *
                                  par$Q[i, , drop = FALSE])

ordrec_probs_d <- function(par, u, i) {
  y  <- calc_scores_d(par, u, i)
  th <- expand_thresholds(par$t1, par$beta)
  CP <- sapply(th, function(t) plogis(t - y))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- per-user OrdRec (M2e) -------------------------------------------
# par$t1  : numeric length nU      (user-specific t_1)
# par$beta: matrix  nU x 3         (user-specific log-gaps)
# par$bi  : numeric length nI      (item biases; there is NO b_u)
# par$P   : nU x dim, par$Q : nI x dim

calc_scores_e <- function(par, u, i)
  par$bi[i] + rowSums(par$P[u, , drop = FALSE] * par$Q[i, , drop = FALSE])

# Threshold rows for the given user vector: one 4-vector per row of u.
user_threshold_rows <- function(par, u) {
  E <- exp(par$beta[u, , drop = FALSE])          # length(u) x 3
  t1 <- par$t1[u]
  cbind(t1,
        t1 + E[, 1],
        t1 + E[, 1] + E[, 2],
        t1 + E[, 1] + E[, 2] + E[, 3])
}

ordrec_forward_e <- function(y, r, Tmat) {
  n <- length(y)
  g_hi <- ifelse(r == K_LEVELS, 1,
                 sigmoid(Tmat[cbind(seq_len(n), pmin(r, K_LEVELS - 1L))] - y))
  g_lo <- ifelse(r == 1L, 0,
                 sigmoid(Tmat[cbind(seq_len(n), pmax(r - 1L, 1L))] - y))
  A <- g_hi * (1 - g_hi)
  B <- g_lo * (1 - g_lo)
  list(prob = pmax(g_hi - g_lo, 1e-12), A = A, B = B)
}

ordrec_mean_ll_e <- function(par, dat) {
  y  <- calc_scores_e(par, dat$u, dat$i)
  Tm <- user_threshold_rows(par, dat$u)
  mean(log(ordrec_forward_e(y, dat$r, Tm)$prob))
}

ordrec_probs_e <- function(par, u, i) {
  y  <- calc_scores_e(par, u, i)
  Tm <- user_threshold_rows(par, u)
  CP <- plogis(Tm - y)          # n x 4 minus length-n vector: recycles by column, correct
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

# ---- metrics, verbatim from 07/09 ------------------------------------
tidy_probs <- function(P) { P <- pmax(P, 1e-12); P / rowSums(P) }

gauss_probs <- function(y, sigma) {
  y  <- as.numeric(y)
  CP <- sapply(CUTS, function(c) pnorm((c - y) / sigma))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(y))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

cum_probs <- function(theta, eta) {
  eta <- as.numeric(eta)
  CP  <- sapply(theta, function(t) plogis(t - eta))
  if (is.null(dim(CP))) CP <- matrix(CP, nrow = length(eta))
  cbind(CP[, 1], CP[, 2:4] - CP[, 1:3], 1 - CP[, 4])
}

metric_set <- function(P, y) {
  y   <- as.integer(y); P <- tidy_probs(P); n <- nrow(P)
  ll  <- log(P[cbind(seq_len(n), y)])
  E   <- as.numeric(P %*% LEVELS5)
  CP  <- t(apply(P, 1, cumsum))
  IND <- outer(y, LEVELS5, function(a, b) as.numeric(a <= b))
  rps <- rowSums((CP[, 1:(K_LEVELS - 1)] - IND[, 1:(K_LEVELS - 1)])^2) / (K_LEVELS - 1)
  list(row_ll = ll,
       tbl = data.table(meanLL = mean(ll),
                        RMSE   = sqrt(mean((y - E)^2)),
                        RPS    = mean(rps),
                        CatAcc = mean(max.col(P, ties.method = "first") == y)))
}

ndcg_at_10 <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), rel = 2^as.integer(y) - 1)
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0) return(NA_real_)
  per_user <- d[, {
    k    <- min(10L, .N)
    dcg  <- sum(rel[order(-s)][1:k]   / log2(seq_len(k) + 1))
    idcg <- sum(rel[order(-rel)][1:k] / log2(seq_len(k) + 1))
    .(ndcg = if (idcg > 0) dcg / idcg else NA_real_)
  }, by = uid]
  mean(per_user$ndcg, na.rm = TRUE)
}

fcp_stat <- function(score, y, uid, min_items = 5L) {
  d <- data.table(uid = uid, s = as.numeric(score), y = as.integer(y))
  d <- d[, if (.N >= min_items) .SD, by = uid]
  if (nrow(d) == 0)
    return(list(FCP = NA_real_, nc = 0L, nd = 0L, n_pairs = 0L, n_users = 0L))
  per_user <- d[, {
    sy_gt <- outer(y, y, `>`); sy_lt <- outer(y, y, `<`)
    ss_gt <- outer(s, s, `>`); ss_ge <- outer(s, s, `>=`)
    list(nc = sum(ss_gt & sy_gt), nd = sum(ss_ge & sy_lt))
  }, by = uid]
  nc <- sum(per_user$nc); nd <- sum(per_user$nd)
  list(FCP = nc / (nc + nd), nc = nc, nd = nd,
       n_pairs = nc + nd, n_users = uniqueN(d$uid))
}
```

---

## 3. The Mini-Batch SGD Trainer (Per-User Thresholds)

The trainer is 09's trainer with the threshold parameters re-indexed by
user and $b_u$ removed. One update rule *changes as a consequence of the
re-indexing*, and it matters:

**Per-user thresholds get the batch-SUM gradient, not the batch mean.**
In 09 the four global thresholds were touched by every row of a batch, so
their summed gradient scaled with the batch size and had to be divided by
it. A user's thresholds are touched only by that user's rows — a handful
per batch — which makes them *entity parameters*, exactly like $b_u$ was.
Summing over the user's rows reproduces the total movement per-row SGD
would have produced, by the same argument Section 3 of 09 made for
$b_u$/$b_i$/$P$/$Q$. Dividing by the count here would silently shrink the
effective threshold learning rate relative to per-row SGD.

Thresholds remain **unpenalised**, as in 09 — but the justification is now
weaker (4·nU parameters instead of 4) and rests on two guards: every user
has $\ge$ 10 fit ratings by the selection rule, and early stopping on the
calibration set halts before per-user overfitting pays off. The fit–calib
gap is reported and discussed in Section 6.

Clipping, per-occurrence L2 on $b_i$/$P$/$Q$, learning-rate decay, and
early stopping are unchanged from 09.

```{r trainer}
train_ordrec_m2e <- function(dat_fit, dat_calib, init, eta, lambda,
                             batch_size = BATCH_SIZE, max_epochs = MAX_EPOCHS,
                             patience = PATIENCE, seed = SEED, clip = CLIP,
                             eta_decay = ETA_DECAY, verbose = TRUE) {
  par <- init
  n <- length(dat_fit$u)
  best <- list(ll = -Inf, par = par, epoch = 0L)
  hist <- data.table(epoch = integer(), fit_meanLL = numeric(),
                     calib_meanLL = numeric(), clip_rate = numeric())
  set.seed(seed)
  eta_e <- eta
  for (ep in seq_len(max_epochs)) {
    perm <- sample.int(n)
    n_clip <- 0L
    for (b0 in seq(1L, n, by = batch_size)) {
      idx <- perm[b0:min(b0 + batch_size - 1L, n)]
      u <- dat_fit$u[idx]; i <- dat_fit$i[idx]; r <- dat_fit$r[idx]

      y  <- calc_scores_e(par, u, i)
      E  <- exp(par$beta[u, , drop = FALSE])            # length(u) x 3
      t1v <- par$t1[u]
      Tm <- cbind(t1v, t1v + E[, 1], t1v + E[, 1] + E[, 2],
                  t1v + E[, 1] + E[, 2] + E[, 3])
      fw <- ordrec_forward_e(y, r, Tm)
      c_hi <- fw$A / fw$prob; c_lo <- fw$B / fw$prob
      n_clip <- n_clip + sum(c_hi > clip) + sum(c_lo > clip)
      c_hi <- pmin(c_hi, clip); c_lo <- pmin(c_lo, clip)
      dy <- c_lo - c_hi                                 # dl/dy per row

      # per-user thresholds: batch-SUM over each user's rows (see note above)
      g_t1 <- rowsum(c_hi - c_lo, u)
      g_b1 <- rowsum(E[, 1] * (c_hi * (1L <= r - 1L) - c_lo * (1L <= r - 2L)), u)
      g_b2 <- rowsum(E[, 2] * (c_hi * (2L <= r - 1L) - c_lo * (2L <= r - 2L)), u)
      g_b3 <- rowsum(E[, 3] * (c_hi * (3L <= r - 1L) - c_lo * (3L <= r - 2L)), u)
      ug <- as.integer(rownames(g_t1))                  # same grouping for all four
      par$t1[ug]      <- par$t1[ug]      + eta_e * as.numeric(g_t1)
      par$beta[ug, 1] <- par$beta[ug, 1] + eta_e * as.numeric(g_b1)
      par$beta[ug, 2] <- par$beta[ug, 2] + eta_e * as.numeric(g_b2)
      par$beta[ug, 3] <- par$beta[ug, 3] + eta_e * as.numeric(g_b3)

      # user factors: batch-SUM, per-occurrence L2 (no b_u to update)
      SP <- rowsum(dy * par$Q[i, , drop = FALSE], u)
      nu <- as.numeric(rowsum(rep(1, length(u)), u))
      par$P[ug, ] <- par$P[ug, ] + eta_e * (SP - lambda * nu * par$P[ug, , drop = FALSE])

      # items: unchanged from 09
      si <- rowsum(dy, i); ig <- as.integer(rownames(si))
      ni <- as.numeric(rowsum(rep(1, length(i)), i))
      par$bi[ig] <- par$bi[ig] + eta_e * (as.numeric(si) - lambda * ni * par$bi[ig])
      SQ <- rowsum(dy * par$P[u, , drop = FALSE], i)
      par$Q[ig, ] <- par$Q[ig, ] + eta_e * (SQ - lambda * ni * par$Q[ig, , drop = FALSE])

      if (any(!is.finite(par$t1[ug])) || any(!is.finite(par$beta[ug, ])))
        return(list(diverged = TRUE, best_ll = -Inf, best_epoch = 0L,
                    history = hist))
    }
    ll_c <- ordrec_mean_ll_e(par, dat_calib)
    ll_f <- ordrec_mean_ll_e(par, dat_fit)
    hist <- rbind(hist, data.table(epoch = ep, fit_meanLL = ll_f,
                                   calib_meanLL = ll_c, clip_rate = n_clip / n))
    if (verbose) cat(sprintf("    epoch %2d  fit %.5f  calib %.5f  clip %.5f%%\n",
                             ep, ll_f, ll_c, 100 * n_clip / n))
    if (ll_c > best$ll + 1e-5) {
      best <- list(ll = ll_c, par = par, epoch = ep)
    } else if (ep - best$epoch >= patience) break
    eta_e <- eta_e * eta_decay
  }
  list(diverged = FALSE, best_ll = best$ll, par = best$par,
       best_epoch = best$epoch, history = hist)
}
```

### Synthetic self-test

The generating model itself has heterogeneous per-user thresholds (drawn
around a common centre), so the test exercises exactly the machinery that
is new in this file. Requirements as in 09: clearly beat the marginal
baseline and land near the oracle likelihood. The oracle tolerance is
slightly looser than 09's (0.08 vs 0.06 nats) because per-user thresholds
give the model more parameters to estimate from the same data.

```{r trainer-selftest}
set.seed(7)
.nU <- 400L; .nI <- 200L; .K <- 8L
.true <- list(t1   = rnorm(.nU, 0.5, 0.5),
              beta = matrix(rnorm(.nU * 3, rep(log(c(1.2, 1.6, 2.0)), each = .nU), 0.25),
                            .nU, 3),
              bi = rnorm(.nI, 0, 0.5),
              P = matrix(rnorm(.nU * .K, 0, 0.55), .nU, .K),
              Q = matrix(rnorm(.nI * .K, 0, 0.55), .nI, .K))
.gen <- function(n) {
  u <- sample(.nU, n, TRUE); i <- sample(.nI, n, TRUE)
  y  <- calc_scores_e(.true, u, i)
  Tm <- user_threshold_rows(.true, u)
  G  <- cbind(plogis(Tm - y), 1)
  P5 <- cbind(G[, 1], G[, 2:5] - G[, 1:4])
  list(u = u, i = i,
       r = as.integer(apply(P5, 1, function(p) sample.int(5, 1, prob = pmax(p, 0)))))
}
.df <- .gen(60000L); .dc <- .gen(15000L)
.oracle <- ordrec_mean_ll_e(.true, .dc)
.marg   <- prop.table(tabulate(.df$r, 5))
.mll    <- mean(log(.marg[.dc$r]))
.cum    <- qlogis(cumsum(.marg)[1:4])
.cold   <- list(t1   = rep(.cum[1], .nU),
                beta = matrix(log(pmax(diff(.cum), 1e-3)), .nU, 3, byrow = TRUE),
                bi = rep(0, .nI),
                P = matrix(rnorm(.nU * .K, 0, 0.05), .nU, .K),
                Q = matrix(rnorm(.nI * .K, 0, 0.05), .nI, .K))
.res <- train_ordrec_m2e(.df, .dc, .cold, eta = 0.03, lambda = 0.01,
                         verbose = FALSE)
cat(sprintf("synthetic: trained %.5f | oracle %.5f | marginal %.5f\n",
            .res$best_ll, .oracle, .mll))
stopifnot(!.res$diverged, .res$best_ll > .mll + 0.05,
          .res$best_ll > .oracle - 0.08)
cat("TRAINER SELF-TEST: PASS\n")
rm(.df, .dc, .true, .cold, .res)
```

---

## 4. Baselines from Cache (M1, M2a, M2b, M2c, M2d)

All comparison models are reconstructed from the shared caches, so every
number in this file is on exactly the same footing as 09: the SVD factors
come from `svd_factors.rds` (07's tuned fit, dim 16, L2 = 0.05), M2c's
parameters from `m2c_params.rds`, and M2d in full from `m2d_final.rds`.
Nothing is re-trained here except M2e itself.

```{r baselines}
svd_cache <- readRDS(file.path(CACHE_DIR, "svd_factors.rds"))
stopifnot(identical(as.integer(svd_cache$user_map$userId),   as.integer(user_map$userId)),
          identical(as.integer(svd_cache$movie_map$movieId), as.integer(movie_map$movieId)),
          svd_cache$dim == SVD_DIM,
          nrow(svd_cache$P) == nU, nrow(svd_cache$Q) == nI)
Pmat <- svd_cache$P; Qmat <- svd_cache$Q
cat(sprintf("Loaded SVD factors from cache (dim=%d, L2=%g)\n",
            svd_cache$dim, svd_cache$l2))

svd_score_vec <- function(u, i)
  rowSums(Pmat[u, , drop = FALSE] * Qmat[i, , drop = FALSE])
calib[, svd_score := svd_score_vec(user_idx, movie_idx)]
test[,  svd_score := svd_score_vec(user_idx, movie_idx)]
calib[, svd_clip := pmin(pmax(svd_score, 1), 5)]
test[,  svd_clip := pmin(pmax(svd_score, 1), 5)]

# M1
sigma_m1 <- sqrt(mean((as.numeric(calib$rating5) - calib$svd_clip)^2))
P_m1 <- gauss_probs(test$svd_clip, sigma_m1)

# M2a / M2b (fast clm fits, as in 09)
calib[, rating_f := factor(rating5, levels = LEVELS5, ordered = TRUE)]
m2a <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "equidistant")
m2b <- clm(rating_f ~ svd_score, data = calib, link = "logit",
           threshold = "flexible")
th_a <- as.numeric(m2a$alpha); b_a <- as.numeric(m2a$beta)
th_a <- th_a[1] + (0:3) * th_a[2]
th_b <- as.numeric(m2b$alpha); b_b <- as.numeric(m2b$beta)
P_m2a <- cum_probs(th_a, b_a * test$svd_score)
P_m2b <- cum_probs(th_b, b_b * test$svd_score)

# M2c from 09's cache (valid: the svd_score inputs are unchanged)
if (RUN_M2C) {
  m2c_cache <- file.path(CACHE_DIR, "m2c_params.rds")
  stopifnot("Run 09 first: m2c_params.rds not found" = file.exists(m2c_cache))
  m2c_par <- readRDS(m2c_cache)
  bu_c <- m2c_par$ranef[as.character(test$userId)]
  bu_c[is.na(bu_c)] <- 0
  P_m2c <- cum_probs(m2c_par$Theta,
                     m2c_par$beta * test$svd_score + as.numeric(bu_c))
}

# M2d in full from 09's final cache (must be the eta=0.03/lambda=0.02 refit)
m2d_cache <- file.path(CACHE_DIR, "m2d_final.rds")
stopifnot("Run 09 first: m2d_final.rds not found" = file.exists(m2d_cache))
m2d <- readRDS(m2d_cache)
th_d <- expand_thresholds(m2d$par$t1, m2d$par$beta)
cat(sprintf("Loaded M2d (best epoch %d, calib meanLL %.5f)\n",
            m2d$best_epoch, m2d$best_ll))
stopifnot(abs(m2d$best_ll - (-1.10002)) < 2e-4)  # guard: the stale pre-refit
# cache had calib -1.10080, only 8e-4 away, so the tolerance must be tight
P_m2d <- ordrec_probs_d(m2d$par, test$user_idx, test$movie_idx)
```

---

## 5. Initialisation for M2e

Cold start, mirroring the configuration the 09 scan selected: small random
factors, zero item biases, and *every user starting from the same marginal
thresholds* ($t_k = \mathrm{logit}$ of the empirical $P(r \le k)$ on the
fit set). All heterogeneity in the final thresholds is therefore learned,
none is baked into the initialisation.

```{r inits}
dat_fit   <- list(u = fit$user_idx,   i = fit$movie_idx,   r = fit$rating5)
dat_calib <- list(u = calib$user_idx, i = calib$movie_idx, r = calib$rating5)
dat_test  <- list(u = test$user_idx,  i = test$movie_idx,  r = test$rating5)

marg5  <- prop.table(tabulate(dat_fit$r, K_LEVELS))
t_marg <- qlogis(cumsum(marg5)[1:4])
set.seed(SEED)
init_cold <- list(t1   = rep(t_marg[1], nU),
                  beta = matrix(log(pmax(diff(t_marg), 1e-3)), nU, 3, byrow = TRUE),
                  bi = rep(0, nI),
                  P = matrix(rnorm(nU * SVD_DIM, 0, 0.05), nU, SVD_DIM),
                  Q = matrix(rnorm(nI * SVD_DIM, 0, 0.05), nI, SVD_DIM))
cat("cold-start thresholds (all users):", round(t_marg, 4), "\n")
```

---

## 6. Final M2e Training

```{r final-train, fig.height=4}
final_cache <- file.path(CACHE_DIR, "m2e_final.rds")
if (file.exists(final_cache)) {
  m2e <- readRDS(final_cache)
  cat("Loaded cached final M2e.\n")
} else {
  m2e <- train_ordrec_m2e(dat_fit, dat_calib, init_cold,
                          eta = ETA, lambda = LAMBDA, verbose = TRUE)
  saveRDS(m2e, final_cache)
}
stopifnot(!m2e$diverged)
gap <- m2e$history[epoch == m2e$best_epoch,
                   fit_meanLL - calib_meanLL]
cat(sprintf("best epoch %d | calib meanLL %.5f | fit-calib gap %.4f | max clip rate %.5f%%\n",
            m2e$best_epoch, m2e$best_ll, gap, 100 * max(m2e$history$clip_rate)))

hc <- melt(m2e$history[, .(epoch, fit_meanLL, calib_meanLL)], id.vars = "epoch")
p_curve <- ggplot(hc, aes(epoch, value, colour = variable)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.5) +
  geom_vline(xintercept = m2e$best_epoch, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(fit_meanLL = "steelblue",
                                 calib_meanLL = "coral"),
                      labels = c("fit", "calibration")) +
  theme_minimal() +
  labs(title = "M2e training curves",
       subtitle = "Dashed line: epoch with the best calibration likelihood (parameters kept).",
       x = "Epoch", y = "Mean log-likelihood", colour = NULL)
print(p_curve)
ggsave(file.path(OUT_DIR, "fig11_m2e_training_curve.png"), p_curve,
       width = 8, height = 4, dpi = 150)
```

> **Plain read.** Compare the fit–calib gap against M2d's (0.09-nats
> territory at its best epoch). Per-user thresholds add 4·nU parameters
> with no penalty; if the gap widens materially, say so in the write-up —
> early stopping contains the damage but does not hide the capacity.

---

## 7. Test-Month Evaluation

Same protocol as 07/09, one new row. Every test user is warm by
construction, so every test row has trained user thresholds to index.

```{r test-eval}
stopifnot(all(dat_test$u >= 1L & dat_test$u <= nU))
P_m2e <- ordrec_probs_e(m2e$par, dat_test$u, dat_test$i)

P_list <- list(M1 = P_m1, M2a = P_m2a, M2b = P_m2b)
if (RUN_M2C) P_list$M2c <- P_m2c
P_list$M2d <- P_m2d
P_list$M2e <- P_m2e

res_list <- lapply(P_list, metric_set, y = test$rating5)
metrics  <- rbindlist(lapply(names(res_list), function(m)
  cbind(data.table(model = m), res_list[[m]]$tbl)))
kable(metrics, digits = 4, caption = "Test month, all models")

rank_tbl <- rbindlist(lapply(names(P_list), function(m) {
  E <- as.numeric(tidy_probs(P_list[[m]]) %*% LEVELS5)
  f <- fcp_stat(E, test$rating5, test$userId)
  data.table(model = m,
             NDCG10 = ndcg_at_10(E, test$rating5, test$userId),
             FCP = f$FCP, n_pairs = f$n_pairs, n_users = f$n_users)
}))
kable(rank_tbl, digits = 6, caption = "Ranking metrics (users with >= 5 test ratings)")
```

> **Plain read.** Two comparisons matter most. (1) M2e against M2d on
> meanLL/RPS: what do per-user thresholds add on top of joint training?
> (2) M2e against M2d on NDCG/FCP: within a single user, that user's
> thresholds are four monotone constants, so the *ordering* of that
> user's items by expected rating can only change through the factors —
> if ranking moves relative to M2d, it moved because the factors trained
> to a different place, not because the thresholds re-ranked anything.
> Either direction is a reportable result.

```{r}
# 接在 Section 7 thresholds chunk 后面
lv_used <- fit[, .(n_lv = uniqueN(rating5),
                   used_1 = any(rating5 == 1L)), by = user_idx]
gaps_e[, user_idx := seq_len(nU)]
gd <- merge(gaps_e, lv_used, by = "user_idx")
kable(gd[, .(n_users = .N,
             g12 = median(g12), g23 = median(g23), g34 = median(g34)),
         by = .(full_scale = n_lv == 5L)][order(-full_scale)],
      digits = 3,
      caption = "Gap medians split by whether the user used all five levels in the fit set")
cat(sprintf("users using all 5 levels: %d / %d (%.1f%%)\n",
            sum(gd$n_lv == 5L), nU, 100 * mean(gd$n_lv == 5L)))
```

### Per-user thresholds: heterogeneity and gap structure

The absolute threshold values live on M2e's own latent scale; what is
comparable across models is the *gap pattern*. For M2e each user has their
own gaps, so we report their distribution against the single global gap
vectors of M2b/M2c/M2d.

```{r thresholds, fig.height=3.8}
Tall <- user_threshold_rows(m2e$par, seq_len(nU))
gaps_e <- data.table(g12 = Tall[, 2] - Tall[, 1],
                     g23 = Tall[, 3] - Tall[, 2],
                     g34 = Tall[, 4] - Tall[, 3])
gap_summary <- gaps_e[, .(gap = c("g12", "g23", "g34"),
                          median = sapply(.SD, median),
                          q25 = sapply(.SD, quantile, 0.25),
                          q75 = sapply(.SD, quantile, 0.75))]
kable(gap_summary, digits = 3,
      caption = "M2e per-user threshold gaps: median and IQR across users")

ref_gaps <- data.table(model = c("M2b", "M2c", "M2d", "M2e (median)"),
                       g12 = c(th_b[2] - th_b[1],
                               m2c_par$Theta[2] - m2c_par$Theta[1],
                               th_d[2] - th_d[1], median(gaps_e$g12)),
                       g23 = c(th_b[3] - th_b[2],
                               m2c_par$Theta[3] - m2c_par$Theta[2],
                               th_d[3] - th_d[2], median(gaps_e$g23)),
                       g34 = c(th_b[4] - th_b[3],
                               m2c_par$Theta[4] - m2c_par$Theta[3],
                               th_d[4] - th_d[3], median(gaps_e$g34)))
kable(ref_gaps, digits = 3,
      caption = "Threshold gaps across the ladder (does unequal spacing survive per-user freedom?)")

# does t1 track user generosity? (it should: t1_u absorbs -b_u)
user_meanr <- fit[, .(mean_r = mean(rating5)), by = user_idx]
t1_dt <- data.table(user_idx = seq_len(nU), t1 = m2e$par$t1)[user_meanr, on = "user_idx"]
p_t1 <- ggplot(t1_dt, aes(mean_r, t1)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "steelblue") +
  geom_smooth(method = "loess", colour = "coral", se = FALSE, linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Per-user first threshold vs. mean fit-set rating",
       subtitle = sprintf("Spearman rho = %.3f. Generous raters should sit lower (thresholds absorb -b_u).",
                          cor(t1_dt$mean_r, t1_dt$t1, method = "spearman")),
       x = "User mean rating (fit set)", y = expression(t[1]^{(u)}))
print(p_t1)
ggsave(file.path(OUT_DIR, "fig12_m2e_t1_vs_meanrating.png"), p_t1,
       width = 7, height = 3.8, dpi = 150)
```

> **Plain read.** If the median gap pattern still widens towards 4|5, the
> unequal-spacing finding has now survived four estimation routes,
> including one where every user was free to disagree. The IQR shows how
> much users actually disagree. The scatter is the interpretability
> dividend of the paper's configuration: the first threshold should slope
> *downwards* in the user's mean rating, because a generous user's
> thresholds shift down to absorb what $b_u$ used to model.

---

## 8. Cluster Bootstrap

Users resampled with replacement, as in 07/09. Comparisons against M1 for
the whole ladder, plus the two structural questions: M2d $-$ M2c (joint
training, carried over from 09 for completeness) and **M2e $-$ M2d**
(per-user thresholds, the new question of this file).

```{r bootstrap, fig.height=3.9}
ll_mat <- do.call(cbind, lapply(res_list, function(r) r$row_ll))
colnames(ll_mat) <- names(res_list)

rows_by_user <- split(seq_len(nrow(test)), test$userId)
u_names <- names(rows_by_user)

set.seed(SEED)
others <- setdiff(colnames(ll_mat), "M1")
cmp_names <- c(paste0(others, " - M1"),
               if (RUN_M2C) "M2d - M2c", "M2e - M2d")
boot <- matrix(NA_real_, N_BOOT, length(cmp_names),
               dimnames = list(NULL, cmp_names))
for (b in seq_len(N_BOOT)) {
  idx <- unlist(rows_by_user[sample(u_names, length(u_names), replace = TRUE)],
                use.names = FALSE)
  mm  <- colMeans(ll_mat[idx, , drop = FALSE])
  boot[b, ] <- c(mm[others] - mm["M1"],
                 if (RUN_M2C) mm["M2d"] - mm["M2c"],
                 mm["M2e"] - mm["M2d"])
}
obs <- colMeans(ll_mat)
obs_diff <- c(obs[others] - obs["M1"],
              if (RUN_M2C) obs["M2d"] - obs["M2c"],
              obs["M2e"] - obs["M2d"])
ci <- data.table(comparison  = colnames(boot),
                 diff_meanLL = as.numeric(obs_diff),
                 lo = apply(boot, 2, quantile, 0.025, na.rm = TRUE),
                 hi = apply(boot, 2, quantile, 0.975, na.rm = TRUE))
ci[, significant := (lo > 0) | (hi < 0)]
kable(ci, digits = 5,
      caption = sprintf("Cluster bootstrap, %d replicates, users resampled", N_BOOT))

p_ci <- ggplot(ci, aes(comparison, diff_meanLL)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "steelblue") +
  coord_flip() + theme_minimal() +
  labs(title = "Gain in mean log-likelihood",
       subtitle = "95% cluster bootstrap intervals",
       x = NULL, y = "Difference in mean log-likelihood")
print(p_ci)
ggsave(file.path(OUT_DIR, "fig13_m2e_bootstrap.png"), p_ci,
       width = 7, height = 3.9, dpi = 150)
```

---

## 9. Save Everything

```{r save}
fwrite(metrics,     file.path(OUT_DIR, "m2e_test_metrics.csv"))
fwrite(rank_tbl,    file.path(OUT_DIR, "m2e_ranking_metrics.csv"))
fwrite(ci,          file.path(OUT_DIR, "m2e_bootstrap_ci.csv"))
fwrite(ref_gaps,    file.path(OUT_DIR, "m2e_threshold_gaps.csv"))
fwrite(gap_summary, file.path(OUT_DIR, "m2e_gap_distribution.csv"))

saveRDS(list(eta = ETA, lambda = LAMBDA,
             calib_meanLL = m2e$best_ll, best_epoch = m2e$best_epoch,
             fit_calib_gap = gap,
             metrics = metrics, ranking = rank_tbl, ci = ci,
             gap_summary = gap_summary, sigma_m1 = sigma_m1),
        file.path(CACHE_DIR, "m2e_results.rds"))
cat("Written to", OUT_DIR, "\n")
```

---

## 10. What Is Not Here

1. **A hyperparameter scan for M2e.** Deliberate: hyperparameters are
   inherited from the 09 scan so that M2e $-$ M2d isolates the structural
   change. The asymmetry (M2d's settings were tuned for M2d) is a caveat
   for the write-up, and it cuts *against* M2e — if M2e wins anyway, the
   conclusion is conservative.
2. **Regularisation or partial pooling of per-user thresholds.** They are
   unpenalised, guarded only by the $\ge$ 10-ratings selection floor and
   early stopping. A hierarchical prior (shrinking each user's thresholds
   towards the global ones) is the natural next model; it is discussed,
   not fitted.
3. **User–item joint thresholds** (paper's eqs. 14–15): future work in the
   original too. Not touched.
4. **Confidence estimation** (paper's Section 7): planned as a short
   analysis for the write-up using M2e's predictive distributions, since
   M2e is the model whose distribution shapes vary most across user–item
   pairs.

## References

Koren, Y. and Sill, J. (2011) OrdRec: an ordinal model for predicting
personalized item rating distributions. *RecSys '11*, pp. 117–124.

McCullagh, P. (1980) Regression models for ordinal data. *JRSS-B*, 42(2),
pp. 109–142.
